---
title: "Exercise Sheet 3 — One-Way ANOVA"
subtitle: "Study Guide: Worked Solutions with Explanations"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    highlight-style: github
    df-print: paged
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

> 📝 **How to use this guide:** Each question is answered step by step, with plain-English explanations *before* the formal answer. Code is shown so you can reproduce every result in R. Leave the annotation boxes empty — those are for your own notes later!

---

# Key Topics Covered

- **Experimental design vocabulary** — factors, levels, treatments
- **Randomized block design** — what it is and when to use it
- **Multiple testing problem** — Type I error accumulation
- **ANOVA assumptions** — what needs to hold before you run the test
- **One-Way ANOVA** — hypothesis test, ANOVA table, F-statistic
- **Effect size η²** — how big is the effect, not just is it significant?
- **Tukey's HSD post hoc test** — which groups are actually different?
- **ANOVA model** — writing the regression-style equation
- **Model diagnostics** — checking your assumptions after fitting
- **Interpreting outputs** — Fligner-Killeen, Levene, diagnostic plots

---

# Core Concepts Explained

## What is ANOVA? (The Big Picture)

Imagine you want to know whether three different fertilizers produce different crop yields. You could run lots of t-tests comparing them two at a time, but that inflates your chance of getting a false positive. **ANOVA (Analysis of Variance)** solves this — it compares *all* groups in a single test by asking: "Is the variation *between* groups larger than the variation *within* groups?"

If yes, something is going on. If no, any group differences are probably just random noise.

## Experimental Design Vocabulary

| Term | Plain English | Formal Definition |
|---|---|---|
| **Factor** | The thing you're changing on purpose | A controlled explanatory variable whose levels are set by the experimenter |
| **Level** | The specific settings of that thing | Different values of the controlled variable |
| **Treatment** | Each unique experimental condition | Each combination of factor levels |

> 💡 **Memory trick:** A *factor* is like a dial on a machine. The *levels* are the positions you can turn it to. A *treatment* is one specific dial position (or combination of positions if there are multiple dials).

## The ANOVA Model

The ANOVA model looks like this:

$$y_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

| Symbol | Meaning |
|---|---|
| $y_{ij}$ | The observation for subject $j$ in group $i$ |
| $\mu$ | The overall mean (grand mean) |
| $\tau_i$ | The effect of treatment $i$ (how much group $i$ differs from overall mean) |
| $\varepsilon_{ij}$ | Random error — the bit we can't explain |

In R's regression-style output, this becomes:

$$\hat{y}_{ij} = \hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 + \cdots$$

Where $\hat{\beta}_0$ is the mean of the **reference group** (usually group 1), and the $\hat{\beta}$ values are how much each other group differs from it.

## The ANOVA Table — Decoded

```
          Df   Sum Sq   Mean Sq   F value   Pr(>F)
Treatment  k-1   SS_T    MS_T      F         p
Residuals  N-k   SS_E    MS_E
```

- **Df** = degrees of freedom. For the treatment: number of groups minus 1. For residuals: total observations minus number of groups.
- **Sum Sq** = total variability attributable to this source.
- **Mean Sq** = Sum Sq / Df (average variability per degree of freedom).
- **F value** = MS_Treatment / MS_Residuals. Big F = groups are more different than random noise.
- **Pr(>F)** = p-value. If < 0.05, we reject H₀ (at least one group mean differs).

## Effect Size η² (Eta-Squared)

The p-value tells you *if* an effect exists. η² tells you *how big* it is.

$$\eta^2 = \frac{SS_{Treatment}}{SS_{Total}} = \frac{SS_{Treatment}}{SS_{Treatment} + SS_{Residuals}}$$

| η² value | Interpretation |
|---|---|
| ~0.01 | Small effect |
| ~0.06 | Medium effect |
| ~0.14+ | Large effect |

> 💡 η² is literally "the proportion of total variation explained by the treatment."

## Tukey's HSD — The Pairwise Test

After ANOVA tells you *something* is different, Tukey's Honestly Significant Difference (HSD) test tells you *which pairs* are different. It controls the family-wise error rate (the chance of any false positive across all comparisons).

**Reading Tukey output:** Look at `p adj`. If it's < 0.05, those two groups are significantly different.

## Model Diagnostics — The Four Plots

When you run `plot(model)` in R, you get four plots:

1. **Residuals vs Fitted:** Should show random scatter with no pattern. A pattern = violated assumptions.
2. **Q-Q Plot:** Residuals should fall on the diagonal line. Deviations = non-normality.
3. **Scale-Location:** Should be flat. A funnel shape = unequal variances (heteroscedasticity).
4. **Residuals vs Leverage:** Points with high leverage AND high residuals are *influential* — they may be pulling your results.

---

# Questions and Worked Solutions

## Q1 — Experimental Design Vocabulary

### Question
In the context of experimental design, what is: (i) a factor? (ii) a level? (iii) a treatment?

### Step-by-step explanation

Think of making coffee. You decide to test whether grind size and water temperature affect taste.

- The things you're **controlling** (grind size, water temperature) → **factors**
- The specific settings you try (coarse/fine/medium, 85°C/90°C/95°C) → **levels**
- Each specific combination you actually brew (fine + 90°C, coarse + 95°C, etc.) → **treatment**

### Answer

**(i) Factor:** A factor is a controlled explanatory variable — a variable whose levels are deliberately set by the experimenter.

**(ii) Level:** A level represents a specific value (or category) of a factor.

**(iii) Treatment:** A treatment is one specific experimental condition. If there is only one factor, each level is a treatment. If there are multiple factors, each **combination** of factor levels is a treatment.

> 📝 *Annotation space — add your own notes here:*

---

## Q2 — Drill Press Experiment

### Question
A mechanical engineer studies thrust force developed by a drill press. He suspects drilling speed and feed rate affect thrust force. He selects 4 feed rates and 2 drilling speeds (high and low).

(i) How many factors? (ii) How many levels each? (iii) How many treatments?

### Step-by-step explanation

Count what is being deliberately varied:

- **Factor 1:** Drilling speed (2 levels: high, low)
- **Factor 2:** Feed rate (4 levels)
- **Treatments:** Every possible combination = 2 × 4 = **8 treatments**

### Answer

**(i)** Two factors: drilling speed and feed rate.

**(ii)** Drilling speed has 2 levels (high, low). Feed rate has 4 levels.

**(iii)** 2 × 4 = **8 treatments** total.

> 📝 *Annotation space:*

---

## Q3 & Q4 — Randomized Block Design

### Question
Give an example of a randomized block design. Why is it appropriate?

### Plain English Explanation

A **randomized block design** is used when you know there's a nuisance variable that could affect your results, but you're not interested in it — you just want to *control* for it.

**Example:** A manufacturer wants to test 4 chemical formulations. Each formulation is made from a batch of raw material, but each batch is only large enough for 4 formulations. Different batches might have natural variation. So "batch" is a nuisance variable.

**Design:**

| Formula | Batch 1 | Batch 2 | Batch 3 | Batch 4 | Batch 5 |
|---|---|---|---|---|---|
| 1 | ✓ | ✓ | ✓ | ✓ | ✓ |
| 2 | ✓ | ✓ | ✓ | ✓ | ✓ |
| 3 | ✓ | ✓ | ✓ | ✓ | ✓ |
| 4 | ✓ | ✓ | ✓ | ✓ | ✓ |

Each batch = one "block." Within each block, formulas are randomly assigned. This controls for batch-to-batch variation.

### Answer

**Q3:** A manufacturer testing 4 formulations across 5 batches of raw material, where each batch introduces natural variation, uses a randomized block design (see table above).

**Q4:** The randomized block design is appropriate because it **controls for the effect of batch variation**, which would otherwise mask or confound the treatment effects.

> 📝 *Annotation space:*

---

## Q5 — Multiple Testing and Type I Error

### Question
If we test **five** null hypotheses (all true) at significance level α = 0.01, what is the probability of making **at least one** Type I error?

### Plain English Explanation

A **Type I error** is a false positive — rejecting H₀ when it's actually true. At α = 0.01, each individual test has a 1% chance of a false positive.

The trick is to use the complement: instead of calculating P(at least one error), calculate P(no errors at all) and subtract from 1.

$$P(\text{at least one error}) = 1 - P(\text{no errors})$$

$$P(\text{no error on one test}) = 1 - 0.01 = 0.99$$

$$P(\text{no errors on 5 tests}) = 0.99^5 = 0.9510$$

$$P(\text{at least one error}) = 1 - 0.9510 = \mathbf{0.049}$$

### Answer

$$P(\text{at least one Type I error}) = 1 - (1-0.01)^5 = 1 - 0.99^5 \approx \mathbf{0.049}$$

Even though each test had only a 1% false positive rate, running 5 tests brings the combined risk up to nearly 5%! This is why we need methods like Tukey's HSD when making multiple comparisons.

> 📝 *Annotation space:*

---

## Q6 — ANOVA Assumptions

### Question
Name two assumptions that should be satisfied before carrying out an ANOVA.

### Answer

1. **Independence (Random Sampling):** Observations must be independent of each other. No observation should influence another.
2. **Homogeneity of Variance (Equal Variances):** The variance within each group should be approximately equal across all groups. Tested using the **Fligner-Killeen** or **Levene** test.

> **Bonus (not always tested but important to know):** Normality — the residuals (errors) should be approximately normally distributed. This is checked via the Q-Q plot in model diagnostics.

> 📝 *Annotation space:*

---

## Q7 — Kenton Foods: Full ANOVA Worked Example

### Question
Kenton Food Company tests 4 package designs for cereal across 20 stores (5 stores per design). Analyse the R output provided.

### Setup and Data

```{r}
#| label: kenton-setup
library(tidyverse)

# Load the data
kenton <- read_tsv("Kenton_Foods.txt")
kenton$Package <- as.factor(kenton$Package)

# Quick look
glimpse(kenton)
```

### Step 1: Exploratory Data Analysis

```{r}
#| label: kenton-eda

# Summary statistics by package
kenton %>%
  group_by(Package) %>%
  summarise(
    mean_Cases = mean(Cases),
    sd_Cases   = sd(Cases),
    n          = n()
  )

# Boxplot
boxplot(kenton$Cases ~ kenton$Package,
        xlab = "Package Design",
        ylab = "Cases Sold",
        main = "Cases Sold by Package Design",
        col  = c("#AED6F1", "#A9DFBF", "#F9E79F", "#F1948A"))
```

**How to read the boxplot:** Each box shows the middle 50% of sales for that package. The line in the middle is the median. Points outside the "whiskers" are potential outliers. Here, Package 4 clearly has higher sales.

### Step 2: Q7(i) — Fligner-Killeen Test

#### Plain English
Before running ANOVA, we must check that the **variance in sales is similar** across all package designs. The Fligner-Killeen test does this. 

- H₀: all groups have equal variance
- Hₐ: at least one group has different variance

If p > 0.05, we're fine to proceed.

```{r}
#| label: kenton-fligner
fligner.test(kenton$Cases ~ kenton$Package)
```

**Result from output:** Fligner-Killeen: med chi-squared = 1.1321, df = 3, **p-value = 0.7693**

**Interpretation:** p = 0.7693 >> 0.05. We **fail to reject H₀**. There is no significant difference in variance across package designs. ✅ The equal variance assumption is satisfied — we can proceed with ANOVA.

### Step 3: Q7(ii) — Hypotheses

$$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4$$

"The mean number of cases sold is the same for all four package designs."

$$H_A: \text{At least one } \mu_i \text{ is different}$$

"At least one package design leads to a different mean number of cases sold."

### Step 4: Q7(iii) — Degrees of Freedom

**Why does Package have 3 degrees of freedom?**

The formula is: df = (number of groups) − 1 = 4 − 1 = **3**

Think of it this way: once you know 3 group means and the overall mean, the 4th group mean is already determined. You "use up" one df by estimating the overall mean.

### Step 5: Q7(iv) — Run the ANOVA

```{r}
#| label: kenton-anova
model_kenton <- aov(kenton$Cases ~ kenton$Package)
summary(model_kenton)
```

**Reading the ANOVA Table:**

| Source | Df | Sum Sq | Mean Sq | F value | p-value |
|---|---|---|---|---|---|
| Package | 3 | 586.8 | 195.6 | 19.56 | 1.33e-05 *** |
| Residuals | 16 | 160.0 | 10.0 | | |

- **F(3, 16) = 19.56** — The between-group variation is nearly 20× the within-group noise. That's huge.
- **p < 0.001** — Extremely strong evidence against H₀.

**Effect Size:**

$$\eta^2 = \frac{SS_{Package}}{SS_{Package} + SS_{Residuals}} = \frac{586.8}{586.8 + 160.0} = \frac{586.8}{746.8} \approx \mathbf{0.786}$$

This means **78.6% of the total variation** in cases sold is explained by the package design. That is a very large effect.

### Step 6: Tukey's HSD — Pairwise Comparisons

```{r}
#| label: kenton-tukey
TukeyHSD(model_kenton)
plot(TukeyHSD(model_kenton))
```

**Reading the Tukey output:**

| Comparison | Difference | 95% CI Lower | 95% CI Upper | p adj | Significant? |
|---|---|---|---|---|---|
| 2 vs 1 | −1.2 | −6.92 | 4.52 | 0.931 | ❌ No |
| 3 vs 1 | 4.6 | −1.12 | 10.32 | 0.140 | ❌ No |
| 4 vs 1 | 12.6 | 6.88 | 18.32 | < 0.001 | ✅ Yes |
| 3 vs 2 | 5.8 | 0.08 | 11.52 | 0.046 | ✅ Yes |
| 4 vs 2 | 13.8 | 8.08 | 19.52 | < 0.001 | ✅ Yes |
| 4 vs 3 | 8.0 | 2.28 | 13.72 | 0.005 | ✅ Yes |

**How to interpret the Tukey plot:** A confidence interval that **does not cross zero** = significant difference. If the CI includes zero, the two groups could plausibly have the same mean.

**Summary grouping table (for exam reports):**

| Package | Mean Cases | Grouping |
|---|---|---|
| 1 | 14.6 | A |
| 2 | 13.4 | A |
| 3 | 19.2 | B |
| 4 | 27.2 | C |

*Groups sharing a letter are NOT significantly different from each other.*

**Full written report (exam-style):**

> A one-way analysis of variance indicated that package design has a significant effect on the number of cases of cereal sold, F(3, 16) = 19.56, MSE = 10.0, p < 0.001, η² = 0.786. Tukey's HSD test indicated that the mean number of cases sold was significantly different between all designs **except** designs 1 vs 2 (p = 0.931) and designs 1 vs 3 (p = 0.140). Package 4 had the highest mean sales (M = 27.2).

### Step 7: Q7(v) — The ANOVA Model

```{r}
#| label: kenton-model
summary.lm(model_kenton)
```

From the regression-style output:

$$\hat{y}_{ij} = \hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 + \hat{\beta}_3 x_3$$

$$\hat{y}_{ij} = 14.6 - 1.2x_1 + 4.6x_2 + 12.6x_3$$

**What each term means:**

| Term | Value | Meaning |
|---|---|---|
| $\hat{\beta}_0 = 14.6$ | Intercept | Mean cases sold for Package 1 (the reference group) |
| $\hat{\beta}_1 = -1.2$ | Package 2 effect | Package 2 sells 1.2 fewer cases on average than Package 1 |
| $\hat{\beta}_2 = 4.6$ | Package 3 effect | Package 3 sells 4.6 more cases on average than Package 1 |
| $\hat{\beta}_3 = 12.6$ | Package 4 effect | Package 4 sells 12.6 more cases on average than Package 1 |

**The dummy variable coding:**

| Group | $x_1$ | $x_2$ | $x_3$ |
|---|---|---|---|
| Package 1 | 0 | 0 | 0 |
| Package 2 | 1 | 0 | 0 |
| Package 3 | 0 | 1 | 0 |
| Package 4 | 0 | 0 | 1 |

> 📝 *Annotation space:*

---

## Q8 — Vitamin C in Multivitamins

### Question
Twelve measurements of Vitamin C content (mg) were taken for each of three multivitamin brands (A, B, C). Interpret the R output.

### Plain English Setup
We're asking: "Do these three vitamin brands actually deliver different amounts of Vitamin C, or is any difference just random variation?"

### Q8(i) — Hypotheses

$$H_0: \mu_A = \mu_B = \mu_C$$

"The mean Vitamin C content is the same across all three brands."

$$H_A: \text{At least one brand has a different mean Vitamin C content}$$

### Q8(ii) — Levene Test Hypotheses

$$H_0: \sigma^2_A = \sigma^2_B = \sigma^2_C$$

"The variance in Vitamin C content is equal across all three brands."

$$H_A: \text{At least one brand has different variance}$$

### Q8(iii) — Interpreting the Levene Test

**From the output:**
```
Levene's Test: Df F value Pr(>F)
group   2    0.6815   0.5128
       33
```

**Interpretation:** p = 0.5128 >> 0.05. We **fail to reject H₀**. There is not enough evidence to suggest that variances differ across brands. ✅ The homogeneity of variance assumption is met — we can proceed with ANOVA.

### Q8(iv) — Completing the ANOVA Table

The original output had missing values:
```
         Df   Sum Sq   Mean Sq   F value   Pr(>F)
variety   -   29.8      -         -        0.355
Residuals -   459.4     -
```

**Filling in the blanks:**

- **variety Df:** 3 groups − 1 = **2**
- **Residuals Df:** 36 total − 3 groups = **33**
- **variety Mean Sq:** 29.8 / 2 = **14.9**
- **Residuals Mean Sq:** 459.4 / 33 = **13.92**
- **F value:** 14.9 / 13.92 = **1.069** ✓ (matches the given value)

**Completed ANOVA table:**

| Source | Df | Sum Sq | Mean Sq | F value | p-value |
|---|---|---|---|---|---|
| variety | 2 | 29.8 | 14.9 | 1.069 | 0.355 |
| Residuals | 33 | 459.4 | 13.92 | | |

**Effect Size:**

$$\eta^2 = \frac{29.8}{29.8 + 459.4} = \frac{29.8}{489.2} \approx \mathbf{0.061}$$

A small effect (6% of variation explained by brand).

**Written report:**

> A one-way analysis of variance did not provide evidence to suggest that the mean Vitamin C content varied between multivitamin varieties, F(2, 33) = 1.069, MSE = 13.92, p = 0.355, η² = 0.061. Tukey's HSD test confirmed that no pairwise comparisons were significantly different at the 5% level.

### Q8(v) — ANOVA Model

$$\hat{y}_{ij} = 147.254 + 1.949x_1 + 1.907x_2 + e_{ij}$$

| Term | Value | Meaning |
|---|---|---|
| 147.254 | Intercept | Mean Vitamin C for Variety A (reference group) |
| 1.949 | Variety B effect | Variety B has 1.949 mg more Vitamin C than A (not significant) |
| 1.907 | Variety C effect | Variety C has 1.907 mg more Vitamin C than A (not significant) |
| $e_{ij}$ | Residual | The unexplained random variation |

> 📝 *Annotation space:*

---

## Q9 — Soft Drink End-Aisle Displays

### Question
Three end-aisle display designs were tested in 15 stores (5 per design). The response is percentage increase in sales. Perform a full analysis.

### Setup

```{r}
#| label: drinks-setup
drinks <- read_tsv("soft_drink.txt")
drinks$Display <- as.factor(drinks$Display)
glimpse(drinks)
```

### Q9(i) — Exploratory Data Analysis

```{r}
#| label: drinks-eda

# Summary statistics
drinks %>%
  group_by(Display) %>%
  summarise(
    mean_Increase = mean(Increase),
    sd_Increase   = sd(Increase),
    max_Increase  = max(Increase),
    min_Increase  = min(Increase),
    n             = n()
  )

# Boxplot
boxplot(drinks$Increase ~ drinks$Display,
        xlab = "Display Design",
        ylab = "% Increase in Sales",
        main = "Sales Increase by Display Design",
        col  = c("#AED6F1", "#A9DFBF", "#F9E79F"))
```

**Summary statistics table:**

| Design | Mean % Increase | SD | Max | Min | n |
|---|---|---|---|---|---|
| 1 | 5.73 | 0.388 | 6.22 | 5.29 | 5 |
| 2 | 6.24 | 0.434 | 6.71 | 5.66 | 5 |
| 3 | 8.32 | 0.670 | 9.20 | 7.55 | 5 |

**How to interpret the boxplot:** Display 3 clearly has higher median sales. Displays 1 and 2 look similar. No obvious outliers. This already suggests Display 3 might be the winner.

**Check for equal variances:**

```{r}
#| label: drinks-fligner
fligner.test(drinks$Increase ~ drinks$Display)
```

> Fligner-Killeen: p = 0.3601. Since p > 0.05, variances are equal across groups. ✅

### Q9(ii) — ANOVA Table

```{r}
#| label: drinks-anova
model_drinks <- aov(drinks$Increase ~ drinks$Display)
summary(model_drinks)
```

**ANOVA Table:**

| Source | Df | Sum Sq | Mean Sq | F value | p-value |
|---|---|---|---|---|---|
| Display | 2 | 18.78 | 9.392 | 35.77 | < 0.001 |
| Residuals | 12 | 3.15 | 0.263 | | |

**Interpreting this table:** F(2, 12) = 35.77. The F-statistic is very large — the between-group variation is 35× the within-group noise. p < 0.001 → very strong evidence that at least one display performs differently.

**Effect Size:**

$$\eta^2 = \frac{18.78}{18.78 + 3.15} = \frac{18.78}{21.93} \approx \mathbf{0.856}$$

**Massive effect** — 85.6% of variation in sales is explained by which display was used.

### Q9(iii) — Pairwise Comparisons (Tukey HSD)

```{r}
#| label: drinks-tukey
TukeyHSD(model_drinks)
plot(TukeyHSD(model_drinks))
```

**Tukey results:**

| Comparison | Difference | 95% CI | p adj | Significant? |
|---|---|---|---|---|
| 2 vs 1 | 0.506 | (−0.47, 1.48) | 0.358 | ❌ No |
| 3 vs 1 | 2.586 | (1.61, 3.56) | < 0.001 | ✅ Yes |
| 3 vs 2 | 2.080 | (1.11, 3.05) | 0.001 | ✅ Yes |

**Summary grouping table:**

| Design | Mean % Increase | SD | Group |
|---|---|---|---|
| 1 | 5.73 | 0.388 | A |
| 2 | 6.24 | 0.434 | A |
| 3 | 8.32 | 0.670 | B |

*Designs 1 and 2 are not significantly different from each other (same letter A), but both are significantly different from Design 3.*

### Q9(v) — ANOVA Model

```{r}
#| label: drinks-model
summary.lm(model_drinks)
```

$$\hat{y}_{ij} = 5.732 + 0.506x_1 + 2.586x_2$$

| Term | Meaning |
|---|---|
| 5.732 | Mean % increase for Display 1 (reference) |
| 0.506 | Display 2 sells 0.506% more than Display 1 (not significant) |
| 2.586 | Display 3 sells 2.586% more than Display 1 (significant) |

**Dummy variable coding:**

| Group | $x_1$ | $x_2$ |
|---|---|---|
| Display 1 | 0 | 0 |
| Display 2 | 1 | 0 |
| Display 3 | 0 | 1 |

### Q9(vi) — Model Diagnostics

```{r}
#| label: drinks-diagnostics
par(mfrow = c(2, 2))
plot(model_drinks)
```

**How to read each diagnostic plot:**

**1. Residuals vs Fitted:**
The points should be scattered randomly around the horizontal line at zero, with no obvious pattern. A U-shape or funnel would indicate a problem. Here the scatter looks random → ✅

**2. Normal Q-Q Plot:**
The points should follow the diagonal dashed line closely. If they curve away (especially at the ends), the normality assumption is violated. Here the points sit close to the line → ✅

**3. Scale-Location Plot:**
The red line should be roughly flat. An upward slope means higher fitted values have more variance (heteroscedasticity). Here it looks approximately flat → ✅

**4. Residuals vs Leverage:**
High-leverage points that also have large residuals are *influential* — they could be distorting results. Look for points outside the Cook's distance dashed lines. Here no points are outside → ✅

**Written report for Q9:**

> A one-way analysis of variance indicated that the end-aisle display design has a significant effect on the percentage increase in sales, F(2, 12) = 35.77, MSE = 0.263, p < 0.001, η² = 0.856. Tukey's HSD test indicated that Display 3 produced significantly higher sales increases than both Display 1 (p < 0.001) and Display 2 (p = 0.001), while Displays 1 and 2 were not significantly different from each other (p = 0.358). Model diagnostic plots confirmed that all ANOVA assumptions were satisfied.

> 📝 *Annotation space:*

---

## Q10 — Brick Firing Temperature

### Question
Four firing temperatures (100°C, 125°C, 150°C, 175°C) were tested to see if they affect brick density. Run a full analysis including handling an influential outlier.

### Setup

```{r}
#| label: bricks-setup
library(outliers)

bricks <- read_tsv("brick.txt")
bricks$Temperature <- as.factor(bricks$Temperature)
glimpse(bricks)
```

### Q10(i) — Exploratory Data Analysis

```{r}
#| label: bricks-eda

# Summary statistics
bricks %>%
  group_by(Temperature) %>%
  summarise(
    mean_Density = mean(Density),
    sd_Density   = sd(Density),
    var_Density  = var(Density),
    n            = n()
  )

# Boxplot
boxplot(bricks$Density ~ bricks$Temperature,
        xlab = "Firing Temperature (°C)",
        ylab = "Brick Density",
        main = "Brick Density by Firing Temperature",
        col  = c("#AED6F1", "#A9DFBF", "#F9E79F", "#F1948A"))
```

**Checking for outliers — Grubbs Test:**

```{r}
#| label: bricks-outlier

# Check overall dataset
grubbs.test(bricks$Density)

# Check within the 175°C group specifically
grubbs.test(bricks$Density[bricks$Temperature == "175"])
```

**Grubbs test interpretation:**
- Overall dataset: outlier is **not** significant (p > 0.05) — no individual value is unusual enough to be an outlier when looking at all temperatures together.
- Within the 175°C group only: the outlier **is** significant (p < 0.05) — observation 37 (density = 22.1) stands out as unusual within that temperature group.

> ⚠️ **Important:** The observation 22.1 at 175°C looks suspicious. It may be a data entry error or a genuine anomaly. Since we don't have additional information, we need to run the analysis **both with and without** this point.

**Checking equal variances:**

```{r}
#| label: bricks-fligner
fligner.test(bricks$Density ~ bricks$Temperature)
```

### Q10(ii) — ANOVA (Full Dataset)

```{r}
#| label: bricks-anova
model_bricks <- aov(bricks$Density ~ bricks$Temperature)
summary(model_bricks)
summary.lm(model_bricks)
```

### Q10(iii) — Pairwise Comparisons (Full Dataset)

```{r}
#| label: bricks-tukey
TukeyHSD(model_bricks)
plot(TukeyHSD(model_bricks))
```

### Q10(vi) — Model Diagnostics (Full Dataset)

```{r}
#| label: bricks-diagnostics
par(mfrow = c(2, 2))
plot(model_bricks)
```

**What to look for here:** Observation 37 (the 22.1 density at 175°C) will likely show up as a point of high influence. Check the "Residuals vs Leverage" plot — if it falls outside the Cook's distance lines, it is highly influential and the analysis should be rerun without it.

> The diagnostic plots reveal that **observation 37 is highly influential**. We should remove it and rerun.

### Reanalysis — Outlier Removed

```{r}
#| label: bricks-removed
bricks_rm <- bricks[-37, ]

# Check boxplots again
boxplot(bricks_rm$Density ~ bricks_rm$Temperature,
        xlab = "Firing Temperature (°C)",
        ylab = "Brick Density",
        main = "Brick Density by Temperature (Outlier Removed)",
        col  = c("#AED6F1", "#A9DFBF", "#F9E79F", "#F1948A"))

# Test variance equality again
fligner.test(bricks_rm$Density ~ bricks_rm$Temperature)

# Refit model
model_bricks_rm <- aov(bricks_rm$Density ~ bricks_rm$Temperature)
summary(model_bricks_rm)
summary.lm(model_bricks_rm)

# Tukey again
TukeyHSD(model_bricks_rm)
plot(TukeyHSD(model_bricks_rm))

# Diagnostics
par(mfrow = c(2, 2))
plot(model_bricks_rm)
```

### How to Write Up Both Results

> Since we have no additional information about the outlier (observation 37, density = 22.1 at 175°C), and it is a plausible value for brick density, we report both analyses:

> **With outlier:** A one-way ANOVA [report F, df, MSE, p, η²]. Tukey's HSD indicated [report which groups differ].

> **Without outlier:** Removing observation 37 changes the pairwise comparisons [describe how results change]. Both sets of results are presented as the outlier cannot be definitively excluded.

> 💡 **Exam tip:** When an outlier is influential, always (a) report it, (b) investigate it, (c) run the analysis both ways, and (d) explain what changes.

> 📝 *Annotation space:*

---

# Summary Points — Key Takeaways

- **Factor** = what you control. **Level** = the settings. **Treatment** = one specific combination.
- **Randomized block design** controls for a known nuisance variable by including it as a "block."
- **Multiple testing** inflates Type I error: P(at least one error) = 1 − (1 − α)^k.
- **ANOVA assumptions:** independence, equal variances (tested with Fligner or Levene), approximately normal residuals.
- **ANOVA hypotheses:** H₀: all group means equal. Hₐ: at least one differs.
- **F statistic** = MS_Treatment / MS_Residuals. Big F → more evidence against H₀.
- **Degrees of freedom:** Treatment = k − 1. Residuals = N − k.
- **η²** = SS_Treatment / SS_Total. Proportion of variation explained. Exam standard: report alongside F, df, MSE, p.
- **Tukey HSD:** controls family-wise error when doing all pairwise comparisons. Groups sharing a superscript letter are not significantly different.
- **The ANOVA model:** ŷ = β₀ + β₁x₁ + β₂x₂ + ··· where β₀ = mean of reference group and β coefficients = differences from reference group.
- **Diagnostics:** Check Residuals vs Fitted (patterns?), Q-Q plot (normality?), Scale-Location (constant variance?), Leverage plot (influential points?).
- **Influential outliers:** Always run the analysis with AND without them if they appear influential. Report both and explain.

---

# Quick Formula Reference

```{r}
#| label: formulas
#| eval: false

# η² (Effect Size)
eta_sq <- SS_Treatment / (SS_Treatment + SS_Residuals)

# Degrees of freedom
df_treatment <- k - 1       # k = number of groups
df_residuals <- N - k       # N = total observations

# Mean Squares
MS_Treatment <- SS_Treatment / df_treatment
MS_Residuals <- SS_Residuals / df_residuals

# F statistic
F_stat <- MS_Treatment / MS_Residuals

# Type I error (multiple tests)
P_at_least_one_error <- 1 - (1 - alpha)^num_tests
```

---

> ✏️ **General annotations — add your own notes, mnemonics, or diagrams below:**
```

