---
title: "Week 10: Model Identification — ACF & PACF"
subtitle: "STAT9005 — Time Series Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
---

---

::: {.callout-note}
## 📌 How to use this guide
This is the most **exam-critical** week of the module. The skill of reading ACF and PACF plots to identify ARIMA models is tested directly in both the MCQ and the practical exam. Every section ends with a self-check question — answer it before moving on.
:::

---

::: {.callout-important}
## ⭐ Assessment Reminders

- **Written test (MCQ / Short Answer) — 20%** — Week 10/11 — 45 minutes, ~20–25 questions, **−0.25 penalty per wrong answer**
- **Practical test (R/RStudio lab) — 40%** — Week 11/12 — open book, all R scripts available
- **Project & Presentation — 40%** — Report due **18 May 2026 at 23:59**

*Professor's note (Week 10): "I'm not going to give you two options that can be one or the other, or anything. It's very straightforward."* — ACF/PACF identification questions in the exam have clear, unambiguous answers.
:::

---

# What This Week Is About

At this point you know:

- How to check stationarity (ADF/KPSS tests)
- How to transform data (log, differencing)
- How to fit ARIMA and SARIMA models

**What was missing:** how do you decide what values of p and q to use? That's what this week answers. The ACF and PACF plots are your **diagnostic tools** for choosing p (AR order) and q (MA order).

---

# What Are ACF and PACF?

## ACF — AutoCorrelation Function

The **ACF** measures the correlation between your time series and a lagged version of itself.

In plain words: how similar is today's value to yesterday's? To the value from 2 periods ago? From 5 periods ago?

$$\hat{\rho}_k = \frac{\hat{\gamma}_k}{\hat{\gamma}_0} = \frac{\sum_{t=k+1}^{T}(y_t - \bar{y})(y_{t-k} - \bar{y})}{T}$$

Where $k$ is the **lag** (the time gap you are measuring).

::: {.callout-note}
## 📌 Analogy: Your mood and your past mood

If you are happy today, you are probably somewhat happy tomorrow (high autocorrelation at lag 1). But are you still happy 30 days later? Maybe not (low autocorrelation at lag 30). The ACF plot shows exactly how this correlation fades over time.
:::

## PACF — Partial AutoCorrelation Function

The **PACF** measures the correlation between your series and a lagged version — but **after removing the influence of all the lags in between**.

$$\hat{\phi}_{kk} = \hat{\rho}_{t, t-k \,|\, t-1, t-2, \ldots, t-k-1}$$

In plain words: the PACF at lag 3 asks "is today's value correlated with the value from 3 periods ago, *after I've already accounted for* the influence of lags 1 and 2?"

::: {.callout-note}
## 📌 The key difference between ACF and PACF

| | ACF | PACF |
|---|---|---|
| **Measures** | Total correlation at each lag | *Direct* correlation at each lag (indirect effects removed) |
| **Useful for** | Identifying MA order (q) | Identifying AR order (p) |
| **Intuition** | "How much of my past is still in my data?" | "Which specific past values directly predict today?" |
:::

---

# Reading the Plots

## The Dashed Lines (Significance Bands)

Both ACF and PACF plots have two dashed horizontal lines. These are the **95% confidence bands**. Any bar that extends **beyond** these lines represents a statistically significant correlation at that lag.

A bar **inside** the bands → not significant (likely random chance).
A bar **outside** the bands → significant correlation at that lag.

::: {.callout-important}
## ⭐ Exam tip: A few bars outside the bands is normal

*Professor's note: "There are a few values outside the barely lines. I agree with you. But there is no pattern — this might be due by chance. If I observe a pattern: 1, 2, 3, 4, 5… every 5 you observe a pattern — then there is a problem. But here, no pattern."*

**The key is PATTERN, not individual bars.** If 2 out of 30 bars are barely outside the lines with no pattern, that is still consistent with white noise (you expect ~5% outside by chance at 95% confidence).
:::

## What White Noise Looks Like

If your data (or residuals) are **white noise** — pure random noise with no structure — both ACF and PACF should show:

- All bars within the significance bands
- No pattern (no regular spikes, no slow decay)

This is what you *want* to see in your **residuals** after fitting a model. If you see this, your model has captured all the structure.

---

# The Four Identification Patterns

::: {.callout-important}
## ⭐ The Rules to Memorise — These Appear in the Exam

| What you see | What it means | ARIMA model |
|---|---|---|
| ACF: **slow decay** → PACF: **cuts off sharply** after lag p | Autoregressive process | **AR(p)** |
| ACF: **cuts off sharply** after lag q → PACF: **slow decay** | Moving average process | **MA(q)** |
| ACF: **slow decay** → PACF: **slow decay** | Mixed process | **ARMA(p,q)** |
| ACF: **very slow decay** (all bars high, barely reducing) | Non-stationary — needs differencing | **Apply diff()** first |

The "slow decay" in ACF for AR means the bars gradually get smaller — still significant for many lags. The "cuts off" in PACF means suddenly drops to zero (inside the significance bands) after lag p.
:::

---

# Each Model Type Explained

## AR(p) — Autoregressive

**What it is:** Today's value is a function of p past values.

$$y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \cdots + \phi_p y_{t-p} + \varepsilon_t$$

**What the plots look like:**

- **ACF:** Slow, gradual decay — still significant at many lags
- **PACF:** Sharp cut-off after lag p — drops inside the bands at lag p+1

**How to find p:** Count the number of significant bars in the PACF before it cuts off. If only lag 1 is significant in the PACF, then p = 1.

::: {.callout-note}
## 📌 Analogy: Learning from your commute history

AR(1): "Today's commute = some fraction of yesterday's commute + random noise."
AR(2): "Today's commute = some fraction of yesterday's + some fraction of 2 days ago + noise."

The more past days you need, the higher the p.
:::

*Professor's note: "Autoregressive — slow reduction in ACF. Drop after one period. PACF: after 1, the next one is not significant. So: autoregressive is slow reduction, with P equal 1. Because after P equal 2 is not significant. And with value equals 0.9."*

```{r}
#| eval: false

library(astsa)

# Simulate an AR(1) process
ar1_sim <- arima.sim(list(order = c(1, 0, 0), ar = 0.9), n = 200)
plot(ar1_sim, main = "AR(1) simulation")

# Plot ACF and PACF together
acf2(ar1_sim, main = "AR(1): slow decay in ACF, sharp cut-off in PACF at lag 1")
```

---

## MA(q) — Moving Average

**What it is:** Today's value depends on q past *forecast errors* (shocks), not past values.

$$y_t = \varepsilon_t + \theta_1 \varepsilon_{t-1} + \theta_2 \varepsilon_{t-2} + \cdots + \theta_q \varepsilon_{t-q}$$

**What the plots look like:**

- **ACF:** Sharp cut-off after lag q — drops inside the bands at lag q+1
- **PACF:** Slow, gradual decay — still significant for many lags

**How to find q:** Count the significant bars in the ACF before it cuts off. If only lag 1 is significant in ACF, then q = 1.

::: {.callout-note}
## 📌 The mirror image of AR

Notice that MA is the exact **opposite** of AR in terms of which plot shows the cut-off:

- AR → PACF cuts off (use PACF to find p)
- MA → ACF cuts off (use ACF to find q)
:::

*Professor's note: "The moving average — we have slow reduction in the PACF and drop after so many values. So we have moving average, and drop after 8."*

```{r}
#| eval: false

# Simulate an MA(1) process
ma1_sim <- arima.sim(list(order = c(0, 0, 1), ma = -0.5), n = 200)
acf2(ma1_sim, main = "MA(1): sharp cut-off in ACF at lag 1, slow decay in PACF")
```

---

## ARMA(p,q) — Mixed

**What it is:** A combination of both AR and MA components.

**What the plots look like:**

- **ACF:** Slow decay
- **PACF:** Slow decay

When *both* plots show slow decay, neither alone cuts off cleanly — this means you need both AR and MA terms. The exact values of p and q are harder to read visually and usually require trial and error (or `auto.arima()`).

*Professor's note: "ARMA — a combination of both. We have a slow reduction in the ACF. And slow reduction in the PACF."*

---

## Non-Stationary Series (Needs Differencing)

**What it looks like in the ACF:**

- ACF bars are **very high** across many lags (near 1.0 for early lags)
- The reduction is **extremely slow** — the series "remembers" its past for a very long time
- PACF shows a spike at lag 1 only

This pattern means the series has a **trend** or **random walk** behaviour — it is not stationary and must be differenced before you can identify p and q.

::: {.callout-important}
## ⭐ Exam tip: Long memory in ACF = non-stationary = need differencing

*Professor's note: "We have a long memory. A very tiny reduction — so there is a huge correlation between my observations and faraway observations. This means that we need some transformations. Most likely: you have to take differences. For sure it means you have a trend, or you have a behaviour that simulates a random walk."*

Once you difference the data and re-plot ACF/PACF, the bars should fall within the significance bands (if white noise) or show a recognisable AR/MA pattern.
:::

---

# The Complete Identification Workflow

This is the process to follow every time you have a (stationary) dataset and need to choose p and q:

```
Step 1: Make sure the series is stationary
        → ADF test (p < 0.05 = stationary)
        → If not: apply diff() and/or log(), then retest

Step 2: Plot acf2() on the stationary series

Step 3: Look at PACF first
        → Does it cut off sharply after lag p?
           Yes → Suggests AR(p)
        → Does it decay slowly?
           Yes → Not purely AR

Step 4: Look at ACF
        → Does it cut off sharply after lag q?
           Yes → Suggests MA(q)
        → Does it decay slowly?
           Yes → Not purely MA

Step 5: Combine your reading
        → PACF cuts off, ACF decays → AR(p)
        → ACF cuts off, PACF decays → MA(q)
        → Both decay slowly → ARMA(p,q)

Step 6: Use auto.arima() as a confirmation / starting point
        → Compare your manual reading to R's suggestion
        → Check the AICc values

Step 7: Validate with checkresiduals()
        → If residuals look like white noise → model is good
        → If pattern remains → try different p, q values
```

---

# The acf2() Function — Your Main Tool

In this module, the preferred function is `acf2()` from the `astsa` package. It plots ACF and PACF side by side in one call.

```{r}
#| eval: false

library(astsa)
library(tseries)
library(forecast)

# Plot ACF and PACF together
acf2(my_stationary_series, max.lag = 30)

# If using base R separately:
par(mfrow = c(2, 1))
acf(my_stationary_series, lag.max = 30)
pacf(my_stationary_series, lag.max = 30)
```

---

# Worked Example: AirPassengers

The AirPassengers dataset is the main teaching example for SARIMA in this module. Here is the full identification workflow:

```{r}
#| eval: false

library(astsa)
library(tseries)
library(forecast)

AP <- AirPassengers

# Step 1: Plot the original series
autoplot(AP, main = "AirPassengers")
# Observation: clear upward trend AND growing seasonal swings
# → non-constant mean (needs differencing)
# → non-constant variance (needs log transformation)

# Step 2: Log transformation to stabilise variance
logAP <- log(AP)
autoplot(logAP, main = "log(AirPassengers)")
autoplot(decompose(logAP))

# Step 3: Check stationarity
adf.test(logAP, alternative = "stationary", k = 0)
kpss.test(logAP)
# Both tests confirm non-stationary → need differencing

# Step 4: First difference to remove trend
dif_LAP <- diff(logAP)
autoplot(dif_LAP, main = "D(log(AirPassengers))")
adf.test(dif_LAP)  # Re-check stationarity

# Step 5: Seasonal differencing (lag = 12 for monthly data)
D12dif_LAP <- diff(dif_LAP, lag = 12)
autoplot(D12dif_LAP)
adf.test(D12dif_LAP)  # Should now be stationary

# Step 6: Plot ACF and PACF on the fully transformed series
acf2(D12dif_LAP)
# Reading: ACF cuts off after lag 1, PACF decays → MA(1) in non-seasonal part
# At seasonal lags (12, 24, 36): similar patterns → SARIMA(0,1,1)(0,1,1)12

# Step 7: Auto.arima confirmation
auto.arima(log(AP))
# R suggests: ARIMA(0,1,1)(0,1,1)[12]

# Step 8: Fit and check residuals
fit <- auto.arima(log(AP))
checkresiduals(fit)
# Looking for: residuals ~white noise, Ljung-Box p > 0.05

# Step 9: Forecast (undo log transformation)
fc <- forecast(fit, h = 24)
fc$mean  <- exp(fc$mean)
fc$lower <- exp(fc$lower)
fc$upper <- exp(fc$upper)
fc$x     <- exp(fc$x)
plot(fc, main = "AirPassengers 24-month forecast")
```

::: {.callout-important}
## ⭐ Why auto.arima(log(AP)) not auto.arima(AP)?

*Professor's note: "auto.arima(AirPassengers) — this is WRONG. auto.arima(log(AirPassengers)) — this helps the ARIMA to stabilise the variance."*

Always apply the log transformation manually first, then run `auto.arima()`. If you skip the log, ARIMA is trying to work with data that violates the constant variance assumption, and the model quality will be poor.
:::

---

# Identifying SARIMA Patterns

For seasonal data (frequency > 1), you read the ACF/PACF at **two levels**:

1. **Non-seasonal lags** (lags 1, 2, 3…) → determine p and q
2. **Seasonal lags** (lags at multiples of m: 12, 24, 36 for monthly data) → determine P and Q

::: {.callout-important}
## ⭐ Reading seasonal correllograms

*Professor's note: "In this second part, you might think that there is seasonality — but there is no repetition here in 2, in 3, in 4. So we have removed this correlation. What happened here is a SARIMA. The correlations have a smaller effect, so things are more difficult to identify. What I have here is: slow reduction, ACF drop after 1 — autoregressive one in the seasonal part. The opposite: slow reduction, moving average, drop after one — moving average one."*

The seasonal part is read **exactly the same way** as the non-seasonal part — just looking at the spikes at lags 12, 24, 36 instead of 1, 2, 3. If the seasonal ACF cuts off after lag 12, Q = 1.
:::

| Pattern at seasonal lags | Interpretation |
|---|---|
| ACF spike only at lag m (12), then nothing | Seasonal MA(1): Q = 1 |
| PACF spike only at lag m (12), then nothing | Seasonal AR(1): P = 1 |
| ACF spikes at m, 2m, 3m; PACF spike at m only | Seasonal AR(1): P = 1 |
| ACF spike at m only; PACF at m, 2m, 3m | Seasonal MA(1): Q = 1 |

---

# auto.arima() — How It Works

`auto.arima()` systematically tries many combinations of p, d, q (and P, D, Q for seasonal data) and picks the one with the lowest **AICc**.

```{r}
#| eval: false

library(forecast)

# Basic usage — let R decide everything
best_model <- auto.arima(my_ts)
summary(best_model)

# With restrictions
auto.arima(my_ts, d = 1, D = 0, seasonal = FALSE)  # Force non-seasonal
auto.arima(my_ts, d = 1, D = 1, seasonal = TRUE)   # Force seasonal differencing

# The output tells you:
# ARIMA(p,d,q)(P,D,Q)[m]
# For example: ARIMA(0,1,1)(0,1,1)[12] means:
#   p=0, d=1, q=1 (non-seasonal)
#   P=0, D=1, Q=1 (seasonal, period 12)
```

::: {.callout-note}
## 💡 Use auto.arima() as a starting point, not a final answer

*Professor's note: "After we've gone through all these confusing ways of identifying the datasets, identifying parameters, analyzing correlograms and all these things — there is an option in R called auto.arima. And this auto.arima is giving you a possible option, according to R, the best model."*

The workflow in the exam is: identify by eye from correlograms → confirm with auto.arima → validate with checkresiduals.
:::

---

# sarima() and checkresiduals() — Fitting and Validating

## sarima() from the astsa package

```{r}
#| eval: false

library(astsa)

# Fit and see full diagnostic output at once
sarima(log(AirPassengers), 
       p = 0, d = 1, q = 1,    # non-seasonal
       P = 0, D = 1, Q = 1,    # seasonal
       S = 12)                  # period

# sarima() automatically plots:
# 1. Standardised residuals over time
# 2. ACF of residuals
# 3. Q-Q plot of residuals
# 4. Ljung-Box p-values at each lag
```

## checkresiduals() from the forecast package

```{r}
#| eval: false

library(forecast)

fit <- auto.arima(log(AirPassengers))
checkresiduals(fit)

# Produces:
# 1. Residual time plot
# 2. ACF of residuals
# 3. Histogram of residuals
# Plus: Ljung-Box test result

# What you want to see:
# - Residuals scattered randomly around zero (time plot)
# - All ACF bars within significance bands (no remaining structure)
# - Roughly normal histogram
# - Ljung-Box p-value > 0.05
```

::: {.callout-important}
## ⭐ The Ljung-Box test for residuals

H₀: residuals are **uncorrelated** (white noise)

- **p > 0.05** → fail to reject H₀ → residuals are white noise → **good model** ✅
- **p < 0.05** → reject H₀ → residuals still have structure → **model needs improvement** ❌

*Professor's note: "Random, not correlated, normal, independent — this is a good model. So I have a model that I can use for prediction."*
:::

---

# Exam-Style Question: Reading a Correlogram

This question type appears directly in the sample MCQ paper.

::: {.callout-important}
## ⭐ Sample Exam Question (from the actual sample paper)

**Question 7:** According to the following ACF and PACF of a time series, identify what type of ARIMA(p,d,q) process best describes the series.

Options: A. MA(1)  B. ARIMA(2,1,0)  C. AR(1)  D. AR(2)

**How to approach this:**

1. Look at the ACF: does it cut off sharply (→ MA) or decay slowly (→ AR or need differencing)?
2. Look at the PACF: does it cut off sharply (→ AR) or decay slowly (→ MA)?
3. Check for a d component: does the ACF show very slow decay (almost no reduction)? → non-stationary → d ≥ 1
4. Count the significant bars after the cut-off to determine p or q

*Professor's note (on this question): "We have slow reduction in ACF. Drop after one period. Okay, so autoregressive. D equal 1 means that we have to take differences to remove the trend. I don't see trend here. I don't see trend in this ACF. If we have trend, we have long memory — very long memory. You can see clearly that we have an auto-aggressive one."*

**Answer: C — AR(1)**

The ACF shows slow decay (characteristic of AR), the PACF cuts off sharply after lag 1 (p = 1). There is no evidence of a trend (no extremely slow ACF decay), so d = 0. Therefore ARIMA(1,0,0) = AR(1).
:::

---

# Summary: The Pattern Table (Memorise This)

| Model | ACF | PACF | Intuition |
|---|---|---|---|
| **White noise** | All bars inside bands | All bars inside bands | No structure — random |
| **AR(1)** | Slow exponential decay | Sharp cut-off after lag 1 | Today depends on yesterday |
| **AR(2)** | Slow sinusoidal decay | Sharp cut-off after lag 2 | Today depends on 2 past values |
| **MA(1)** | Sharp cut-off after lag 1 | Slow decay | Reacts to 1 past error |
| **MA(q)** | Sharp cut-off after lag q | Slow decay | Reacts to q past errors |
| **ARMA(p,q)** | Both decay slowly | Both decay slowly | Mix of AR and MA |
| **Non-stationary** | Very slow decay (all bars high) | Spike at lag 1 only | Has trend/drift — needs diff() |

---

# R Code Summary for Model Identification

```{r}
#| eval: false

library(astsa)
library(tseries)
library(forecast)

# ---- 1. Check stationarity ----
adf.test(my_ts)     # H0: non-stationary → want p < 0.05
kpss.test(my_ts)    # H0: stationary → want p > 0.05

# ---- 2. Transform if needed ----
log_ts   <- log(my_ts)          # stabilise variance
diff_ts  <- diff(my_ts)         # remove trend (d = 1)
sdiff_ts <- diff(diff_ts, lag = 12)  # seasonal diff (D = 1)

# ---- 3. Plot correlograms ----
acf2(diff_ts)          # ACF and PACF side by side
acf(diff_ts, lag = 30)
pacf(diff_ts, lag = 30)

# ---- 4. Auto identification ----
auto.arima(log(my_ts))   # Remember: log first for variance issues!

# ---- 5. Manual fit ----
fit <- arima(my_ts, order = c(1, 1, 0))         # ARIMA(1,1,0)
fit_s <- arima(my_ts,                             # SARIMA
               order = c(0, 1, 1),
               seasonal = list(order = c(0, 1, 1), period = 12))

# ---- 6. Validate ----
checkresiduals(fit)
Box.test(residuals(fit), lag = 20, type = "Ljung-Box")

# ---- 7. Forecast ----
fc <- forecast(fit, h = 12)
autoplot(fc)
```

---

# Summary Checklist

- ✅ I know what ACF measures (total correlation at each lag)
- ✅ I know what PACF measures (direct correlation at each lag, with shorter lags removed)
- ✅ I can identify the four main patterns in ACF/PACF plots
- ✅ I know: **AR → PACF cuts off** (use PACF to find p)
- ✅ I know: **MA → ACF cuts off** (use ACF to find q)
- ✅ I know: **both decay slowly → ARMA**
- ✅ I know: **very slow ACF decay → non-stationary → diff()**
- ✅ I understand that a few bars outside the significance bands is normal (random chance)
- ✅ I can use `acf2()`, `auto.arima()`, `checkresiduals()` in R
- ✅ I know that `auto.arima(log(AirPassengers))` is correct; `auto.arima(AirPassengers)` is wrong
- ✅ I can interpret the Ljung-Box test result on residuals (want p > 0.05)
- ✅ I can read seasonal correllograms at lags 12, 24, 36 for SARIMA models

---

# Pre-Topic Checklist

Before the practical exam, make sure you can:

- [ ] Look at an ACF/PACF plot and identify the model type in under 60 seconds
- [ ] Explain why you look at PACF for AR and ACF for MA
- [ ] Run the full workflow: `adf.test()` → `log()` + `diff()` → `acf2()` → `auto.arima()` → `checkresiduals()` → `forecast()`
- [ ] Interpret `checkresiduals()` output and say whether the model is acceptable
- [ ] Explain what `ARIMA(0,1,1)(0,1,1)[12]` means in plain words

---

# Practice Questions

::: {.callout-important}
## ⭐ Q1 — Pattern identification

You plot ACF and PACF on a stationary series. The ACF shows an exponential decay that is still significant at lags 1, 2, 3, 4, 5 (all above the significance bands) and the PACF shows only lag 1 is significant. What model do you suggest?

**Answer:** AR(1). The ACF decays slowly (characteristic of AR), and the PACF cuts off sharply after lag 1 — meaning p = 1. Since the series is already stationary, d = 0. Model: ARIMA(1,0,0).
:::

::: {.callout-important}
## ⭐ Q2 — Pattern identification

You plot ACF and PACF. The ACF shows a single large spike at lag 1 and nothing significant beyond that. The PACF shows oscillating, gradually decaying values. What model do you suggest?

**Answer:** MA(1). The ACF cuts off sharply after lag 1 (q = 1), and the PACF decays slowly — the mirror-image signature of a moving average process. Model: ARIMA(0,0,1).
:::

::: {.callout-important}
## ⭐ Q3 — Non-stationarity detection

You run acf2() on a raw dataset before any transformations. All 30 bars of the ACF are significant and barely decrease from left to right (all near 0.9–1.0). The PACF has one very large spike at lag 1 and nothing else. What does this tell you and what do you do?

**Answer:** This is the signature of a non-stationary series — very long memory in the ACF. The series has a trend or random walk behaviour. You must apply `diff()` to the data (first-order differencing, d = 1) and then re-plot the ACF/PACF on the differenced series before trying to identify p and q.
:::

::: {.callout-important}
## ⭐ Q4 — SARIMA reading

After transforming AirPassengers with `diff(log(AP), lag=12)`, you plot `acf2()`. At the non-seasonal lags (1, 2, 3…) you see: ACF cuts off after lag 1, PACF decays slowly. At the seasonal lags (12, 24, 36) you see: ACF has a spike at lag 12 only (then nothing), PACF has a spike at lag 12 only (then nothing). What SARIMA model would you try first?

**Answer:** The non-seasonal part suggests MA(1) (ACF cuts off, PACF decays) → q = 1, p = 0. The seasonal part — spike in both ACF and PACF at lag 12 suggests trying Q = 1, P = 0 or P = 1. Since we applied d = 1 and D = 1 (seasonal difference at lag 12), the model is `SARIMA(0,1,1)(0,1,1)[12]`. This is exactly what `auto.arima(log(AirPassengers))` returns.
:::

---

# Additional Resources

- 📘 [Forecasting: Principles and Practice — Chapter 9 (ARIMA)](https://otexts.com/fpp3/arima.html)
- 📘 [Forecasting: Principles and Practice — Chapter 9.5 (ACF/PACF)](https://otexts.com/fpp3/non-seasonal-arima.html)
- 📘 [Forecasting: Principles and Practice — Chapter 9.8 (Seasonal ARIMA)](https://otexts.com/fpp3/seasonal-arima.html)

---

# Annotations Space

*Use this space for your own notes as you practice reading correlograms.*

---

## Patterns I find easy:

> *...*

---

## Patterns I still find confusing:

> *...*

---

## Examples I practised on:

> *Write the dataset name, what the ACF/PACF looked like, and what model you identified...*

---

*End of Week 10 Study Guide — STAT9005 Time Series Analysis*
