---
title: "Week 1: Introduction to Data Mining & Machine Learning"
subtitle: "STAT9013 — Data Mining & Machine Learning"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "📚 Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show R code"
    highlight-style: github
    smooth-scroll: true
    df-print: paged
---

---

::: {.callout-important}
## ⚠️ Assessment & Course Admin Reminders

- **Assignments are 100% practical.** You will receive a dataset and must complete a machine learning task using R.
- **Assignment datasets are unique.** Gen AI tools trained on public data will not recognise them — relying on Gen AI to solve the whole task will not work.
- **Gen AI policy:**
  - ✅ **OK:** Using Gen AI for a small helper task (e.g. "how do I split a dataset into train and test in R?")
  - ❌ **NOT OK:** Using Gen AI to write major parts of a solution.
- **Week 2 note:** Week 2 class is **pre-recorded** (lecturer on holidays). The Zoom recording will be posted in Canvas by Friday midnight before Week 2.
- **Where to find material:** Slides are uploaded to the **Discussions page on Canvas** each week. The R code is uploaded in a **separate file** — the lecturer strongly encourages running it on your own machine.
- **Lab format:** The last part of each 3-hour lecture is a **live lab**. Follow along in RStudio while the professor codes.
- **Feedback is welcome:** The lecturer explicitly asked for emails with comments, criticism, or topics you want covered more deeply.
:::

---

# Quick Revision Cheat-Sheet

If you have only five minutes before an exam, read this section.

| Big Idea | One-Line Summary |
|----------|------------------|
| **AI ⊃ ML ⊃ Deep Learning** | Nested hierarchy. Data Mining ≈ Machine Learning (older name). |
| **Three Types of ML** | Supervised (with labels), Unsupervised (no labels), Reinforcement (reward/penalty). |
| **Supervised splits into** | Classification (predict a **category**) and Regression (predict a **number**). |
| **CRISP-DM (6 phases)** | Business → Data Understanding → Data Prep → Modelling → Evaluation → Deployment. **Iterative, not linear.** |
| **Traditional software vs ML** | Deterministic (fixed rules) vs Stochastic (learned from data). |
| **R-squared (R²)** | 0 to 1. Measures how much variation the model explains. |
| **Correlation (r)** | −1 to +1. Strength and direction of a linear relationship. R² = r². |
| **Gradient Descent** | The "nudge left, check, nudge right" algorithm behind most ML training. |
| **Accuracy trap** | A 99.999% accurate fraud model may catch **zero** fraud. Always check business impact. |
| **ML bias** | The algorithm is a pattern-matcher — bias lives in the **data**, not the algorithm. |

---

# Key Topics This Week

This week covers the big-picture foundations of the **entire course**. Every topic below will reappear in later weeks — treat this as your reference point.

- What is Data Mining, Machine Learning, Artificial Intelligence, and Deep Learning — and how do they relate?
- Real-world industry applications of AI/ML
- How data scientists manage ML projects (the **CRISP-DM** workflow)
- Ethics and fairness in Machine Learning
- The three types of Machine Learning: Supervised, Unsupervised, Reinforcement
- Supervised Learning algorithms (classification vs. regression)
- A complete worked example: Linear Regression in R
- Key statistical concepts: R-squared, correlation, gradient descent
- A **preview of Weeks 2–12** so you know where this course is going

---

# What Is Data Mining? How Does It Relate to AI?

## The Four Core Terms

Understanding the relationship between these four terms is **the single most important concept from Week 1**. They are not the same thing — they are nested inside each other like Russian dolls.

::: {.callout-note}
## 🪆 Analogy: Russian Dolls

Think of it like nesting dolls:

- The **biggest** doll is **Artificial Intelligence** — the broadest idea.
- Inside it sits **Machine Learning** — a specific approach to building AI.
- Inside Machine Learning sits **Deep Learning** — a specific family of ML algorithms.
- **Data Mining** is an older term that sits roughly where Machine Learning sits — consider them interchangeable in modern usage.
:::

| Term | Plain-Language Definition | Example |
|------|--------------------------|---------|
| **Artificial Intelligence (AI)** | When a computer does something that normally requires human intelligence | Google Maps route planning; a chess program; a self-driving car |
| **Machine Learning (ML)** | Using statistical algorithms so machines improve automatically with more data and experience | A spam filter that gets better the more emails it sees |
| **Data Mining** | Extracting patterns and rules from large amounts of data. An older term — treat as interchangeable with ML today | Finding which supermarket products are bought together |
| **Deep Learning** | A family of ML algorithms that works especially well for images and text | Face recognition; ChatGPT; image generation |

*Professor's note: "Data mining — extracting rules from large quantities of data — is an older term, rarely used actually anymore in the industry. I consider it interchangeable with machine learning."*

## How They Fit Together — The Nested Diagram

```
┌─────────────────────────────────────────┐
│          Artificial Intelligence        │
│  ┌───────────────────────────────────┐  │
│  │        Machine Learning           │  │
│  │  ┌─────────────────────────────┐  │  │
│  │  │       Deep Learning         │  │  │
│  │  └─────────────────────────────┘  │  │
│  └───────────────────────────────────┘  │
└─────────────────────────────────────────┘
```

**Concrete example — Self-driving car:**

- The whole car = **AI** (doing something human-like: driving)
- Its vision system = **Deep Learning** (recognising objects, lanes, pedestrians)
- That vision system is also **ML** because Deep Learning is a subset of ML

::: {.callout-important}
## ⭐ Exam Tip: Don't Confuse Gen AI with the Whole of ML

*Professor's note: "Don't be fooled with the Gen AI boom. Gen AI is a whole other massive area of artificial intelligence. There are wholesale businesses out there that are using machine learning every day of the week, and Gen AI is not involved in it."*

**Gen AI (like ChatGPT) is just one small part of a very large field.** Most ML in industry is about **tabular data**, predictions, and classification — not text generation. This course focuses on tabular ML.
:::

## Deterministic vs. Stochastic Systems

This is a key distinction that separates traditional software from machine learning.

| Feature | Traditional Software (Deterministic) | Machine Learning (Stochastic) |
|---------|--------------------------------------|-------------------------------|
| **Rules** | Programmer writes all the rules | Algorithm **learns** rules from data |
| **Same input → same output?** | Always yes | Can change as model updates |
| **Handles new patterns?** | Only if programmer adds new rules | Yes — learns automatically |
| **Can improve without code change?** | No | Yes (retrain on fresh data) |
| **Example** | A spreadsheet formula `=A1*B1` | A fraud detection model |
| **Failure mode** | Crashes / wrong output | Quietly produces biased or confident-but-wrong predictions |

::: {.callout-note}
## 💡 Analogy: Recipe vs. Cooking Lessons

- **Traditional software** is like handing someone a precise recipe: *"add exactly 200g of flour."* Same ingredients → same cake, every time.
- **Machine Learning** is like giving someone cooking lessons. They practise, taste, adjust. Over time they develop judgement. Two different trained cooks given the same ingredients may produce slightly different cakes — because they have **learned** different intuitions from **different training**.
:::

::: {.callout-warning}
## ⚠️ Common Misconception: "ML is magic" / "ML is objective"

Many non-technical stakeholders believe ML is either sorcery or flawless objective truth. **It is neither.**

- It is **maths** — specifically, statistics and linear algebra applied to patterns in data.
- It is **only as good as the data** you feed it.
- It is **pattern-matching**, not reasoning. A model that has never seen a situation will make poor predictions about it.
:::

---

# AI Applications in Industry

The professor spent considerable time on this — understanding **where** ML is applied helps you understand **why** certain techniques matter.

## Summary Across All Industries

| Industry | Most Common ML Problems |
|----------|------------------------|
| **Marketing** | Churn prediction, customer segmentation, recommendation engines, cross-sell/upsell, survey analysis |
| **Finance & Banking** | Fraud detection, credit risk, algorithmic trading, document extraction |
| **Manufacturing** | Predictive maintenance, anomaly detection on production lines, supply & demand forecasting |
| **Telecoms** | Churn, speech recognition for customer service, cell-tower siting |
| **Pharmaceuticals** | Drug discovery, clinical trial analysis, radiology imaging, outbreak prediction |

Notice how **churn** and **anomaly detection** appear in multiple industries. These are *general-purpose* ML tasks — the same technique solves different business problems. Most of Weeks 3–10 builds up the tools to tackle them.

## Marketing — Deep Dive

| Application | What It Does | Why It Matters |
|-------------|-------------|----------------|
| **Customer survey analysis** | Uses NLP to categorise written feedback into topics | Automates work that would take humans weeks |
| **Churn prediction** | Identifies customers likely to leave 1–2 months before they decide | By the time someone cancels, it's too late — you need to **predict it early** |
| **Cross-sell / upsell** | Recommends *related* products, not the same product again | A customer who just bought a TV won't buy another TV — offer them a soundbar |
| **Customer segmentation** | Groups customers by behaviour, lifestyle, or spending | Enables targeted marketing instead of mass campaigns |

::: {.callout-important}
## ⭐ Exam-Worthy Concept: Churn Prediction Is Predictive, Not Reactive

*Professor's note: "By the time a customer has decided they're going to opt out or unsubscribe, it's probably too late to change their mind. You really need to look at their behaviour maybe a month or two beforehand. So it gets clever — you're not asking 'who will leave in January', you're asking in November, 'who are the likely customers that will leave in the next 6 to 8 weeks?'"*

**The key idea:** you must build a model that predicts churn *before* it happens, so the business has time to intervene (e.g. discount, special offer).
:::

::: {.callout-note}
## 📖 The Tesco Clubcard Story — A Classic Data Mining Case Study

*Professor's note: "They went in to do a presentation to the board of Tesco's. The chairman waited maybe 30 seconds after the presentation — they thought they'd blown it. And he said: 'I've worked in this industry for over 30 years. I've worked in Tesco's for over 20 years, and it scares me that you know more about my customers than I do.'"*

In the early 1990s, Tesco launched the Clubcard. A data analytics company (**Dunnhumby**) used **customer segmentation** on the transaction data to group customers — e.g. "double income, no kids", "young families", "early retirees". Instead of sending every customer the same vouchers, Tesco could target each segment, and could tell suppliers: *"We can reach your exact target demographic."*

The result was so powerful that **Tesco bought a controlling share in Dunnhumby** to stop competitors accessing the same capability.

**The technique used:** Clustering / Segmentation (covered in Week 6).
:::

::: {.callout-note}
## ☕ The Petrol Station Coffee Story

*Professor's note: "If somebody comes in for €50 worth of fuel, but the person buys a coffee — the company makes as much money from the cup of coffee as they do from the entire €50 worth of fuel. The fuel itself has a very low margin, maybe 5%. If you sell €50 of fuel and make 5% profit, you have €2.50. But they can make €2.50 just selling a coffee."*

A petrol station the lecturer worked with used loyalty data to spot customers who came in for fuel but never bought food or coffee. By targeting them with specific promotions, the business significantly increased the sale of **high-margin** items. The key insight: **all customers are not equally profitable, and the same customer at different points in time isn't equally profitable either.**
:::

## Finance & Banking — Deep Dive

| Application | Notes |
|-------------|-------|
| **Fraud detection** | One of the earliest and most successful ML applications. Contactless payments only work because fraud is kept below ~0.001% |
| **Credit risk** | Automates loan decisions — but **legal restrictions apply**: race, gender, etc. cannot be used as features |
| **Automated trading** | Well over 99% of stock exchange transactions are executed by algorithmic bots, not humans |
| **Document processing** | AI/GenAI extracts data from mortgage applications, bank statements, payslips |
| **Digital banks** | Revolut, N26, Monzo operate without physical branches — entirely ML-driven backend |

::: {.callout-important}
## ⭐ Fraud Detection: The "Accuracy Trap" (Preview of Week 3)

*Professor's note: "If you have a machine learning algorithm that classifies every transaction as non-fraudulent, that algorithm is going to have 99.999% accuracy. Which shows that a machine learning metric of just judging something on accuracy is a worthless business metric — because it hasn't found any fraudulent transactions at all."*

This is a **critical exam concept**. Raw accuracy is often misleading. With fraud data:

| What the model does | Accuracy | Fraud caught |
|---|---|---|
| Predicts "not fraud" for every transaction | **99.999%** | **0** |
| A properly tuned ML model | ~99.9% | thousands |

The first model looks better on paper but is **completely useless**. This is the **imbalanced dataset problem** (Week 5). Better metrics: **precision**, **recall**, **F1 score** (Week 3).
:::

## Manufacturing, Telecoms, Pharmaceuticals

- **Manufacturing:** optimal production schedules, supply/demand forecasting, **anomaly detection** (spotting defective products on a production line), **predictive maintenance** (predicting machine failure *before* it happens, avoiding expensive downtime).
- **Telecoms:** speech recognition for customer service, churn prediction, ad-spend analysis across customer segments, spatial analysis for cell-tower placement.
- **Pharmaceuticals:** drug discovery, clinical trial analysis, radiology (image analysis), electronic health records, epidemic outbreak prediction.

---

# Machine Learning Workflows: CRISP-DM

## Why Do We Need a Workflow?

ML projects are **research problems**, not standard software projects. You don't know in advance what will work. You must explore, experiment, and iterate.

*Professor's note: "Essentially, it is a research problem, and needs to be treated as such."*

Two common workflows are mentioned: **CRISP-DM** and **SEMMA**. CRISP-DM is the industry standard and is the one you need to know.

## The CRISP-DM Framework

**CRISP-DM** = **CR**oss **I**ndustry **S**tandard **P**rocess for **D**ata **M**ining.

It is an **agile, iterative** process — you frequently move backwards and forwards between phases. This is normal and expected.

```
Business Understanding ←→ Data Understanding
           ↓                    ↑
    Data Preparation ←──────────┘
           ↓
        Modelling ←──┐
           ↓         │
       Evaluation ───┘  (may loop back to any earlier phase)
           ↓
       Deployment
```

| Phase | Key Question | What Happens |
|-------|-------------|-------------|
| **1. Business Understanding** | What does the business actually need? | Define the problem. What does success look like? What is the business cost of a wrong prediction? |
| **2. Data Understanding** | What data do we have? Is it clean? | Explore, visualise, and assess data quality (Week 2 covers this!) |
| **3. Data Preparation** | How do we organise the data for modelling? | Clean, transform, encode, engineer features. **This takes the most time in a real project — often 60–80% of total effort.** |
| **4. Modelling** | What techniques should we apply? | Train ML algorithms on the prepared data |
| **5. Evaluation** | Which model meets the business objective? | Not just accuracy — does it actually solve the problem? |
| **6. Deployment** | How do stakeholders access the results? | Put the model into production; monitor it afterwards |

::: {.callout-important}
## ⭐ Key Exam Point: CRISP-DM Is Not Linear

*Professor's note: "The sequence of the phases is not rigid. Moving back and forth between different phases is always required. The outcome of each phase determines which phase, or particular task of a phase, has to be performed next."*

You might: build a model → realise the data needs more cleaning → go back to Data Preparation → rebuild the model → evaluate again. This cycling is the **correct** way to work, not a mistake.
:::

::: {.callout-warning}
## ⚠️ Common Mistake: Rushing to Modelling

Junior data scientists often want to jump straight into building models because that feels like "real" data science. The professor has emphasised (and will emphasise again in Week 7) that **feature engineering** and **data understanding** matter far more than the choice of algorithm. A great model on bad data is worthless. A simple model on well-prepared data often wins.
:::

## Ethics and Fairness in Machine Learning

This is not optional — it is a professional and legal responsibility.

::: {.callout-warning}
## ⚠️ Common Mistake: Assuming ML is "Objective"

ML models learn from historical data. If historical data contains bias (which it almost always does), the model will **learn and reproduce** that bias. The algorithm itself is not the problem — the data is. But the data scientist is responsible for recognising and addressing this.

*Professor's note: "ML is only a complex pattern matcher. If your data is biased or unfair to a group, it is almost certainly caused by the data — it's just reflecting existing biases."*
:::

**Real-world examples of ML bias:**

| Case | What Happened | Lesson |
|------|---------------|--------|
| **Amazon's hiring tool** | Trained on historical CVs (mostly male), it penalised CVs that included the word "women's" (e.g. "women's chess club"). Amazon scrapped it. | Historical bias → model bias |
| **Facial recognition** | Shown to be far less accurate on darker skin tones, leading to wrongful arrests. | Unbalanced training data → poor generalisation |
| **Criminal justice AI** | Risk scores used in sentencing decisions showed racial disparities. | High-stakes decisions need extra scrutiny |

::: {.callout-important}
## ⭐ Professional Responsibility

*Professor's note: "You, as the data scientist, have the responsibility to stand up inside the business and say: if I build a model with this dataset, I think it might not work in all cases. And that's a hard thing to say. But the reputational damage companies can suffer from having machine learning models that perform poorly or discriminate — it's worth standing up and saying I don't think this is going to work, if you have doubts about something beforehand."*

The reputational and legal damage from a biased ML model can be severe. It is better to raise concerns early than to deploy a model that discriminates.
:::

---

# Types of Machine Learning

There are three main types. These are **not** overlapping — each has a completely different structure.

| Type | Data Available | Goal | Examples |
|------|----------------|------|---------|
| **Supervised Learning** | Data has both inputs (X) and correct labels (Y) | Learn to predict Y from X | Fraud detection, spam filter, image classification |
| **Unsupervised Learning** | Data has inputs (X) only — no labels | Find hidden structure or patterns | Customer segmentation, dimensionality reduction |
| **Reinforcement Learning** | An agent takes actions and receives rewards or penalties | Learn a strategy to maximise cumulative reward | Self-driving cars, game-playing AI (AlphaGo), robotics |

::: {.callout-note}
## 💡 Analogy: Three Ways to Learn

- **Supervised:** Like learning with a teacher who marks your homework. You have both the questions (X) and the correct answers (Y) — you learn the mapping.
- **Unsupervised:** Like exploring a new city without a map. Nobody tells you what you'll find — you discover natural groupings and landmarks yourself.
- **Reinforcement:** Like learning to ride a bike. You try, fall, adjust, try again. Feedback (reward/penalty) teaches you the right strategy over time.
:::

## Supervised Learning in Detail

This is the main focus of this course (Weeks 2–5, 8–10).

**How it works:**

1. You have historical data with known outcomes (the **label** or **target variable**, Y).
2. You also have **features** (predictor variables, X) — the information used to make predictions.
3. You train an algorithm on this data → it learns the relationship between X and Y.
4. You apply the trained model to new data to make predictions.

*Professor's note: "The process of creating a machine learning model is known as model building or model training. Once we have the model, we can use it on new data to make predictions."*

### Supervised Learning: Classification vs. Regression

| Sub-type | What Is Predicted | Output | Algorithms |
|----------|------------------|--------|-----------|
| **Classification** | A category (class label) | Discrete: "spam" / "not spam"; fraud / not fraud; class 0/1/2/3 | Logistic Regression, Decision Trees, Random Forest, Naive Bayes, SVM, Neural Networks |
| **Regression** | A continuous number | e.g. house price = €345,000; temperature = 18.4 °C | Linear Regression, Generalised Linear Models, Decision Trees, Random Forest, Deep Learning |

::: {.callout-important}
## ⭐ Classification vs. Regression — Know the Difference

**Classification** predicts a **category**. Think: *"Which box does this belong in?"*

**Regression** predicts a **number**. Think: *"How much / how many?"*

- **Binary classification** = only 2 possible classes (e.g. fraud / not fraud).
- **Multi-class classification** = 3 or more possible classes (e.g. handwritten digits 0–9).

Notice that **Decision Trees, Random Forest, and Deep Learning** appear in *both* lists — many algorithms can do either, depending on how you configure them.
:::

## Reinforcement Learning — The AlphaGo Story

While this course will not cover reinforcement learning in depth, the professor shared a memorable example.

::: {.callout-note}
## 🎮 AlphaGo and the Power of Self-Play

**Chess** engines beat the best humans in the late 1990s. But **Go** is vastly harder — the number of possible board positions is astronomical compared to chess.

Researchers programmed the rules of Go and set up two copies of the machine to **play against each other — millions, possibly billions of times**. Over time, the machines discovered strategies no human had ever invented. When AlphaGo eventually played human grandmasters, it won. Some of its moves were initially described as "bad" by commentators — and then, several moves later, revealed themselves as brilliant.

**Why this matters:** reinforcement learning is behind self-driving cars, robotics, and advanced game-playing AI. It is fundamentally different from supervised learning because the "teacher" is the **environment** — the agent receives rewards for good actions and penalties for bad ones, and has to discover the right strategy on its own.
:::

---

# Worked Example: Linear Regression in R

This section walks through the complete regression example from the lecture, step by step. This is also the full **Week 1 Lab**.

## The Setup: What Are We Doing?

We are going to:

1. Create a "true" mathematical relationship: **y = 1.4x − 1.2**
2. Add some random noise to simulate real-world data (real data is **never** perfectly clean).
3. Use R's `lm()` function to estimate the line from the noisy data.
4. Compare the estimated line to the true line.
5. Measure how well the model fits using **R-squared**.
6. Repeat with **more noise** and see R-squared drop.

::: {.callout-note}
## 💡 Why This Matters

In real life, we never know the "true" relationship — we only see noisy data. This example is a teaching tool: we create a known truth, then see how well our model can recover it. **This is the essence of all machine learning.**

*Professor's note: "The green line is reality, which is often hidden from us. The X points on the graph are the observed values. Our blue line is our line of best fit. What we are doing is basically trying to calculate the blue line from our observed values."*
:::

## Step 1: Generate the Data

```{r}
#| eval: false
set.seed(42)        # Makes randomness reproducible — same result every time
m <- 1.4            # True slope (gradient of the line)
b <- -1.2           # True intercept (where line crosses y-axis)
x <- 0:9            # x values: 0, 1, 2, ..., 9

y  <- m*x + b       # The TRUE y values (the perfect line)
py <- y + rnorm(10) # NOISY y values: add random noise from a normal distribution

# Inspect
x
## [1] 0 1 2 3 4 5 6 7 8 9

y
## [1] -1.2  0.2  1.6  3.0  4.4  5.8  7.2  8.6 10.0 11.4

py
## [1]  0.1709584 -0.3646982  1.9631284  3.6328626  4.8042683  5.6938755
## [7]  8.7115220  8.5053410 12.0184237 11.3372859
```

::: {.callout-note}
## 💡 What is `set.seed(42)`?

Random number generators in computers are not truly random — they use a mathematical formula starting from a **seed** number. `set.seed(42)` fixes the starting point, so you get the same "random" numbers every time.

*Professor's note: "Regardless of the operating system, regardless of the version of R, the next person who takes your code should get the same results as you — literally, figuratively the same results."*

**Why this matters:** reproducibility is essential in science and data analysis. Any colleague or marker should be able to run your code and get the same numbers.

The number 42 has no special meaning. Any number works. It is a convention borrowed from *The Hitchhiker's Guide to the Galaxy*.
:::

::: {.callout-note}
## 💡 What is `rnorm(10)`?

`rnorm(10)` generates 10 random numbers from a **normal distribution** (the classic bell-curve shape) with mean 0 and standard deviation 1. Adding this to `y` simulates the kind of random measurement error or noise that exists in all real data.
:::

::: {.callout-note}
## 💡 R Tip: Vectorised Operations

*Professor's note: "One of the nice features of R is that if you mix an array operation with a scalar value, it gives you the output that you expect — another array. In other languages, you'd have to write for-loops."*

`y <- m*x + b` multiplies **every** element of the vector `x` by `m`, then adds `b` — all in one line. No loops needed. This is called **vectorisation** and it makes R very expressive for data work.
:::

## Step 2: Plot the Data and Fit the Model

```{r}
#| eval: false
# Plot the true line in green
plot(x, y, type = "l", col = "green",
     main = paste("y = ", m, "x ", b, sep = ""))

# Add the noisy observed points as X marks
points(x, py, pch = 4)

# Fit a linear model: py as a function of x
model1 <- lm(py ~ x)

# Add the model's estimated line in blue (dashed)
lines(x,
      model1$coefficients[2] * x + model1$coefficients[1],
      lty = 2, col = "blue")
```

**Reading the plot:**

| Element | Colour | What It Represents |
|---------|--------|--------------------|
| Solid green line | Green | The **true** underlying relationship (y = 1.4x − 1.2) — hidden in real life |
| X marks | Black | The **observed** noisy data points |
| Dashed blue line | Blue | The model's **estimated** best-fit line |

## Step 3: Inspect the Model Output

```{r}
#| eval: false
model1 <- lm(py ~ x)
model1

## Call:
## lm(formula = py ~ x)
##
## Coefficients:
## (Intercept)    x
##     -0.7895   1.4304
```

The model estimated:

- **Intercept = −0.7895** (true value was **−1.2**)
- **Slope = 1.4304** (true value was **1.4**)

These are close but not exact — because the model is working from noisy data. **Models are always approximations of reality.** This is a core concept of the entire course.

::: {.callout-note}
## 💡 What does `lm(py ~ x)` mean?

- `lm` stands for **linear model** — R's built-in function for linear regression.
- `py ~ x` is a **formula**: it means *"model `py` as a function of `x`"*.
- The `~` symbol is read as *"is modelled by"* or *"depends on"*.
- R returns a **model object** containing coefficients, residuals, p-values, and more.
:::

## Step 4: The Full Summary

```{r}
#| eval: false
summary(model1)

## Call:
## lm(formula = py ~ x)
##
## Residuals:
##     Min      1Q  Median      3Q     Max
## -1.0056 -0.7056 -0.1180  0.7218  1.3647
##
## Coefficients:
##             Estimate Std. Error t value Pr(>|t|)
## (Intercept) -0.78949    0.51766  -1.525    0.166
## x            1.43040    0.09697  14.752 4.39e-07 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
##
## Residual standard error: 0.8807 on 8 degrees of freedom
## Multiple R-squared:  0.9645,  Adjusted R-squared:  0.9601
## F-statistic: 217.6 on 1 and 8 DF,  p-value: 4.386e-07
```

**Reading the coefficients table:**

| Column | Meaning |
|--------|---------|
| `Estimate` | The model's best guess for the coefficient |
| `Std. Error` | How uncertain we are about the estimate (smaller = more confident) |
| `t value` | A test statistic — larger absolute value = more significant |
| `Pr(>|t|)` | The **p-value**. Below 0.05 is typically considered statistically significant |
| Stars `***` | Significance codes: `***` = very significant; `**` = significant; `.` = borderline |

**Reading the bottom lines:**

| Output | Meaning |
|---|---|
| `Residual standard error` | Average size of the "miss" between predicted and actual values |
| `Multiple R-squared` | Proportion of variation explained by the model (0 to 1) |
| `Adjusted R-squared` | R-squared adjusted for the number of predictors (fairer for comparing models) |
| `F-statistic` + `p-value` | Tests whether the model as a whole is better than no model. Low p-value = yes, the model is useful |

::: {.callout-important}
## ⭐ Statistical Significance vs. Practical Significance

*Professor's note: "Even if a variable is statistically significant — has stars — you should also look at the magnitude of its coefficient. If the coefficient is tiny, multiplying it by the X values won't meaningfully affect predictions."*

**Always check both:**

1. Is the p-value significant? (**statistical significance**)
2. Is the coefficient large enough to matter in practice? (**practical / business significance**)

A variable can be statistically significant but practically irrelevant — especially in very large datasets, where tiny effects look significant just because you have lots of data.
:::

## Step 5: Calculating R-squared by Hand

```{r}
#| eval: false
r_squared1 <- cor(x, py) ^ 2
r_squared1
## [1] 0.9645404
```

This matches the `Multiple R-squared` from `summary(model1)`.

### What Is R-squared?

**R-squared** (also written R²) measures how well your model explains the variation in the data.

- Range: **0 to 1**
- **R² = 1** → perfect fit (model explains 100% of the variation)
- **R² = 0** → terrible fit (model explains nothing)
- **R² = 0.9645** → the model explains about **96.45%** of the variation in `py`

### What Is Correlation?

**Correlation** (often written `r`) measures the strength of the linear relationship between two variables.

- Range: **−1 to +1**
- **+1** = perfect positive relationship (both increase together)
- **−1** = perfect negative relationship (one increases, the other decreases)
- **0** = no linear relationship
- Rule of thumb: |r| > 0.7 is considered strong

**R-squared is literally the square of correlation** (for simple linear regression): **R² = r²**

::: {.callout-warning}
## ⚠️ Suspicious R-squared Values — Data Leakage Warning

*Professor's note (preview of Week 5): "If you ever see an R-squared above ~0.85–0.90 on real-world tabular data, be suspicious. It might indicate data leakage."*

Real-world data is messy. An extremely high R-squared on tabular data is unusual and might mean:

- You accidentally included the target variable (or something very similar) as a predictor — **data leakage** (Week 5).
- The problem is trivially simple and doesn't need ML.
- Something went wrong in your data pipeline.

**In this worked example** R² = 0.9645 is fine because we **made up** the data from a perfect line plus a little noise. Real data rarely behaves this cleanly.
:::

## Step 6: Effect of More Noise

The lecture then repeats the analysis with **more** random noise added:

```{r}
#| eval: false
# Note: set.seed NOT called again — we get a different random sample
py <- y + rnorm(10)   # New, noisier data

model2 <- lm(py ~ x)

r_squared2 <- cor(x, py) ^ 2
r_squared2
## [1] 0.8524506
```

**Comparison of the two models:**

| Model | Noise Level | R-squared | Interpretation |
|-------|-------------|-----------|----------------|
| Model 1 | Low noise | **0.9645** | Model explains 96.5% of variation |
| Model 2 | More noise | **0.8525** | Model explains 85.2% of variation |

**Key insight:** more noise → lower R-squared → less accurate model. The underlying true relationship is the same (y = 1.4x − 1.2), but noisier data makes it harder to recover. **Data quality directly determines model quality.**

## The General Linear Regression Formula

For one predictor:

$$\hat{y} = \beta_0 + \beta_1 x$$

For multiple predictors (which you'll use in practice):

$$\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_n x_n$$

Where:

- $\hat{y}$ = the predicted value (our model's best guess)
- $\beta_0$ = the **intercept** (the value of $\hat{y}$ when all X values are 0)
- $\beta_1, \beta_2, \ldots, \beta_n$ = the **coefficients** (learned from the data)
- $x_1, x_2, \ldots, x_n$ = the **feature values** (inputs to the model)

## How Does the Model Learn? — Gradient Descent (Intuition)

*Professor's note: "You start with a theoretical best fit, and what you try and do is move it slightly to the left, slightly to the right, make a judgment to see — was that an improvement? If it was, continue in that direction. That is effectively the algorithm for gradient descent, and that is the basis of a lot of machine learning algorithms, like SVM and deep learning."*

**Gradient Descent — step by step:**

1. Start with a random line (initial guess for coefficients).
2. Measure how wrong it is (the **error** or **loss**).
3. Nudge the coefficients in the direction that **reduces** error.
4. Repeat until the error stops improving (you've found the best fit).

This algorithm is behind **nearly all ML training**, from simple linear regression to massive deep learning models.

::: {.callout-note}
## 💡 Analogy: Finding the Bottom of a Valley in Fog

Imagine you're in a hilly landscape in thick fog. You want to reach the lowest point (minimum error). You can only see the ground immediately under your feet. So you take a small step in the direction that goes downhill. Repeat. Eventually you reach the bottom. That's gradient descent.

⚠️ The catch: if there are multiple valleys (local minima), you might end up at a "lower-than-starting-point" but **not the lowest** valley. This is an important subtlety in training deep networks.
:::

---

# Complete R Code — Full Lab

All the code from Week 1 in one runnable block. Try it yourself!

```{r}
#| eval: false

# ── Setup ──────────────────────────────────────────────────────────────────
set.seed(42)          # For reproducibility
m <- 1.4              # True slope
b <- -1.2             # True intercept
x <- 0:9              # x values

# ── Generate Data ──────────────────────────────────────────────────────────
y  <- m * x + b       # True y (the perfect line)
py <- y + rnorm(10)   # Observed y (with noise)

# ── Model 1: Low Noise ─────────────────────────────────────────────────────
model1 <- lm(py ~ x)
model1                # View coefficients
summary(model1)       # Full statistical summary

r_squared1 <- cor(x, py) ^ 2
r_squared1            # Should be ~0.9645

# ── Plot Model 1 ───────────────────────────────────────────────────────────
plot(x, y, type = "l", col = "green",
     main = paste("y = ", m, "x ", b, sep = ""))
points(x, py, pch = 4)
lines(x,
      model1$coefficients[2] * x + model1$coefficients[1],
      lty = 2, col = "blue")

# ── Model 2: More Noise ────────────────────────────────────────────────────
py <- y + rnorm(10)   # Re-generate with different noise (no set.seed)

model2 <- lm(py ~ x)
r_squared2 <- cor(x, py) ^ 2
r_squared2            # Should be ~0.8525 — lower due to more noise

# ── Plot Model 2 ───────────────────────────────────────────────────────────
plot(x, y, type = "l", col = "green",
     main = paste("y = ", m, "x ", b, sep = ""))
points(x, py, pch = 4)
lines(x,
      model2$coefficients[2] * x + model2$coefficients[1],
      lty = 2, col = "blue")
```

### Self-Check Exercises

Try these small modifications to test your understanding:

| # | Exercise | Expected Outcome |
|---|----------|------------------|
| 1 | Change `m <- 1.4` to `m <- 3.0`. | The line becomes steeper. Slope estimate should be close to 3. |
| 2 | Change `b <- -1.2` to `b <- 5`. | The line shifts up. Intercept estimate should be close to 5. |
| 3 | Replace `rnorm(10)` with `5 * rnorm(10)`. | R-squared drops significantly because noise is now much larger. |
| 4 | Replace `rnorm(10)` with `0.1 * rnorm(10)`. | R-squared becomes very close to 1 because noise is tiny. |
| 5 | Run the code without `set.seed(42)` twice. | You get different noisy values and slightly different coefficients each run. |

---

# Preview of Coming Weeks

The professor went through the full 12-week map at the start of the class. Knowing the roadmap makes it easier to see how each week fits in.

| Week | Topic | Technique Family |
|------|-------|-----------------|
| **1** | Intro to Data Mining & ML *(this week)* | — |
| **2** | Basic statistics & visualisations in R; data exploration (iris, Titanic) | EDA |
| **3** | Classification, decision trees, confusion matrix, ML vs business metrics | Supervised (classification) |
| **4** | Bias-variance tradeoff, one-hot encoding, logistic regression, Naive Bayes, SVM, Random Forest, Neural Networks | Supervised (classification) |
| **5** | LDA, QDA, data leakage, hyperparameter optimisation, imbalanced data | Supervised + tuning |
| **6** | Clustering: k-means, hierarchical; SVM; communicating results | Unsupervised |
| **7** | Generative A.I. | Special topic |
| **8** | Dimensionality reduction: t-SNE, PCA | Unsupervised |
| **9** | MNIST image data; combining dimensionality reduction with classification | Combined |
| **10** | High-dimensional data; outlier / anomaly detection | Unsupervised |
| **11** | Real-world case studies | Applied |
| **12** | Wrap-up | Review |

---

# Key Concepts Reference Table (Glossary)

| Concept | Definition | First Seen |
|---------|-----------|-----------|
| **Artificial Intelligence** | Computers doing human-like tasks | Week 1 |
| **Machine Learning** | Statistical algorithms that improve with more data | Week 1 |
| **Deep Learning** | ML algorithms great for images and text | Week 1 |
| **Data Mining** | Extracting patterns from large data (≈ ML) | Week 1 |
| **Supervised Learning** | Learning from labelled data (X and Y) | Week 1 |
| **Unsupervised Learning** | Finding patterns in unlabelled data (X only) | Week 1 |
| **Reinforcement Learning** | Learning by trial, error, and reward | Week 1 |
| **Classification** | Predicting a category (discrete output) | Week 1 |
| **Regression** | Predicting a number (continuous output) | Week 1 |
| **CRISP-DM** | 6-phase iterative data science process | Week 1 |
| **R-squared (R²)** | Goodness of fit: 0 = terrible, 1 = perfect | Week 1 |
| **Correlation** | Strength of linear relationship: −1 to +1 | Week 1 |
| **Gradient Descent** | Iterative algorithm to find best model parameters | Week 1 |
| **Deterministic** | Same input → same output (traditional software) | Week 1 |
| **Stochastic** | Output varies; system learns over time (ML) | Week 1 |
| **Imbalanced Dataset** | One class is very rare (e.g. fraud) — most algorithms struggle | Week 5 *(preview)* |
| **Data Leakage** | Target variable accidentally leaks into training data → artificially high R² | Week 5 *(preview)* |
| **Churn** | Customers who leave a subscription or service | Week 1 |
| **Predictor variables (X)** | The input features used to make predictions | Week 1 |
| **Target variable (Y)** | The outcome we want to predict | Week 1 |
| **Model training** | The process of fitting an ML algorithm to data | Week 1 |
| **Feature** | Another word for a predictor variable / input column | Week 1 |
| **Reproducibility** | Getting the same result every time using `set.seed()` | Week 1 |

---

# Exam Practice Questions

::: {.callout-important}
## ⭐ Practice Questions with Answers

**Q1. What is the difference between Machine Learning and Deep Learning?**

*Answer:* Machine Learning is a broad set of statistical algorithms that learn from data. Deep Learning is a specific family of ML algorithms that uses neural networks with many layers — it is particularly good at computer vision and text analysis. Deep Learning is a **subset** of Machine Learning.

---

**Q2. A model that predicts "not fraud" for every transaction achieves 99.999% accuracy on a fraud dataset. Is this a good model? Why or why not?**

*Answer:* No — it is a completely useless model. Because fraud is very rare (0.001% of transactions), predicting "not fraud" always gives high accuracy but catches zero fraud. This is the **imbalanced dataset problem**. Better metrics are **precision**, **recall**, and **F1 score** (Week 3).

---

**Q3. What are the 6 phases of CRISP-DM?**

*Answer:* 1. Business Understanding, 2. Data Understanding, 3. Data Preparation, 4. Modelling, 5. Evaluation, 6. Deployment. The process is **iterative** — you frequently move between phases.

---

**Q4. What does R-squared measure, and what value would a perfect model have?**

*Answer:* R-squared measures how well the model explains the variation in the data. It ranges from 0 to 1. A perfect model has R² = 1 (explains 100% of the variation). A value of 0.9645 means the model explains about 96.45% of the variation.

---

**Q5. What is the difference between supervised and unsupervised learning?**

*Answer:* In supervised learning, each data point has a label (Y) — the correct answer. The algorithm learns to predict Y from the input features X. In unsupervised learning, there are no labels — the algorithm finds hidden structure, groupings, or patterns in the data by itself.

---

**Q6. Give one example of how ML bias can occur, and explain who is responsible for addressing it.**

*Answer:* Amazon's hiring tool was trained on historical CVs that were mostly from male candidates. It learned to penalise CVs containing the word "women's" (e.g. "women's chess club"). The algorithm itself is just a pattern-matcher — the **bias lived in the training data**. The **data scientist** is responsible for recognising such risks and raising concerns with the business before deploying a biased model.

---

**Q7. Classify each of the following as classification or regression:**

| Task | Type |
|---|---|
| a) Predicting house price in € | Regression |
| b) Predicting whether an email is spam | Classification (binary) |
| c) Predicting the digit (0–9) in a handwritten image | Classification (multi-class) |
| d) Predicting tomorrow's temperature | Regression |
| e) Predicting whether a customer will churn | Classification (binary) |

---

**Q8. Why is `set.seed()` important when writing ML code?**

*Answer:* It fixes the starting point of R's random number generator, so every time the code runs it produces the same "random" numbers. This makes results **reproducible** — essential for science, teamwork, and debugging.

---

**Q9. In the linear regression formula $\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2$, what do $\beta_0$, $\beta_1$, and $\beta_2$ represent?**

*Answer:* $\beta_0$ is the **intercept** — the predicted value when all X's are zero. $\beta_1$ and $\beta_2$ are the **coefficients** for each predictor — learned from the data — representing how much $\hat{y}$ changes when each corresponding X increases by 1 unit.

---

**Q10. Why is feature engineering often more important than choice of algorithm?**

*Answer:* A model is only as good as the data you feed it. Well-prepared features (clean, transformed, properly encoded) often improve performance more than swapping a simple algorithm for a complex one. The professor emphasises this repeatedly throughout the course.
:::

---

# Summary Checklist

- ✅ AI ⊃ Machine Learning ⊃ Deep Learning (nested hierarchy)
- ✅ Data Mining ≈ Machine Learning (older, interchangeable term)
- ✅ ML is **stochastic** (learns from data); traditional software is **deterministic** (fixed rules)
- ✅ Three types of ML: Supervised (with labels), Unsupervised (no labels), Reinforcement (reward-based)
- ✅ Supervised learning divides into Classification (category) and Regression (number)
- ✅ **CRISP-DM** is a 6-phase, iterative, agile process for data science projects
- ✅ ML bias comes from biased **data** — always question the fairness of your model
- ✅ `lm(y ~ x)` fits a linear model in R
- ✅ `summary(model)` shows coefficients, p-values, and R-squared
- ✅ R-squared measures goodness of fit (0–1); check both statistical and practical significance
- ✅ More noise in data → lower R-squared → less accurate model
- ✅ Gradient descent is the core optimisation algorithm behind most ML models
- ✅ High accuracy on imbalanced data is misleading — look at precision, recall, F1 (Week 3)
- ✅ Suspicious R-squared above 0.9 on real tabular data? Check for **data leakage** (Week 5)
- ✅ `set.seed()` is essential for reproducibility
- ✅ Feature engineering usually matters more than algorithm choice

---

# Pre-Topic Checklist: Are You Ready for Week 2?

Before Week 2, you should be able to:

- [ ] Explain AI, ML, Deep Learning, and Data Mining in plain English — and give a real-world example of each
- [ ] Name the 3 types of Machine Learning and give an example use case for each
- [ ] Explain the difference between classification and regression, with 3 examples of each
- [ ] List the 6 phases of CRISP-DM in order and explain what happens in each
- [ ] Explain why CRISP-DM is described as "iterative" or "agile"
- [ ] Run the Week 1 R code on your own machine without errors
- [ ] Explain what `lm()`, `summary()`, `cor()`, and `set.seed()` do in R
- [ ] Interpret R-squared: what does a value of 0.85 mean?
- [ ] Explain why a 99.999% accurate fraud detection model might be completely useless
- [ ] Explain what gradient descent does in plain language
- [ ] Install R and RStudio on your machine (homework!)
- [ ] Explain why ML bias is usually a **data** problem, not an algorithm problem

---

# Additional Resources

| Resource | Link | Why Read It |
|----------|------|------------|
| Amazon AI hiring bias (Reuters) | <https://www.reuters.com/article/us-amazon-com-jobs-automation-insight-idUSKCN1MK08G> | Real case study on ML ethics — mentioned in lecture |
| Facial recognition bias (Nature) | <https://www.nature.com/articles/d41586-020-03186-4> | Fairness in ML — discussed in class |
| AI in criminal justice (MIT Tech Review) | <https://www.technologyreview.com/2019/01/21/137783/algorithms-criminal-justice-ai/> | Ethics and consequences of biased models |
| Russell & Norvig — *Artificial Intelligence: A Modern Approach* | Available in library | The classic AI textbook the professor recommended — "really interesting" |
| Kaggle | <https://www.kaggle.com> | Practice datasets and competitions |
| Free R / ML books | Canvas Discussions | Posted by lecturer each week — run the code! |
| RStudio | <https://posit.co/download/rstudio-desktop/> | The IDE the lecturer uses in every lab |

---

# Homework

- [ ] Install **R** from [https://cran.r-project.org](https://cran.r-project.org)
- [ ] Install **RStudio** from [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)
- [ ] Run the Week 1 R code (`week1_stat9013c.R`) on your machine
- [ ] Try all 5 self-check exercises in the Worked Example section
- [ ] Look at the free books listed in Canvas Discussions
- [ ] Watch the Week 2 pre-recorded Zoom class (uploaded before Friday midnight)

---

# My Notes — Annotation Space

*Use this space to add your own observations, questions, and notes from reviewing the lecture recording.*

---

**Questions I still have:**

<!-- Write your questions here -->

---

**Things I want to revisit:**

<!-- Write topics you found confusing here -->

---

**My own examples of the concepts:**

<!-- Write your own analogies or examples here — this is one of the best ways to check understanding -->

---

*Study guide based on STAT9013 Week 1 lecture slides, R code (`week1_stat9013c.R`), and class transcript — Mark Hodnett, Lecturer.*
