---
title: "Week 2: Basic Statistics, Visualisations in R, and Data Exploration"
subtitle: "STAT9013 — Data Mining and Machine Learning"
author: "Comprehensive Study Guide — v2"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: "📚 Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    highlight-style: github
    df-print: paged
    self-contained: true
    css: |
      body { font-size: 1.05em; line-height: 1.7; }
      h1, h2, h3 { margin-top: 1.8em; }
      .callout { margin: 1.2em 0; }
      code { font-size: 0.93em; }
execute:
  eval: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Run this block FIRST before anything else ────────────────────────────
# If you get "could not find function 'qplot'" or similar errors, it's almost
# certainly caused by packages not being loaded. Run this block first.
#
# First time only — install if you have never done so:
# install.packages(c("ggplot2","gridExtra","datasets","corrplot",
#                    "titanic","psych","openintro"))

library(ggplot2)
library(gridExtra)
library(datasets)
library(corrplot)
library(titanic)
library(psych)
library(openintro)
```

---

::: {.callout-important}
## ⭐ Assessment & Course Reminders

**GenAI and coursework policy:**

*Professor's note:* "It's fine to use [GenAI] for small little things… to basically ask GenAI to solve the solution, or to do the whole project — that's not a good idea. It will be very obvious, because your project will look probably similar to other people in the class."

- Using GenAI to write your **entire** assignment is **not acceptable**. Submissions that look cookie-cutter or suspiciously similar to classmates' work will raise red flags.
- Live coding sessions are now common in **hiring interviews**. You need to be able to do this yourself.
- GenAI for debugging a single line or clarifying a concept = fine. GenAI for writing your project = not fine.

**Next week:** Machine Learning Algorithms — Decision Trees, and Lab 3.

**Contact:** mark.hodnett@mtu.ie or through Canvas.
:::

---

## Quick Recap — What You Should Know from Week 1

Before diving in, make sure these are solid from last week:

| Topic | One-line Summary |
|---|---|
| Data → Information → Knowledge | Data is raw facts; processed data becomes information; applied information becomes knowledge |
| AI / ML / Deep Learning | AI is the broadest idea; ML and Deep Learning are subsets of it |
| CRISP-DM (6 stages) | Business Understanding → Data Understanding → Data Preparation → Modelling → Evaluation → Deployment |
| Supervised vs Unsupervised | Supervised = prediction with labelled data; Unsupervised = finding patterns without labels |
| Stochastic vs Deterministic | Traditional software: same inputs → same output. ML: same inputs may → different outputs |

*Questions about Week 1? Email mark.hodnett@mtu.ie or message via Canvas.*

---

## Key Topics Covered This Week

- CRISP-DM revisited — why 85–90% of a project happens *before* modelling
- The PayPal churn case study — real-world business understanding done right
- Summary statistics: mean, median, quartiles, IQR
- The `drawPlots` helper function in R
- Univariate plots: histograms, density plots, box plots
- **Interpreting** the plots (not just producing them)
- Skewness — what it is, why it matters for ML, and how to fix it with log transforms
- The mammals dataset — a concrete R² improvement example
- Working with built-in R datasets (`datasets` package)
- Data types in R — especially **factors**
- Full data exploration workflow on the Iris dataset (8 steps)
- Introduction to the Titanic dataset and binary classification
- Feature engineering overview — why it matters more than algorithm choice

---

## Why This Week Matters — A Preview of What's Coming {#forward-preview}

::: {.callout-note}
## 🔮 Forward Links — Where Week 2 Skills Are Used Later

Week 2 is the foundation of almost every later week. Your exploratory data analysis skills directly feed into:

| Later Week | What Week 2 Skill You'll Use |
|---|---|
| **Week 3 (Decision Trees)** | The setosa `if statement` rule (Petal.Length < 2) becomes the root split of a decision tree. The 177 missing Ages need imputation before any model can run. |
| **Week 5 (Titanic deep dive)** | The cabin encoding (77% missing) gets extracted as a deck letter. The Name column gets parsed for titles (Mr/Mrs/Miss/Master) to help impute Age better. |
| **Week 6 (Logistic Regression)** | Data leakage is introduced — and the key example is "don't calculate your mean-age-imputation on the full dataset, only on the training set." |
| **Week 7 (Clustering)** | The Iris setosa separation seen in Week 2 box plots is confirmed by k-means clustering — the model independently discovers the same pattern. |

**This is why data exploration matters.** The insights you find in Week 2 *become* the features, rules, and imputations of the later weeks. Skip data exploration and you skip the foundation of everything else.
:::

---

## CRISP-DM — Why Data Comes Before Modelling {#crisp-dm}

### The Framework

**CRISP-DM** stands for **C**ross-**I**ndustry **S**tandard **P**rocess for **D**ata **M**ining. It is a flexible, agile process for machine learning projects — not a rigid checklist. The arrows in the diagram go *back and forth* between stages because machine learning is a research process.

| Stage | What you do | Key point |
|---|---|---|
| **1. Business Understanding** | What does the business *actually* need? | Wrong question = wrong model. Focus on revenue, not just accuracy. |
| **2. Data Understanding** | What data exists? Is it in the right format? Is it even collected? | You may discover the data you need doesn't exist yet |
| **3. Data Preparation** | Clean, transform, engineer features | This is where most of your time goes |
| **4. Modelling** | Apply ML algorithms | Only ~10–15% of total project time! |
| **5. Evaluation** | Does the model meet the *business* goal? | A 99% accurate model can still be useless — see fraud example below |
| **6. Deployment** | How do stakeholders access results? | May include a human-in-the-loop review process |

::: {.callout-important}
## ⭐ Exam Tip — The Single Most Important CRISP-DM Quote

*Professor's note:* "You might only spend 10–15% of your time on a project actually in the modelling stage. The rest of it is business understanding, data preparation, and things like that."

**Two of the six stages are entirely about data — before any model is built.** Know *why* this matters, not just what the stages are. The examiner wants to see that you understand the emphasis on preparation over modelling.
:::

### Machine Learning as a Research Problem

Machine learning is fundamentally different from standard software development. The professor made this distinction clearly:

| Traditional Software | Machine Learning |
|---|---|
| Fixed requirements, known goal | Open-ended — you may not know if the problem is solvable |
| Same inputs → same outputs (deterministic) | Same inputs may not always → same outputs (stochastic) |
| Goal is fixed before you start | You discover whether the goal is achievable as you work |
| Jira tickets, two-week sprints | Iterative, experimental, research-like cycles |

*Professor's note:* "Machine learning is essentially a research problem. A lot of times, you're given a dataset and you're not told if you can actually solve the problem. The phrase we use a lot in the company that I work in — it's a **stochastic process**. It's not guaranteed to give you the same answers all the time."

::: {.callout-note}
## 💡 Analogy — Cooking competition vs. assembly line

Think of a cooking competition. Most of your time is spent selecting ingredients, tasting, adjusting, and going back to try again. That's CRISP-DM. An assembly line (traditional software) has fixed steps and a guaranteed output. Machine learning is the cooking competition.
:::

### The Fraud Detection Evaluation Problem

::: {.callout-important}
## ⭐ Exam Tip — Why 99% Accuracy Can Be Useless

The professor gave this example repeatedly. A model that predicts *every transaction is fine* would have 99.999% accuracy on a fraud dataset — because almost all transactions genuinely are fine. But it would catch zero fraudulent transactions.

**The solution:** Use the ML model as a *filter*, not a final decision-maker. Flag the top 1–2% of most suspicious transactions for a human to review. This is called **human-in-the-loop** machine learning.

**Forward link:** You will revisit this when you learn about confusion matrices, precision, recall, and the imbalanced dataset problem in Week 4+.
:::

---

## The PayPal Churn Case Study {#paypal}

::: {.callout-important}
## ⭐ Exam Tip — This Case Study Was Repeated Across Two Weeks

When the professor repeats an example across multiple sessions, it is a strong signal it is exam-relevant. Be prepared to describe the steps and the lesson.
:::

**The problem:** PayPal was losing approximately 1 million merchants per year for 17 years and could not fix it.

**The key insight:** The solution was not a fancy algorithm — it was carefully defining *what* they were actually trying to predict.

### The Filtering Steps They Took

| Step | What they removed | Why |
|---|---|---|
| Exclude cancellations | Accounts that already cancelled | By the time someone cancels, it's too late. Focus on *before* cancellation |
| Exclude "one-and-dones" | Merchants who transacted once with tiny revenue | Not worth effort; negligible revenue impact |
| Separate activation from retention | New merchants with onboarding problems | Different problem, different team required |
| Remove false positives | Seasonal businesses, subscription-only, TOS violations | They're not churning — they're seasonal or were removed for cause |
| Focus on revenue churn | Top 10% of merchants = 90% of revenue | Losing a €100k/month merchant matters far more than losing a €500/month one |

*Professor's note:* "The task is not really to prevent churn. The task is to make money. And to make money, you look at the customers that are spending the most, and you prioritize them."

### What They Actually Did

After filtering, they had ~50,000 high-value merchants who churned. They then:

1. Examined each case individually — looking at all touchpoints (logging, risk, compliance, customer service)
2. Identified **20 "killer scenarios"** — 20 distinct reasons why merchants leave
3. Applied a predictive model to bucket merchants into the relevant scenario
4. Sent results to customer service, who **called the merchants** to fix the problems

*Professor's note:* "Their fancy predictive model was basically a series of SQL queries, and then basically sending the results to customer service, who would, in a lot of cases, call the customers and fix the problems."

::: {.callout-note}
## 💡 The Core Lesson

Good business understanding — knowing *exactly* what you are predicting, for whom, and when — is more valuable than any algorithm. "Always ask: what am I actually trying to solve?" before reaching for a machine learning model.
:::

::: {.callout-warning}
## ⚠️ Common Mistake — Predicting the Wrong Thing (Data Leakage Preview)

The professor said: "Cancelling is too late. By the time someone goes along and cancels their Netflix, it's probably too late for Netflix to do anything about it."

**Data leakage warning:** Never use a variable that only exists *after* the event you are predicting (e.g., "date account was closed" to predict churn). This is called **data leakage** — the model will look great in training but fail completely in production.

**Forward link to Week 6:** Data leakage is formally covered in Week 6, where you'll see another form — accidentally computing transformations (like mean imputation) on the full dataset before the train/test split. Remember: any transformation you apply to predict future data must be based only on *past* (training) data.
:::

---

## Summary Statistics — Mean, Median, Quartiles, IQR {#summary-stats}

### The Five-Number Summary

For any column in a dataset, R's `summary()` function gives you the five key statistics:

```{r summary-example}
summary(iris[, -5])   # summary of all numeric columns (excluding Species)
```

| Statistic | What it means | R function |
|---|---|---|
| **Min** | Smallest value in the column | `min(x)` |
| **1st Quartile (Q1)** | 25% of values are below this | `quantile(x, 0.25)` |
| **Median (Q2)** | Middle value — 50% below, 50% above | `median(x)` |
| **Mean** | Sum of all values ÷ count | `mean(x)` |
| **3rd Quartile (Q3)** | 75% of values are below this | `quantile(x, 0.75)` |
| **Max** | Largest value | `max(x)` |

### Mean vs Median — When Each Is Better

| Measure | Definition | Sensitive to outliers? | Best used for |
|---|---|---|---|
| **Mean** | Sum ÷ count | **Yes** — one extreme value can shift it a lot | Symmetric data with no extreme outliers |
| **Median** | Middle value of sorted data | **No** — extreme values barely affect it | Skewed data (income, house prices, COVID deaths) |

::: {.callout-note}
## 💡 Analogy — Bill Gates walks into a room

11 people earn roughly €50,000 each. Mean and median are both ~€50,000 — they agree.

Now Bill Gates and Jeff Bezos walk in. Their wealth is billions.

- The **mean** salary in the room might jump to €30 million — nobody in the room actually earns that.
- The **median** shifts only slightly — it still represents a "typical" person in the room.

*Professor's note:* "This is why the media report the **median** age of COVID deaths rather than the average — because the average really doesn't mean anything. If the median is 60, that basically means 50 people were younger than 60, and 50 were older than 60. It's more descriptive of what's going on."
:::

::: {.callout-warning}
## ⚠️ Common Mistake — Defaulting to the Mean

Students often assume the mean is always the best measure of "average." This is wrong when data is **skewed** or has **outliers**.

Rule of thumb:
- **Symmetric data** (heights, test scores) → mean ≈ median → either is fine
- **Skewed data or outliers** (income, house prices, body weights) → use the **median**
:::

### Quartiles and Interquartile Range (IQR)

**Quartiles** divide sorted data into four equal parts.

| Name | Position | What it means |
|---|---|---|
| **Q1** (25th percentile) | 25% of data is below this | Lower boundary of the middle half |
| **Q2** (50th percentile) | This *is* the **median** | The exact middle |
| **Q3** (75th percentile) | 75% of data is below this | Upper boundary of the middle half |
| **IQR = Q3 − Q1** | — | Spread of the middle 50% — robust to outliers |

**How to calculate:**

1. Order the data from smallest to largest
2. Cut into 4 equal-size groups
3. Each group = one "quartile"
4. IQR = the range covered by the middle 2 quartiles (Q1 to Q3)

The IQR is used by box plots to define the "normal" range and identify outliers.

---

## Visualising Distributions — Univariate Plots {#univariate}

**Univariate** means "one variable at a time." These three plot types all show the *distribution* (spread and shape) of a single column, but in different formats.

| Plot type | Format | Best for | Audience |
|---|---|---|---|
| **Histogram** | Bars — count of values per bin | First look at shape | Technical and non-technical |
| **Density plot** | Smooth curve (like a smoothed histogram) | Comparing shapes; seeing skewness | Technical |
| **Box plot** | Compact: shows median, IQR, outliers | Side-by-side group comparison | Technical (avoid for business users) |

*Professor's note:* "The disadvantage of the box plot is that it's typically not well understood by non-statistical people. If I'm presenting to business users, I tend to avoid box plots. It's very useful for me and other people who are used to statistics, but not something a lot of people outside of statistics would be familiar with."

::: {.callout-important}
## ⭐ The Most Important Meta-Lesson of Week 2

*Professor's note (from Week 3 recap):* "If you're just producing the plots, and then jumping straight into machine learning, you're wasting your time. The plots don't tell you anything. You should be looking at the plots, **interpreting** the plots, and basically asking yourself questions from those plots to see if it can guide you into your feature engineering."

**Producing plots is easy. Interpreting plots is the skill.** Every plot you make should answer: "What does this tell me about the data? What questions does it raise? What decisions does it lead to?"
:::

### The `qplot` Function

`qplot` is part of the `ggplot2` package — described by the professor as "probably the best graphical code library that exists for any programming language."

| You pass | Default result | Override with |
|---|---|---|
| 1 column | Histogram | `geom = "density"`, `geom = "boxplot"` |
| 2 columns | Scatter plot | `geom = "point"`, `geom = "line"` |

Key parameters:

```{r qplot-template}
qplot(
  data,                        # the column / vector to plot
  geom = "histogram",          # "histogram", "density", "boxplot", "point"
  bins = 30,                   # histogram only: how many bars?
  main = "My Plot Title",      # title
  xlab = "X-axis label",       # x-axis
  ylab = "Y-axis label"        # y-axis
)
```

### Reading a Box Plot

```
     |-------[========|========]-------|    •  •
     ^        ^        ^        ^      ^   outliers
  whisker    Q1     median     Q3   whisker
             (25%)            (75%)
             |<------ IQR ------->|
             (box covers middle 50%)

Whiskers extend to 1.5 × IQR from Q1 and Q3.
Anything beyond the whisker = outlier (shown as a dot).
Blue dot (in drawPlots) = the MEAN.
```

| Part | Represents |
|---|---|
| Left edge of box | Q1 — 25th percentile |
| Line inside box | Median (Q2) |
| Right edge of box | Q3 — 75th percentile |
| Whiskers | Up to 1.5 × IQR from Q1 and Q3 |
| Dots beyond whiskers | Outliers |
| Blue dot (professor's helper) | The mean |

### The `drawPlots` Helper Function

The professor wrote this helper function — it draws all three plots side by side. **Blue line = mean. Green line = median.** When they are in the same place, the data is symmetric. When they diverge, the data is skewed.

```{r drawPlots}
drawPlots <- function(data, title) {
  p01 <- qplot(data, geom = "histogram", bins = 100) +
    geom_vline(xintercept = mean(data), size = 2, color = "blue") +
    geom_vline(xintercept = median(data), size = 2, color = "green")

  p02 <- qplot(data, geom = "density") +
    geom_vline(xintercept = mean(data), size = 2, color = "blue") +
    geom_vline(xintercept = median(data), size = 2, color = "green")

  p03 <- qplot(data, geom = "boxplot") +
    coord_flip() +
    geom_point(aes(x = mean(data), y = 0), size = 3, color = "blue")

  gridExtra::grid.arrange(p01, p02, p03, nrow = 1, top = title)
}
```

---

## Skewness — What It Is and Why It Matters {#skewness}

### What Is Skewness?

**Skewness** describes how *asymmetric* a distribution is. A normal (bell-shaped) distribution has no skewness — the left side mirrors the right side.

| Distribution | Shape | Tail direction | Mean vs Median | Real-life example |
|---|---|---|---|---|
| **Normal** | Symmetric bell curve | No tail | Mean ≈ Median | Heights of adult males in one country |
| **Right (positive) skew** | Most values low; few very high | Tail goes right → | Mean **>** Median | Income, house prices, animal body weights |
| **Left (negative) skew** | Most values high; few very low | Tail goes left ← | Mean **<** Median | Test scores after an easy exam |

::: {.callout-important}
## ⭐ The Key Rule — Mean Chases the Tail

**Right skew:** Mean > Median (a few very high values pull the mean up)
**Left skew:** Mean < Median (a few very low values pull the mean down)
**Normal:** Mean ≈ Median

This rule lets you identify skew direction from a single statistic — without even plotting. Know it for the exam.
:::

```{r skewness-demo}
# Normal — mean ≈ median
data <- rnorm(10000, mean = 0, sd = 1)
drawPlots(data, "Normal distribution (median ~= mean)")

# Left (negative) skew — median > mean
data <- rbeta(10000, 5, 1)
drawPlots(data, "Left skewed (median-green > mean-blue)")

# Right (positive) skew — median < mean
data <- rbeta(10000, 1, 5)
drawPlots(data, "Right skewed (median-green < mean-blue)")
```

### Real-World Examples of Skewed Data

| Variable | Skew type | Why |
|---|---|---|
| Personal income | Right | Most people earn modestly; a few earn billions |
| Average spend per customer | Right | Most customers spend little; a few spend a lot |
| Distance travelled to work | Right | Most commutes are short; some are very long |
| Animal body weights | Right | Most mammals are small; elephants/whales are extreme outliers |
| Test scores on an easy exam | Left | Most students score high; a few score very low |
| Age at retirement | Left | Most retire around 65; very few retire very young |

### Why Skewness Matters for Machine Learning

*Professor's note (from Week 3 recap):* "Most machine learning algorithms that handle numeric values may — it might be an explicit or a hidden assumption — that the data is from a normal distribution. Because that's typically what datasets form for any realistic measurements. If you measure something, typically most of the time it's a normal distribution, so they have a problem with skewed datasets."

When data is strongly skewed:

- The model's performance can degrade significantly
- The relationship between variables is harder to capture linearly
- Outliers exert disproportionate influence on the model

**The fix:** Apply a mathematical transformation to make the data more normal. The most common are:

| Transformation | When to use | Notes |
|---|---|---|
| **log()** | Right-skewed data with all positive values | Simple and very common — used in class |
| **sqrt()** | Mild right skew | Less aggressive than log |
| **Box-Cox** | General-purpose — finds the best transformation automatically | More accurate than log; a real-world standard |

*Professor's note:* "There are other [transformations]. I think **Box-Cox** is probably the most common and more accurate than log. We just use log as a very simple example for our mammals dataset."

::: {.callout-note}
## 💡 Analogy — Compressing a stretched photo

Imagine a photo that is stretched very wide on the right side (a few extreme values). The log transform is like compressing it back to a more normal shape. You're not removing any data — you're just rescaling it so the extreme values don't dominate.
:::

::: {.callout-warning}
## ⚠️ When NOT to Just Transform

*Professor's note:* "Does this mean you should naturally go with all skew data, perform a log function on it? Possibly not. Maybe what you should do is try both — with and without — and see if it has an effect on the data. In this case, we only have one variable [brain weight], so of course it's highly predictive. But it doesn't work in all cases."

Also consider: **are your outliers real or errors?** If they're data entry errors, remove them. If they're real extreme cases (like elephants), keep them and transform.
:::

---

## The Mammals Dataset — A Concrete Log Transform Example {#mammals}

This is one of the most important examples in Week 2. It shows *numerically* how a simple data transformation can dramatically improve a model.

### Step 1 — Plot the Raw Data

```{r mammals-raw}
p01 <- qplot(mammals$body_wt, geom = "density",
             main = "Mammals body weight", xlab = "body weight")
p02 <- qplot(mammals$brain_wt, geom = "density",
             main = "Mammals brain weight", xlab = "brain weight")
gridExtra::grid.arrange(p01, p02, nrow = 1)
```

Both plots are **extremely right-skewed** — a few large mammals (elephants, hippos) pull everything to the right.

*Professor's note:* "If we try and plot [body weight vs brain weight], it's almost impossible to draw a line that encompasses all this data, because the two elephants are so big that linear regression is going to be pulled… it's going to underestimate the body weight."

### Step 2 — Build Model 1 on Raw Data

```{r mammals-model1}
model1 <- lm(body_wt ~ brain_wt, data = mammals)
r_sq1  <- cor(model1$fitted.values, mammals$body_wt)^2
r_sq1   # → 0.87
```

R² = **0.87** — not bad, but the skewed data is limiting the model.

### Step 3 — Apply the Log Transform

```{r mammals-log-transform}
mammals$log_body_wt  <- log(mammals$body_wt)
mammals$log_brain_wt <- log(mammals$brain_wt)

# Compare raw vs transformed distribution:
p01 <- qplot(mammals$body_wt,     geom = "density",
             main = "Mammals body weight",       xlab = "body weight")
p02 <- qplot(mammals$log_body_wt, geom = "density",
             main = "Mammals body weight (log)", xlab = "body weight (log)")
gridExtra::grid.arrange(p01, p02, nrow = 1)
```

The log-transformed distribution is now approximately **normal**. The elephants are still in the dataset — they are just no longer extreme outliers relative to the others.

### Step 4 — Why Not Just Delete the Elephants?

*Professor's note:* "I don't want to just throw out my larger records that I have data for. An **outlier should be a record that was recorded incorrectly, or information not likely to occur in future**. On the other hand, if my task is to predict body weight from brain weight, there certainly are larger mammals like elephants, hippos, giraffes — so I shouldn't throw out the larger records. Also, if we delete the elephants, the cows and horses would become the new outliers. And if I ever encountered a mammoth, my model wouldn't work for that either."

This is an important principle for handling outliers:

| Outlier Type | What to do |
|---|---|
| **Data-entry error** (e.g., age = 200) | **Remove** — it doesn't represent reality |
| **Real but extreme value** (e.g., an elephant in a mammals dataset) | **Keep and transform** — it is a legitimate future case |
| **A unique sub-population** (e.g., the two elephants) | Investigate — may indicate a bimodal distribution |

### Step 5 — Plot the Relationship (Raw vs Log)

```{r mammals-scatter}
p01 <- qplot(body_wt,     brain_wt,     data = mammals,
             main = "Body weight v brain weight")
p02 <- qplot(log_body_wt, log_brain_wt, data = mammals,
             main = "Log of Body weight v log of brain weight")
gridExtra::grid.arrange(p01, p02, nrow = 1)
```

The raw plot is a messy cloud — nearly impossible to draw a line through. The log-transformed plot shows a **clear, near-linear relationship**.

### Step 6 — Build Model 2 on Log Data

```{r mammals-model2}
model2 <- lm(log_body_wt ~ log_brain_wt, data = mammals)
r_sq2  <- cor(model2$fitted.values, mammals$log_body_wt)^2
r_sq2   # → 0.92
```

::: {.callout-important}
## ⭐ The Numbers to Memorise

| | Model 1 (raw) | Model 2 (log-transformed) |
|---|---|---|
| **R²** | 0.87 | 0.92 |
| **Error rate** | 1 − 0.87 = **0.13** | 1 − 0.92 = **0.08** |
| **Error reduction** | — | (0.13 − 0.08) / 0.13 = **37.79% fewer errors** |

The algorithm did not change. The improvement came **purely from transforming the data.**

*Professor's note:* "Don't just focus on the absolute value of the increase in accuracy. Focus on the relative difference in error rate. I decreased my error rate from 0.13 down to 0.08 — that's a 30–40% improvement on the existing model."
:::

---

## R Data Types and the `datasets` Package {#datatypes}

### R's Basic Data Types

Understanding and setting the correct data types is a critical and often underestimated task in data preparation.

| Type | Description | Example |
|---|---|---|
| `numeric` | Numbers with decimal points | 5.1, 3.14, -0.5 |
| `integer` | Whole numbers | 1L, 42L |
| `character` | Text strings | "male", "setosa" |
| `logical` | True or False | TRUE, FALSE |
| `factor` | Categorical variable — a fixed set of possible values | Species = {setosa, versicolor, virginica} |

::: {.callout-important}
## ⭐ Factors — A Special R Data Type Worth Knowing

*Professor's note:* "Factor is a special variable type, really only seen in R. It basically says there are three levels — the species must be one of these three. It's almost like an **enum** — constrained to be one of a value in a set."

**Why it matters for ML:** Some machine learning algorithms in R need categorical variables stored as `factor`, not as strings or integers. For example, if `Pclass` (1, 2, 3) is stored as `integer`, an algorithm may treat it as a continuous number — as if class 3 is "three times worse" than class 1. Converting it to `factor` tells the algorithm: these are categories, not numbers.

*Professor's note:* "Crucial to get the column types correct!"

**Forward link:** In Week 3's decision tree, the `Survived` column must be a factor — otherwise `rpart` will try to do regression instead of classification.
:::

```{r type-conversion}
# Converting types — the function usually starts with "as."
as.numeric("3")      # → 3       (works fine)
as.numeric("three")  # → NA      (fails — R can't convert text to number)

# Converting a column to factor:
titanic_train$Survived <- as.factor(titanic_train$Survived)
titanic_train$Pclass   <- as.factor(titanic_train$Pclass)
```

::: {.callout-warning}
## ⚠️ Watch Out for Encoded Missing Values

*Professor's note:* "A lot of systems might use minus 1 as the default value for empty. If you're calculating revenue and the underlying system didn't want to use an empty value, it might use minus 1 to say 'no data'. If you apply that data blindly into a machine learning algorithm, the algorithm is not going to know the difference between 999 [meaning 'missing'] and any other value."

**Common "fake" missing value encodings to look for:**

- `-1` (especially in older systems)
- `999` or `9999`
- `0` (when 0 is not a valid value)
- Empty strings `""`
- `NA` (R's official missing value marker)
:::

### The `datasets` Package

R comes with a built-in collection of datasets for practice. Some important ones:

| Dataset | Description | Good for practising |
|---|---|---|
| `iris` | 150 flower measurements across 3 species | Classification, correlation, EDA |
| `titanic_train` | Titanic passenger survival data | Binary classification, missing data |
| `mammals` | Brain and body weights of 62 mammals | Regression, log transforms |
| `airquality` | New York air quality measurements | Missing data, time series |
| `USArrests` | Violent crime rates by US state | Clustering, PCA |
| `AirPassengers` | Monthly airline passengers 1949–1960 | Time series |

```{r datasets}
library(help = "datasets")   # see all available datasets
data(iris)                   # load a dataset by name
```

---

## Data Exploration — The 8-Step Workflow on the Iris Dataset {#iris-eda}

### Why Data Exploration Is Critical

*Professor's note:* "Data exploration is critical to the success of your project. It is more important than anything you do in the machine learning model build stage, including picking an algorithm. The work done at this stage will mean a more robust and accurate machine learning model."

### About the Iris Dataset

The Iris dataset was created by British biologist and statistician **Ronald Fisher** in 1936. It contains 150 measurements from three species of iris flower:

- **Iris setosa**
- **Iris versicolor**
- **Iris virginica**

Four features were measured from each of the 50 samples per species: sepal length, sepal width, petal length, and petal width.

*Professor's note:* "The Iris dataset is the 'Hello World' of the machine learning process. It's a very old dataset, but that's why it's always given as the example — because the patterns within it are well-understood and can be verified. In 1936, there wasn't even the idea of colour photography, so the only way they had was measurements."

### The 8 Steps at a Glance

| Step | Function | Purpose |
|---|---|---|
| 1 | `dim()`, `names()`, `str()` | Understand structure |
| 2 | `summary()` | Get summary stats |
| 3 | `qplot(..., geom="histogram")` | See distributions |
| 4 | `qplot(..., geom="density")` | Smooth view of distributions |
| 5 | `qplot(..., geom="boxplot")` with grouping | Compare groups |
| 6 | `qplot(x, y, colour=group)` | See relationships between variables |
| 7 | `cor()`, `corrplot.mixed()` | Measure linear relationships |
| 8 | `pairs.panels()` | Everything at once |

### Step 1 — Load and Inspect Structure

```{r iris-step1}
data(iris)

dim(iris)    # 150 rows, 5 columns
names(iris)  # "Sepal.Length" "Sepal.Width" "Petal.Length" "Petal.Width" "Species"
str(iris)    # column types — numeric for measurements, factor for Species
```

::: {.callout-important}
## ⭐ The `str()` Function — Your First EDA Step

Always call `str()` on a new dataset first. It tells you:

- How many rows and columns
- The data type of every column
- A preview of the first few values

This one line often reveals problems immediately — wrong types, unexpected values, or suspicious patterns.
:::

### Step 2 — Summary Statistics

```{r iris-step2}
summary(iris[, -5])   # all columns except Species (column 5)
```

Iris summary statistics:

| Stat | Sepal.Length | Sepal.Width | Petal.Length | Petal.Width |
|---|---|---|---|---|
| Min | 4.300 | 2.000 | 1.000 | 0.100 |
| Q1 | 5.100 | 2.800 | 1.600 | 0.300 |
| Median | 5.800 | 3.000 | 4.350 | 1.300 |
| Mean | 5.843 | 3.057 | 3.758 | 1.199 |
| Q3 | 6.400 | 3.300 | 5.100 | 1.800 |
| Max | 7.900 | 4.400 | 6.900 | 2.500 |

**Key observation:** For `Petal.Length`, the median (4.35) is much higher than Q1 (1.60). This large gap hints at a bimodal distribution — confirmed in the next step.

### Step 3 — Histograms

```{r iris-step3}
p01 <- qplot(iris$Sepal.Length, geom = "histogram", bins = 10,
             main = "Histogram on Sepal.Length", xlab = "Sepal.Length", ylab = "")
p02 <- qplot(iris$Sepal.Width,  geom = "histogram", bins = 10,
             main = "Histogram on Sepal.Width",  xlab = "Sepal.Width",  ylab = "")
p03 <- qplot(iris$Petal.Length, geom = "histogram", bins = 10,
             main = "Histogram on Petal.Length", xlab = "Petal.Length", ylab = "")
p04 <- qplot(iris$Petal.Width,  geom = "histogram", bins = 10,
             main = "Histogram on Petal.Width",  xlab = "Petal.Width",  ylab = "")
gridExtra::grid.arrange(p01, p02, p03, p04)
```

**What to look for — Iris histograms:**

| Variable | What you see | What it means |
|---|---|---|
| `Sepal.Length` | Roughly bell-shaped | Normal-ish spread across species |
| `Sepal.Width` | Roughly normal | Similar across species |
| `Petal.Length` | **Bimodal (two humps!)** | Two distinct groups exist in the data |
| `Petal.Width` | **Bimodal (two humps!)** | Same reason — setosa is distinct |

::: {.callout-important}
## ⭐ Bimodal = Two Groups — A Red Flag in EDA

*Professor's note:* "This is a **red flag**. I'm seeing two different distributions within my dataset. There's something going on here. Already my exploratory data analysis is giving me tingly feelings that something different is within this dataset."

*Professor's note (Week 3 recap):* "A bimodal distribution in a single column is a really… it's a red flag. It's either saying data was not recorded correctly, or you have two different types of data, or two different data distributions within a single dataset."

**What bimodal means:** If your data has two peaks (two humps), it strongly suggests there are **two distinct sub-populations** mixed together. For the iris dataset, that sub-population is the setosa species, which has much smaller petals than the other two species.

**Analogy:** If you measured the height of everyone in Ireland — men, women, and children all mixed together — you'd see a bimodal or trimodal distribution. That's not one group; it's three overlapping groups.

**Why it matters for ML:** It tells you the data is naturally separable. That's extremely useful for a classification problem.
:::

### Step 4 — Density Plots

```{r iris-step4}
p01 <- qplot(iris$Sepal.Length, geom = "density",
             main = "Density plot on Sepal.Length", xlab = "Sepal.Length", ylab = "")
p02 <- qplot(iris$Sepal.Width,  geom = "density",
             main = "Density plot on Sepal.Width",  xlab = "Sepal.Width",  ylab = "")
p03 <- qplot(iris$Petal.Length, geom = "density",
             main = "Density plot on Petal.Length", xlab = "Petal.Length", ylab = "")
p04 <- qplot(iris$Petal.Width,  geom = "density",
             main = "Density plot on Petal.Width",  xlab = "Petal.Width",  ylab = "")
gridExtra::grid.arrange(p01, p02, p03, p04)
```

Density plots confirm the same findings as histograms. The smooth curves make the bimodal shape of `Petal.Length` and `Petal.Width` even clearer — you can see the dip between the two peaks more easily.

### Step 5 — Box Plots (by Species)

Splitting by species is powerful — it reveals how the groups differ.

```{r iris-step5}
p01 <- qplot(Species, Sepal.Length, data = iris, geom = "boxplot",
             main = "Boxplot on Sepal.Length")
p02 <- qplot(Species, Sepal.Width,  data = iris, geom = "boxplot",
             main = "Boxplot on Sepal.Width")
p03 <- qplot(Species, Petal.Length, data = iris, geom = "boxplot",
             main = "Boxplot on Petal.Length")
p04 <- qplot(Species, Petal.Width,  data = iris, geom = "boxplot",
             main = "Boxplot on Petal.Width")
gridExtra::grid.arrange(p01, p02, p03, p04)
```

**Key findings from box plots by species:**

| Variable | Observation | Implication |
|---|---|---|
| `Petal.Length` | Setosa box has **zero overlap** with versicolor and virginica | Petal length alone can perfectly separate setosa |
| `Petal.Width` | Same pattern — setosa is clearly distinct | Very useful feature for classification |
| `Sepal.Length` | Some overlap between species, but still broadly separated | Useful but not perfect |
| `Sepal.Width` | Surprisingly, setosa has *wider* sepals — but all species overlap | Less useful for separating species |

::: {.callout-important}
## ⭐ What "No Overlap" Means for Classification

*Professor's note:* "If you can draw a horizontal line that does not cross either box plot, then you can basically make a rule that perfectly separates those two classes."

When box plots for two groups do not overlap at all, a **simple rule** can separate them nearly perfectly:

```{r setosa-rule}
# Test the rule: if Petal.Length < 2, it's setosa
table(iris$Species)                         # 50 of each

iris_temp <- iris[iris$Petal.Length < 2, ]
table(iris_temp$Species)                    # ALL 50 are setosa!
```

This is a 100% accurate rule — discovered purely through exploratory data analysis. No algorithm needed.

**Forward link to Week 3:** This exact `Petal.Length < 2` rule becomes the first split in your decision tree. A decision tree is, in essence, a sequence of these "is there no overlap?" rules chained together automatically.

**Forward link to Week 7:** When you run k-means clustering on Iris, the algorithm *independently discovers* this same setosa group — confirming that the EDA insight was real.
:::

### Step 6 — Scatter Plots (Multivariate Exploration)

Scatter plots show the relationship between **two variables simultaneously**. Colouring by species shows how well the groups separate:

```{r iris-step6}
p01 <- qplot(Sepal.Width,  Sepal.Length, data = iris, colour = Species,
             main = "Sepal.Width v Sepal.Length")
p02 <- qplot(Petal.Length, Petal.Width,  data = iris, colour = Species,
             main = "Petal.Length v Petal.Width")
p03 <- qplot(Petal.Length, Sepal.Length, data = iris, colour = Species,
             main = "Petal.Length v Sepal.Length")
p04 <- qplot(Sepal.Width,  Petal.Width,  data = iris, colour = Species,
             main = "Sepal.Width v Petal.Width")
gridExtra::grid.arrange(p01, p02, p03, p04)
```

**Key insight:** The `Petal.Length vs Petal.Width` plot shows the **clearest separation** between all three species. This tells you these two features will be the most powerful predictors for a classification model.

::: {.callout-note}
## 💡 Visual Pattern Recognition in Scatter Plots

*Professor's note (Week 7):* "When you can do a vertical or horizontal line in a scatter plot, that means you can clearly differentiate them on a single variable using an if statement. If, when you do a scatter plot, you can draw a sloped line that differentiates the records completely, that indicates clearly that you can use linear regression to differentiate the two categories."

Summary of scatter-plot separability patterns:

| What you can draw | What ML model will work | Example |
|---|---|---|
| Horizontal or vertical line | Simple `if` statement / decision tree | Setosa separation on Petal.Length |
| Sloped straight line | Linear regression / logistic regression | Versicolor vs Virginica (partially) |
| Curved boundary | Tree-based methods, SVM with kernel | More complex real-world datasets |
| No clear boundary | More features or more complex models | — |
:::

### Step 7 — Correlation Analysis

**Correlation** measures the linear relationship between two numeric variables. It ranges from −1 to +1.

| Correlation value | Meaning |
|---|---|
| Close to **+1** | Strong positive: as X goes up, Y goes up |
| Close to **0** | No linear relationship |
| Close to **−1** | Strong negative: as X goes up, Y goes down |

```{r iris-step7}
corr_mat <- cor(iris[, -5])    # compute correlation matrix (exclude Species)
corr_mat                        # show the numbers
corrplot.mixed(corr_mat)        # visualise with circles and numbers
```

**Correlation values in the Iris dataset:**

| Pair | Correlation | Strength |
|---|---|---|
| Petal.Length ↔ Petal.Width | **0.96** | Very strong positive |
| Sepal.Length ↔ Petal.Length | **0.87** | Strong positive |
| Sepal.Length ↔ Petal.Width | **0.82** | Strong positive |
| Sepal.Width ↔ Petal.Length | **−0.43** | Moderate negative |
| Sepal.Length ↔ Sepal.Width | **−0.12** | Very weak / negligible |

::: {.callout-warning}
## ⚠️ Correlation ≠ Causation

A correlation of 0.96 between Petal.Length and Petal.Width does not mean one causes the other. Both are driven by the underlying species and overall flower size. Always interpret correlations in the context of your domain knowledge.

Correlation tells you *what* varies together. It says nothing about *why*.
:::

### Step 8 — Pairs Plot (Everything at Once)

The `pairs.panels()` function from the `psych` package combines all of the above into a single view:

- **Diagonal:** histogram + density curve for each variable
- **Below diagonal:** scatter plots for every pair of variables
- **Above diagonal:** correlation coefficients (the numbers)

```{r iris-step8}
pairs.panels(iris[, -5],
             method   = "pearson",     # Pearson correlation
             hist.col = "#00AFBB",     # histogram fill colour
             density  = TRUE,          # show density curve on histogram
             ellipses = FALSE)         # no confidence ellipses
```

This is one of the most efficient exploratory tools available — a full picture of all pairwise relationships in a single plot. Use it early in any data exploration.

---

## The Titanic Dataset — Binary Classification Example {#titanic}

**The question:** Can machine learning predict which passengers survived the Titanic shipwreck?

This is a **binary classification** problem: the output is 0 (did not survive) or 1 (survived).

*Professor's note:* "The Titanic dataset is more complicated than the Iris dataset. There are more columns, but we also have decisions to make about how we're going to manage the various columns."

::: {.callout-note}
## 💡 Binary Classification Encoding Convention

*Professor's note:* "The default [convention] is usually that the more interesting or the most unusual class is the one that gets the value of 1, and the normal case gets the value of 0. So:

- Fraud detection: 0 = non-fraud (common), 1 = fraud (unusual case we care about)
- Medical: 0 = healthy (common), 1 = sick (unusual case we care about)
- Titanic: 0 = did not survive (majority), 1 = survived (outcome of interest)"
:::

### Loading the Dataset

```{r titanic-load}
data(titanic_train)
str(titanic_train)   # 891 observations, 12 variables
```

### Variable Dictionary with ML Decisions

| Variable | Definition | Values / Notes | ML Decision |
|---|---|---|---|
| `PassengerId` | Unique row ID | 1–891 (every row unique) | **Drop** — unique IDs have zero predictive value |
| `Survived` | Did they survive? | 0 = No, 1 = Yes | **Keep** — this is the **target variable**; convert to factor |
| `Pclass` | Ticket class | 1 = 1st, 2 = 2nd, 3 = 3rd | **Keep** — very important; convert to factor |
| `Name` | Full passenger name | 891 unique values | **Drop raw** — but **extract title** (Mr/Mrs/Dr/Miss/Master) |
| `Sex` | Gender | male / female | **Keep** — women were much more likely to survive |
| `Age` | Age in years | 89 unique values; **177 missing (NA)** | **Impute** (fill in) — use mean from training data |
| `SibSp` | # siblings/spouses aboard | 0–8 | **Investigate** — family size may matter |
| `Parch` | # parents/children aboard | 0–6 | **Investigate** — family size may matter |
| `Ticket` | Ticket number | 681 unique (not 891) | **Drop raw** — but some share tickets (family groups) |
| `Fare` | Price paid | 248 unique values | **Keep** — proxy for wealth/class |
| `Cabin` | Cabin number | **687 blanks (77% missing!)** | **Extract deck letter** (A, B, C...) as a new feature |
| `Embarked` | Port of boarding | C=Cherbourg, Q=Queenstown, S=Southampton; 2 blank | **Keep** — easy to fill 2 missing values; convert to factor |

### Checking Unique Values and Missing Data

```{r titanic-eda}
# How many unique values does each column have?
# If unique count = total rows, that column is a unique ID → drop it
uniq_len <- function(x) { length(unique(x)) }
sapply(titanic_train, uniq_len)

# How many blank / NA values does each column have?
countBlank <- function(x) { sum(is.na(x) | x == "") }
sapply(titanic_train, countBlank)
```

*Professor's note:* "If you have a column that has a unique value for every single row inside it — **drop it**. It has to be dropped, because there is nothing informative in that column that a machine learning algorithm can use to differentiate the data."

### Titanic Survival Statistics — Cross-Tabulation

The professor demonstrated live in class. You can reproduce this:

```{r titanic-crosstab}
# Survival rate by gender
table(titanic_train$Survived, titanic_train$Sex)
# Result: males ~20% survival rate; females ~75% survival rate

# Survival rate by ticket class
table(titanic_train$Survived, titanic_train$Pclass)
# Result: 1st class >50% survived; 3rd class <20% survived
```

::: {.callout-important}
## ⭐ Titanic Cross-Tabulation Results — Know These Numbers

| Group | Survived | Did Not | Survival Rate |
|---|---|---|---|
| Male | 109 | 468 | ~19% |
| Female | 233 | 81 | ~74% |
| 1st class | >50% | <50% | >50% |
| 2nd class | ~47% | ~53% | ~47% |
| 3rd class | 119 | ~372 | ~24% |

*Professor's note:* "You can imagine if you are third class and an adult male, your chances of survival were very, very, very low."

These numbers are the result of simple cross-tabulation — no machine learning algorithm was needed to find this insight. It is a perfect example of why data exploration comes before modelling.
:::

---

## Feature Engineering — Transforming Raw Data for ML {#feature-engineering}

**Feature engineering** is the process of transforming raw data into features (columns) that are more useful for machine learning. It happens *after* data exploration and *before* model training.

::: {.callout-important}
## ⭐ Exam Tip — Feature Engineering > Algorithm Choice

*Professor's note:* "Feature engineering is more important than your choice of algorithm. If you're given a realistic dataset, feature engineering and picking an underperforming algorithm will outperform just blindly applying random forests or gradient-boosted machines to the raw data."

This is a direct, quotable professor statement. Know it. Be able to justify it.
:::

### Common Feature Engineering Tasks

| Task | Example from Titanic |
|---|---|
| **Remove irrelevant columns** | Drop `PassengerId`, raw `Name` |
| **Fix column types** | Convert `Pclass` from `integer` → `factor` |
| **Handle missing values** | Impute `Age` with mean; decide what to do with `Cabin` |
| **Extract information from text** | Pull title ("Mr", "Mrs", "Dr", "Master") out of `Name` |
| **Create new binary features** | Create `is_child` (TRUE if Age < 18) from `Age` |
| **Log-transform skewed columns** | Apply `log()` to right-skewed numeric variables |
| **Scale numeric variables** | Normalise so all variables are on a similar scale |

::: {.callout-note}
## 💡 Why Scaling Matters

If you have `Age` (values 0–80) and `Fare` (values 0–500) in the same model without scaling, some algorithms will treat `Fare` as more important simply because its numbers are bigger. Scaling puts all variables on the same scale — for example, 0 to 1, or mean 0 with standard deviation 1.

*Professor's note (Week 7):* "Because these things [iris measurements] are measuring the same part of the flower and they are all in the same unit of measurement, it doesn't really matter whether they're inches or centimetres — [for Iris] it doesn't matter. But in a real dataset where you have Age (years) and Fare (dollars), scaling absolutely matters."
:::

### How Your Week 2 Titanic Observations Become Week 5 Features

::: {.callout-note}
## 🔮 Forward Link — Your EDA Findings Turn Into Features

This is one of the most important concepts of the course. Every observation you made about the Titanic dataset in Week 2 becomes a concrete feature engineering task in later weeks:

| Week 2 Observation | Week 5/6 Feature Engineering |
|---|---|
| "Age has 177 missing values (NAs)" | Impute missing ages using the mean from the training set (not the full dataset!) |
| "Cabin has 687 blanks (77% missing)" | Extract the first letter as a "deck" feature (A, B, C…); the rest treat as missing |
| "Name has 891 unique values" | Extract the **title** (Mr / Mrs / Miss / Master / Dr) — this reveals gender + age hints |
| "Mr Miss Master Dr are titles" | Discovery: "Master" is used for young boys — helps predict age for boys with missing Age |
| "Age becomes significant mainly for children" | Create `is_child` (Age < 18) as a binary feature — it captures the real signal |
| "SibSp and Parch are family sizes" | Create `family_size = SibSp + Parch + 1` or `is_alone` (family_size == 1) |

**Better age imputation (Week 5):** Rather than just using one mean for all missing ages, compute a mean age **by ticket class AND title**:

- 1st class Miss: average age ~29.7
- 1st class Mr: average age ~42.4
- 1st class Mrs: average age ~40.4
- 1st class Master: average age ~5.3 ← children!

Then plug in the appropriate mean for each missing record. This gives a much better estimate than a blanket "average age = 29.7" imputation.
:::

::: {.callout-warning}
## ⚠️ Always Verify Data Encoding

*Professor's note:* "I'm told survival is either 0 or 1. I certainly should be checking that. I'm told ticket class is 1, 2, or 3 — I should be checking that. Is the spread of age reasonable? Do we have any person with a negative age, or age over 120?"

Never trust the data dictionary alone. Always verify using `table()`, `unique()`, and `summary()`.
:::

---

## Summary Checklist ✅

- ✅ I understand CRISP-DM and can explain why only 10–15% of project time is spent on modelling
- ✅ I can explain the PayPal churn case study and what it teaches about business understanding and defining the right problem
- ✅ I know the difference between mean and median, and when each is more appropriate
- ✅ I understand quartiles, IQR, and how a box plot is constructed and read
- ✅ I can identify left skew, right skew, and normal distributions from a plot
- ✅ I know the rule: **right skew → mean > median; left skew → mean < median**
- ✅ I understand why skewed data is a problem for ML and how log / Box-Cox transforms help
- ✅ I can reproduce the mammals example: R² went from 0.87 → 0.92 after log transform (37.79% fewer errors)
- ✅ I understand what a factor data type is in R and why getting column types correct matters
- ✅ I can run a full 8-step data exploration workflow in R on the Iris dataset
- ✅ I know the Iris bimodal finding — petal measurements reveal two distinct sub-populations
- ✅ I can interpret a correlation matrix and correlation plot
- ✅ I know the Titanic variable dictionary and what to do with each column
- ✅ I know the Titanic survival cross-tabulation results (~19% men, ~74% women, <20% 3rd class)
- ✅ I understand feature engineering and why it is more important than algorithm choice
- ✅ I can explain human-in-the-loop machine learning and the fraud detection use case
- ✅ I understand **interpretation > production** — producing plots is only useful if I interpret them
- ✅ I can spot visual patterns in scatter plots (horizontal/vertical/sloped lines → different ML approaches)
- ✅ I can explain how Week 2 EDA insights become concrete features in later weeks

---

## Exercise/Exam Scenarios Table

This table maps common exercise and exam scenarios to the relevant Week 2 concept:

| Scenario you might be given | Concept to apply | Answer pattern |
|---|---|---|
| "Mean > Median in a dataset" | Skewness direction | Right (positive) skew |
| "Mean < Median in a dataset" | Skewness direction | Left (negative) skew |
| "Histogram shows two humps" | Bimodal distribution | Two sub-populations; consider splitting the data |
| "Column has unique value for every row" | Unique ID column | Drop it — no predictive value |
| "Column has >70% missing values" | Heavy missing data | Either drop column, extract partial info (e.g., deck letter from Cabin), or treat missingness itself as a feature |
| "Column has <5% missing values" | Light missing data | Impute — use median (for skewed) or mean (for normal) |
| "Model has 99%+ accuracy but catches no fraud" | Imbalanced dataset | Use human-in-the-loop; report precision and recall, not just accuracy |
| "Predicting churn — should I use cancellation date?" | Data leakage | No — it's known only after the event you're predicting |
| "R² improves from 0.80 to 0.90 after transformation" | Relative error reduction | Error dropped from 0.20 to 0.10 — that's a 50% error reduction |
| "Scatter plot shows a clear horizontal gap" | Perfectly separable groups | Simple if-statement or decision tree will work |
| "Algorithm treats class 3 as 'worse' than class 1" | Wrong data type | Convert integer → factor |
| "Income data is very skewed" | Right skew | Report median instead of mean; consider log transform |

---

## Pre-Topic Checklist for Week 3

Before moving on to Decision Trees, make sure you can:

- [ ] Load all required packages in R without errors
- [ ] Call `str()`, `dim()`, `names()`, `summary()` on a dataset and interpret every line of output
- [ ] Write and run the `drawPlots` function from memory
- [ ] Create histograms, density plots, and box plots using `qplot()`
- [ ] Combine multiple plots with `gridExtra::grid.arrange()`
- [ ] Identify skewness from a plot and state the mean/median rule
- [ ] Apply `log()` to a skewed column and verify the result looks more normal
- [ ] Use `cor()` and `corrplot.mixed()` to produce and interpret a correlation matrix
- [ ] Use `pairs.panels()` for a comprehensive multivariate view
- [ ] Write and use the `uniq_len` and `countBlank` functions on a new dataset
- [ ] Describe all 12 Titanic variables and state what you would do with each one
- [ ] Run `table()` for cross-tabulation and interpret the output
- [ ] Explain what a factor is, and convert a column to factor with `as.factor()`
- [ ] Explain the `if (Petal.Length < 2) → setosa` rule and why it matters
- [ ] Interpret a plot (don't just produce one) — always ask "what decision does this lead to?"

---

## Practice Questions

::: {.callout-important}
## ⭐ Practice Questions with Full Answers

---

**Q1:** If the mean of a dataset is much larger than the median, what type of skew does the distribution have? How can you tell from a plot?

*Answer:* **Right (positive) skew.** A few very large values pull the mean upward but leave the median (middle value) relatively unchanged. On a histogram or density plot, the data clusters on the left with a long tail extending to the right. On a `drawPlots` output, the blue line (mean) will be to the right of the green line (median).

---

**Q2:** You are building a churn prediction model. A colleague suggests using "date account was closed" as a predictor. What is the problem?

*Answer:* This is **data leakage**. The account closing date only exists *after* the churn event. At the time you would want to make a prediction (i.e., before the customer leaves), this value is not yet known. Including it gives artificially high model performance in testing, but the model would fail completely in production because the feature would be unavailable.

---

**Q3:** The Iris dataset shows a bimodal distribution in Petal.Length. What causes this, and why does it matter for ML?

*Answer:* There are two distinct groups (sub-populations) in the data. The setosa species has much shorter petals (around 1–2cm) than versicolor and virginica (3–7cm). On a histogram or density plot, this appears as two separate humps. For ML, this is valuable: it suggests the data is naturally separable, and petal length will be a very powerful classification feature. A simple rule (`if Petal.Length < 2, predict setosa`) is already 100% accurate. This rule becomes the root split of a decision tree in Week 3.

---

**Q4:** A mammal body weight model gives R² = 0.87 on raw data and R² = 0.92 after log-transforming both variables. How much did the error rate improve?

*Answer:* **37.79% reduction in error rate.** Raw error = 1 − 0.87 = 0.13. Transformed error = 1 − 0.92 = 0.08. Improvement = (0.13 − 0.08) / 0.13 = 0.05 / 0.13 ≈ 38% (stated as 37.79% in the slides). The algorithm did not change — only the data preparation did. This demonstrates why feature engineering matters more than algorithm choice.

---

**Q5:** Why is `PassengerId` dropped from the Titanic ML model?

*Answer:* It has 891 unique values for 891 rows — every row has a different value. A column that is unique for every row contains no pattern that a machine learning algorithm can learn from. It cannot help predict survival. In general, any column that functions as a **primary key** or unique identifier should be removed before model training.

---

**Q6:** A fraud detection model predicts every transaction as "not fraudulent" and achieves 99.999% accuracy. Is this a good model? What should you do instead?

*Answer:* No — this is useless. Because fraud is extremely rare, predicting "not fraudulent" for everything will be technically accurate but catches zero actual fraud cases. This is the **imbalanced dataset** problem. A better approach is to use the model as a filter: flag the top 1–2% of most suspicious transactions for human review. This is called **human-in-the-loop machine learning**. The model is not the final arbiter — it is a prioritisation tool.

---

**Q7:** What is the difference between a `character` and a `factor` in R? Why does it matter?

*Answer:* A `character` is a free-text string — any value is valid. A `factor` is a categorical variable with a fixed set of allowed values (called "levels"). For example, `Species` in the iris dataset is a factor with levels {setosa, versicolor, virginica} — R knows no other value is valid. It matters for ML because some algorithms require categorical variables to be stored as factors, not strings. If a column like `Pclass` (1, 2, 3) is stored as integer, an algorithm may treat it as a continuous number; as a factor, it is treated as a category.

---

**Q8:** The Age column in the Titanic dataset has 177 missing values (NA). What are your three options, and which is usually best?

*Answer:* Three options:

1. **Drop the column entirely.** Not good here — Age is an important predictor (children were prioritised for lifeboats).
2. **Drop the rows with missing Age.** Not good — you lose 177 records (20% of data), and your production model can't handle new records with missing Age.
3. **Impute** — fill in the missing values. This is usually best. You can use a blanket mean (simple), or better, a grouped mean (e.g., mean age by ticket class and title — giving much better estimates).

**Important:** The mean must be calculated on the **training set only**, then applied to both train and test sets. Using the full dataset's mean to impute is data leakage.

---

**Q9:** You discover a scatter plot where you can draw a **sloped** line that perfectly separates two classes. What ML model would you try first, and why?

*Answer:* **Linear regression** (for a numeric target) or **logistic regression** (for a binary classification target). A sloped line is a linear decision boundary — that's exactly what these models produce. If you could draw a horizontal/vertical line (axis-aligned), a simple `if` statement or shallow decision tree would suffice. If the boundary is curved, you need more flexible models (trees, SVMs, neural networks).

---

**Q10:** What does the professor mean when he says "if you're just producing plots and jumping straight into ML, you're wasting your time"?

*Answer:* Exploratory plots are only valuable if you **interpret** them — draw conclusions, form hypotheses, and let them guide your feature engineering. Examples from class: the setosa bimodal distribution → create an `is_setosa` rule; the Age distribution → identify missing values; the Cabin blanks → extract deck letters. A plot without interpretation teaches you nothing. A plot with interpretation can change your entire modelling approach.
:::

---

## Additional Resources

| Resource | Link |
|---|---|
| Iris dataset (Wikipedia) | <https://en.wikipedia.org/wiki/Iris_flower_data_set> |
| ggplot2 documentation | <https://ggplot2.tidyverse.org/> |
| corrplot package | <https://cran.r-project.org/web/packages/corrplot/> |
| Kaggle Titanic competition | <https://www.kaggle.com/c/titanic> |
| Kaggle: filter R notebooks by most votes | Sort by Most Votes; filter Language = R |
| Kaggle Spaceship Titanic (fun variation) | <https://www.kaggle.com/c/spaceship-titanic> |
| Kaggle House Prices | <https://www.kaggle.com/c/house-prices-advanced-regression-techniques> |
| R for Data Science (free book) | <https://r4ds.had.co.nz/> |
| StatQuest YouTube (visual explanations) | Search "StatQuest" on YouTube |
| Box-Cox transformation reference | Search "Box-Cox R caret preProcess" |

*Professor's tip:* On Kaggle's Titanic competition, filter notebooks by language = R and sort by "Most Votes" to find high-quality examples.

---

## Concept Relationship Map

The key ideas from Week 2 connect like this:

```
CRISP-DM Process
│
├── Business Understanding
│     └── PayPal Case: define the right problem first
│
├── Data Understanding + Data Preparation (most of your time)
│     │
│     ├── Summary Statistics
│     │     ├── Mean vs Median → sensitive vs robust to outliers
│     │     └── Quartiles / IQR → used in box plots
│     │
│     ├── Univariate Plots (INTERPRET, don't just produce)
│     │     ├── Histogram → shape, frequency, bimodal detection
│     │     ├── Density plot → smooth shape comparison
│     │     └── Box plot → median, IQR, outliers (statisticians only!)
│     │
│     ├── Skewness
│     │     ├── Right skew: Mean > Median → tail right
│     │     ├── Left skew: Mean < Median → tail left
│     │     └── Fix: log transform (simple) or Box-Cox (accurate)
│     │
│     ├── Multivariate Analysis
│     │     ├── Scatter plots → relationships between two variables
│     │     │     (horizontal line = if-statement; sloped = linear; curved = tree)
│     │     ├── Correlation matrix (cor()) → strength of linear relationships
│     │     └── pairs.panels() → everything in one view
│     │
│     └── Feature Engineering (informed by EDA)
│           ├── Drop: unique ID columns, irrelevant text
│           ├── Convert: integer→factor for categorical variables
│           ├── Impute: fill in missing values (Age in Titanic, by title + class)
│           ├── Transform: log() for skewed variables
│           ├── Extract: title from Name, deck letter from Cabin
│           └── Create: binary features like is_child, is_alone
│
└── Modelling (~10-15% of project time)
      │
      └── Forward to later weeks:
            ├── Week 3: Decision trees use the if-statement rules you found
            ├── Week 5/6: Titanic features you engineered become model inputs
            ├── Week 6: Data leakage — transformations on training data only
            └── Week 7: Clustering confirms the groups you saw in EDA
```

---

## Annotation Space

*Use the space below for your own notes, questions, or diagrams during or after class:*

\

\

\

\

\

\

\

\

\

\

\

\

---

*Enhanced study guide v2, based on STAT9013 Week 2 lecture slides, class transcript (closed captions), week2_code.R, and forward references from Week 3, Week 5, Week 6, and Week 7 materials. Module: Data Mining and Machine Learning. Professor: Mark Hodnett (mark.hodnett@mtu.ie). Contact via Canvas or email.*
