---
title: "Week 10 – ANCOVA (Analysis of Covariance)"
subtitle: "STAT9004: Regression Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    highlight-style: github
    code-fold: true
    code-tools: true
    self-contained: true
execute:
  warning: false
  message: false
---

::: {.callout-important}
## ⚠️ Assessment Reminders

- **Lab Exam** — Week 13, approximately **6th of May**. Professor said: *"When you're prepping for the lab exam, you're also prepping for the final exam."*
- **Final Exam** — timetable to be confirmed by the exams office. Check Canvas and your email.
- **Coursework (25%)** — Group critique assignment, due **Week 7**.
- **End of Module Exam (50%)** — covers all topics including ANCOVA.

> 💡 Review the **past exam paper** (`STAT9004_Exam_19.pdf`) on Canvas — the Professor confirmed exam questions follow the same format as the R output exercises in these notes.
:::

---

# Key Topics Covered This Week

- What ANCOVA is and why we use it
- The difference between a **covariate** and a **factor**
- The ANCOVA model formula (with and without interaction)
- How to identify the **experimental design** in a question
- Model **assumptions** (independence, normality, equal variance)
- How to **interpret R output** from an ANCOVA model
- Hypothesis testing for each term in the model
- The **Parsimony Principle** — dropping non-significant terms
- Comparing **R²** vs **Adjusted R²**
- The **Birds dataset** exercise with log transformations

---

# What is ANCOVA?

## The Big Picture

**ANCOVA** stands for **Analysis of Covariance**. The Professor put it simply in class:

::: {.callout-note icon=false}
## 💬 Professor's Note
*"It's ANOVA with a C — and the C stands for Covariance. That's a perfect way to describe it."*
:::

Think of it as a combination of the two methods you already know:

| Method | Explanatory Variable Type |
|--------|--------------------------|
| **ANOVA** | Categorical only (e.g. treatment groups) |
| **Linear Regression** | Continuous only (e.g. size, temperature) |
| **ANCOVA** | **Both at the same time** |

ANCOVA is the right tool when you have **one main categorical variable** you care most about, *plus* a **continuous variable** that also influences the outcome but isn't your main focus.

## What is a Covariate?

A **covariate** is a continuous (quantitative) variable that:

- You are **not primarily interested in**, but
- You **include in the model** to **control for its effect** on the response variable

::: {.callout-note icon=false}
## 💡 Analogy — Why Include Something You Don't Care About?
Think of a race where you want to compare two training programmes. But some runners are naturally faster. If you ignore natural speed when comparing programmes, your results will be unfair. The covariate (natural speed) is included to *level the playing field* and make the comparison between the two programmes fair.
:::

::: {.callout-note icon=false}
## 💬 Professor's Note
*"The categorical explanatory variables are of major interest. The continuous variable is of lesser interest — but it's included to control for its effect on the response variable."*
:::

---

# The Fruit & Grazing Example

## The Experiment

This is the central example for Week 10. The Professor confirmed the exam format mirrors this kind of analysis — you will be given R output and asked to interpret it.

**Setup:**

- 40 plants split into two groups:
  - **Grazed** — exposed to rabbits for the first two weeks of stem growth
  - **Ungrazed** — left alone, then fenced
- At the end of the growing season, **fruit production** (dry weight in mg) was recorded

**The three variables:**

| Variable | Type | Role in Model |
|----------|------|--------------|
| `Fruit` | Continuous (mg) | **Response variable** — what we are measuring |
| `Grazing` | Categorical (2 levels) | **Main explanatory variable** — what we care about |
| `Root` | Continuous | **Covariate** — initial plant size (diameter of rootstock) |

## Why Not Just Do a t-test?

::: {.callout-warning}
## ⚠️ Common Mistake — Using a t-test When You Shouldn't
Your first instinct might be: *"Just split the data into grazed and ungrazed and run a t-test."*

**This is wrong.** The Professor emphasised this in class.

The problem: the amount of fruit produced also depends on the **initial size of the plant**. A bigger plant will produce more fruit regardless of grazing. Ignoring root size makes the group comparison unfair.
:::

::: {.callout-note icon=false}
## 💬 Professor's Note
*"This is not the correct way to proceed, since the amount of fruit produced may also depend on the initial size of the plant."*
:::

## The Counter-Intuitive Result

Without ANCOVA, a t-test shows that **grazed plants produce significantly MORE fruit**. This seems biologically wrong — rabbits eating plants shouldn't increase fruit production!

The reason this happens: the **grazed plants happened to be larger** to begin with. Once ANCOVA adjusts for initial size (Root), the misleading signal disappears. This is exactly why the covariate matters.

---

# The Experimental Design

::: {.callout-important}
## ⭐ Exam Skill — Identifying Experimental Design
The Professor said: *"These are the types of questions that you should be thinking about when you're looking back through the notes."* Given a scenario, you must be able to identify all design components.
:::

## Design Components — What to Identify

For any experiment, ask yourself these questions:

| Question | Answer for This Example |
|----------|------------------------|
| What are the **experimental units**? | The 40 plants |
| How many **factors**? | One — Grazing |
| What are the **levels** of each factor? | Grazed and Ungrazed (2 levels) |
| What are the **treatments**? | Grazed and Ungrazed |
| Is there a **covariate**? | Yes — Root (initial plant size) |
| What is the **response variable**? | Fruit production (mg) |

## Design Classification

This experiment is:

- A **Single-Factor Design** — only one independent variable (Grazing)
- A **Completely Randomized Design (CRD)** — each plant has an equal probability of being assigned to either treatment

::: {.callout-note icon=false}
## 💡 What Does "Completely Randomized" Mean?
It means that the treatment (grazed or ungrazed) is assigned to each plant **at random** — like flipping a fair coin for each one. No plant is more likely to end up in one group than the other. This is the simplest and most common experimental design.
:::

## Factorial Design — What Would Change?

| Design Type | What It Requires |
|------------|-----------------|
| **Single-Factor** | 1 factor, 2+ levels |
| **Factorial** | 2+ factors, each with 2+ levels |
| **Full Factorial** | All possible combinations of factor levels tested (all main effects + all interactions) |

> **Example:** To make this a full factorial design, you could add a second factor like `Soil Type` (e.g. Sandy / Clay). Then all combinations — Grazed+Sandy, Grazed+Clay, Ungrazed+Sandy, Ungrazed+Clay — would be tested.

---

# Assumptions

::: {.callout-important}
## ⭐ Exam Skill — Checking and Interpreting Assumptions
The Professor said: *"Ensuring that these assumptions are valid is an extremely important part of the design of the experiment."* And: *"Make sure that you're able to interpret the plots from R."*
:::

These three assumptions are the same ones you have seen in ANOVA and regression. They all apply to ANCOVA.

## 1. Independence of Errors

- Each observation must be **independent** — one result must not influence another
- **How to check:** Plot residuals in **time order** of data collection — a positive pattern = violation
- **How to ensure:** Proper **randomisation** before the experiment begins

## 2. Normality of Errors

- The errors (residuals) must follow a **normal distribution**
- **How to check:**

| Tool | What to Look For |
|------|-----------------|
| **QQ plot** | Points should fall on a straight diagonal line |
| **Histogram of residuals** | Should look like a bell curve |
| **Shapiro-Wilk test** | p-value > 0.05 → no evidence against normality |
| **Kolmogorov-Smirnov test** | Alternative normality test |

::: {.callout-note icon=false}
## 💡 Which Tool to Use?
The histogram is better for large datasets (you can see the shape clearly). The QQ plot is useful at any sample size. A useful guide for interpreting QQ plots is linked in the worksheet: [http://seankross.com/2016/02/29/A-Q-Q-Plot-Dissection-Kit.html](http://seankross.com/2016/02/29/A-Q-Q-Plot-Dissection-Kit.html)
:::

## 3. Homogeneity of Variance (Equal Variance)

- The **spread** (variance) of residuals should be the same across all groups and fitted values
- **How to check:**

| Tool | What to Look For |
|------|-----------------|
| **Levene's Test** | p-value > 0.05 → equal variances |
| **Residuals vs Fitted plot** | No pattern; points scattered randomly around zero |

---

# The ANCOVA Model

## Full Model (With Interaction)

$$Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ + e$$

| Symbol | Meaning in This Example |
|--------|------------------------|
| $Y$ | Fruit production (response) |
| $X$ | Root size — the **continuous** covariate |
| $Z$ | Dummy variable: **0 = Grazed, 1 = Ungrazed** |
| $XZ$ | Interaction between Root and Grazing |
| $e$ | Error (residual) term |

::: {.callout-note icon=false}
## 💬 Professor's Note
*"Zero is grazed — that's your reference group. The same as the previous models you've run with ANOVA, you have one reference group."*
:::

## What Each Coefficient Means

| Coefficient | Plain-English Meaning |
|-------------|----------------------|
| $\beta_0$ | Expected Fruit when Root = 0, for **Grazed** plants (the reference group) |
| $\beta_1$ | For every **1-unit increase in Root**, Fruit increases by $\beta_1$ — for **Grazed** plants |
| $\beta_2$ | **Difference in intercept** when switching from Grazed → Ungrazed |
| $\beta_3$ | **Difference in slope** when switching from Grazed → Ungrazed (the interaction) |

## Model Equations for Each Group

The power of ANCOVA is that it gives you a **separate regression line for each group**, derived from one model.

**For Grazed plants** (Z = 0 — the reference group):

$$\text{Fruit} = \beta_0 + \beta_1 \cdot \text{Root} + e$$

**For Ungrazed plants** (Z = 1):

$$\text{Fruit} = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) \cdot \text{Root} + e$$

::: {.callout-note icon=false}
## 💡 How to Remember This
When Z = 0 (Grazed), the $\beta_2$ and $\beta_3$ terms both drop out (multiplied by 0).  
When Z = 1 (Ungrazed), they stay in and **adjust** both the intercept and the slope.
:::

---

# Reading the R Output

::: {.callout-important}
## ⭐ Exam Skill — Step-by-Step Output Interpretation
The Professor confirmed: *"In a lot of exam questions, they will ask you to state the model and state the hypotheses. You generally get the R output and have to put together the model or interpret the outputs."*

**Always follow this order:**
1. Overall model significance (F-test at the bottom)
2. Interaction term first
3. Main effects (only if interaction is not significant)
:::

## Full Model (With Interaction): `lm(Fruit ~ Root * Grazing)`

```r
lm(formula = Fruit ~ Root * Grazing, data = fruit)

Coefficients:
                     Estimate Std. Error t value Pr(>|t|)
(Intercept)          -125.173    12.811   -9.771  1.15e-11 ***
Root                   23.240     1.531   15.182  < 2e-16 ***
GrazingUngrazed        30.806    16.842    1.829   0.0757 .
Root:GrazingUngrazed    0.756     2.354    0.321   0.7500

Residual standard error: 6.831 on 36 degrees of freedom
Multiple R-squared:  0.9293,    Adjusted R-squared:  0.9234
F-statistic: 157.6 on 3 and 36 DF,  p-value: < 2.2e-16
```

### Step-by-Step Interpretation

**Step 1 — Overall model (F-statistic, bottom line):**  
p-value < 2.2e-16 → much less than 0.05 → **the overall model is statistically significant**. At least one predictor has an effect on fruit production.

**Step 2 — Interaction term (`Root:GrazingUngrazed`):**  
p-value = 0.75 → much greater than 0.05 → **not significant**. The slopes for grazed and ungrazed plants are **not different**. The two regression lines are parallel.

**Step 3 — Main effects:**

| Term | Estimate | p-value | Significant? | Interpretation |
|------|----------|---------|-------------|----------------|
| `(Intercept)` | −125.173 | 1.15e-11 | ✅ Yes | Expected Fruit when Root = 0, for Grazed plants. Not scientifically meaningful (Root = 0 doesn't exist). |
| `Root` | 23.240 | < 2e-16 | ✅ Yes | For every 1-unit increase in Root, Fruit increases by 23.24 mg — for Grazed plants |
| `GrazingUngrazed` | 30.806 | 0.0757 | ❌ No (at 5%) | Ungrazed plants produce ~31 mg more fruit than Grazed plants — but this is **not significant** in this model |
| `Root:GrazingUngrazed` | 0.756 | 0.7500 | ❌ No | The effect of Root on Fruit does NOT differ between Grazed and Ungrazed |

## Fitted Equations From This Model

**Overall:**
$$\hat{\text{Fruit}} = -125.173 + 23.24 \cdot \text{Root} + 30.806 \cdot Z + 0.756 \cdot Z \cdot \text{Root}$$

**Grazed** (Z = 0):
$$\hat{\text{Fruit}} = -125.173 + 23.24 \cdot \text{Root}$$

**Ungrazed** (Z = 1):
$$\hat{\text{Fruit}} = (-125.173 + 30.806) + (23.24 + 0.756) \cdot \text{Root} = -94.367 + 23.996 \cdot \text{Root}$$

---

# Hypothesis Tests

::: {.callout-important}
## ⭐ Exam Skill — Writing Hypotheses
The Professor: *"In a lot of exam questions, they will ask you to state the model and state the hypotheses."* For any term in the model, you need to be able to write $H_0$ and $H_A$ correctly.
:::

## All Hypotheses You Need to Know

### Test 1 — Overall Model (Global F-test)
$$H_0: \beta_1 = \beta_2 = \beta_3 = 0 \quad \text{(no predictor has any effect)}$$
$$H_A: \text{at least one } \beta_i \neq 0$$

**How to find it in R output:** The F-statistic and p-value at the very bottom of `summary()`.

### Test 2 — Interaction Term (Are the slopes the same?)
$$H_0: \beta_3 = 0 \quad \text{(lines are parallel — same slope)}$$
$$H_A: \beta_3 \neq 0 \quad \text{(lines have different slopes)}$$

::: {.callout-note icon=false}
## 💡 Plain English
This test asks: *does the effect of Root on Fruit change depending on whether a plant is grazed or not?*  
If $H_0$ is not rejected → the effect is the same → the lines are **parallel**.
:::

### Test 3 — Root (Covariate)
$$H_0: \beta_{\text{Root}} = 0 \quad \text{(Root has no effect on Fruit)}$$
$$H_A: \beta_{\text{Root}} \neq 0$$

### Test 4 — Grazing (Main Factor)
$$H_0: \beta_{\text{Grazing}} = 0 \quad \text{(Grazing has no effect on the intercept)}$$
$$H_A: \beta_{\text{Grazing}} \neq 0$$

### Test 5 — Do the Lines Coincide? (Same slope AND same intercept)
$$H_0: \beta_2 = \beta_3 = 0$$
$$H_A: \beta_2 \neq 0 \text{ or } \beta_3 \neq 0$$

---

# The Parsimony Principle

## What It Means

::: {.callout-note icon=false}
## 💬 Professor's Note
*"If we have two or more possible models, each of which fit the data adequately, the one with the fewest explanatory variables is preferred."*
:::

**Simpler is better** — as long as the simpler model fits the data just as well.

::: {.callout-warning}
## ⚠️ Important Nuance
The Professor highlighted: *"This isn't always the case. You may actually want to keep non-significant terms in your model, depending on whether they're of scientific interest."* This is a principle, not a rule.
:::

## How to Drop Terms

1. Identify all **non-significant** terms (p-value > 0.05)
2. Start by dropping the term with the **highest p-value** (most insignificant)
3. Rerun the model and check again
4. Continue until all remaining terms are significant (or you have a reason to stop)

In our example: `Root:GrazingUngrazed` has p = 0.75 — the highest p-value. Drop it first.

## Comparing R² and Adjusted R²

After removing the interaction term:

| Metric | With Interaction | Without Interaction | Change |
|--------|-----------------|---------------------|--------|
| **R²** | 0.9293 | 0.9291 | Decreased ↓ |
| **Adjusted R²** | 0.9234 | **0.9252** | Increased ↑ ✅ |

::: {.callout-important}
## ⭐ Key Exam Concept — R² vs Adjusted R²

- **R²** *always* decreases when you remove a term — even if that term was completely useless. Do not use R² alone to compare models of different sizes.
- **Adjusted R²** penalises for adding extra terms. It goes *up* when you drop a term that wasn't helping.
- A small **increase in Adjusted R²** after removing a term confirms the term was not contributing to the model.

The Professor: *"The adjusted R-squared is more appropriate here as it takes into account the number of terms in the model."*
:::

## The Simpler (Final) Model: `lm(Fruit ~ Root + Grazing)`

```r
lm(formula = Fruit ~ Root + Grazing, data = fruit)

Coefficients:
                Estimate Std. Error t value Pr(>|t|)
(Intercept)     -127.829      9.664  -13.23  1.35e-15 ***
Root              23.560      1.149   20.51  < 2e-16 ***
GrazingUngrazed   36.103      3.357   10.75  6.11e-13 ***

Residual standard error: 6.747 on 37 degrees of freedom
Multiple R-squared:  0.9291,    Adjusted R-squared:  0.9252
F-statistic: 242.3 on 2 and 37 DF,  p-value: < 2.2e-16
```

All three terms are now significant. The final fitted equations are:

**Grazed** (Z = 0):
$$\hat{\text{Fruit}} = -127.829 + 23.56 \cdot \text{Root}$$

**Ungrazed** (Z = 1):
$$\hat{\text{Fruit}} = (-127.829 + 36.103) + 23.56 \cdot \text{Root} = -91.726 + 23.56 \cdot \text{Root}$$

::: {.callout-note icon=false}
## 💡 Key Observation
In the final model, both lines have the **same slope (23.56)** — they are **parallel**. The only difference is the intercept. Ungrazed plants produce approximately 36.1 mg more fruit than Grazed plants, *at any given root size*.
:::

---

# Practical R Code

```{r}
#| eval: false
#| code-fold: false

library(tidyverse)
library(car)

# --- LOAD DATA ---
fruit <- read_tsv("fruit.txt")
fruit$Grazing <- as.factor(fruit$Grazing)

# --- EXPLORATORY PLOTS ---
# Scatterplot: Fruit vs Root, grouped by Grazing
ggplot(fruit, aes(x = Root, y = Fruit)) +
  geom_point(aes(shape = factor(Grazing)), size = 2)

# Boxplot: Fruit by Grazing treatment
plot(Fruit ~ Grazing, data = fruit)

# --- WRONG APPROACH (for comparison only) ---
# t-test ignores Root — misleadingly shows Grazed > Ungrazed
t.test(Fruit ~ Grazing, data = fruit)

# --- ANCOVA MODEL 1: With Interaction ---
m1 <- lm(Fruit ~ Root * Grazing, data = fruit)
summary(m1)

# --- ANCOVA MODEL 2: Without Interaction (Parsimony Principle) ---
m2 <- lm(Fruit ~ Root + Grazing, data = fruit)
summary(m2)

# Compare models formally
anova(m2, m1)

# --- PLOT PARALLEL REGRESSION LINES ---
ggplot(fruit, aes(Root, Fruit)) +
  geom_point(aes(shape = factor(Grazing)), size = 2) +
  geom_abline(intercept = -127.829, slope = 23.56, linetype = "solid") +   # Grazed
  geom_abline(intercept = -91.726,  slope = 23.56, linetype = "dashed")    # Ungrazed

# --- DIAGNOSTIC PLOTS ---
par(mfrow = c(2, 2))
plot(m2)

# --- ASSUMPTION TESTS ---
leveneTest(Fruit ~ Grazing, data = fruit)   # Equal variance
shapiro.test(residuals(m2))                  # Normality of residuals
```

---

# Exercise: Birds Dataset

This is the **exam-level exercise** from Worksheet 8. It extends ANCOVA to a real ecological dataset with multiple predictors and required transformations.

## Background

The `Birds.txt` dataset records bird population densities in 56 forest patches in south-eastern Victoria, Australia. The goal is to identify which habitat variables affect bird abundance (`ABUND`).

## Variables

| Variable | Type | Description |
|----------|------|-------------|
| `ABUND` | Continuous | Bird population density — **response variable** |
| `AREA` | Continuous | Size of the forest patch |
| `DIST` | Continuous | Distance to the nearest patch |
| `LDIST` | Continuous | Distance to the nearest *larger* patch |
| `ALT` | Continuous | Mean altitude of the patch |
| `YR.ISOL` | Continuous | Year of isolation by land clearing |
| `GRAZE` | Categorical (1–5) | Grazing intensity: 1 = light, 5 = intensive |

## Step-by-Step Worksheet Questions

### Step 1 — Exploratory Data Analysis

- Make **boxplots** and **histograms** for all variables
- Use a **pairs plot** to examine relationships between variables and check for correlations
- Comment on the shape of distributions — are any variables skewed?

```{r}
#| eval: false

Birds <- read.table("Birds.txt", header = TRUE)
Birds$GRAZE <- as.factor(Birds$GRAZE)

# Summary statistics
summary(Birds)

# Boxplots
par(mfrow = c(2, 3))
boxplot(Birds$ABUND, main = "Abundance")
boxplot(Birds$AREA,  main = "Area")
boxplot(Birds$DIST,  main = "Distance")
boxplot(Birds$LDIST, main = "L Distance")
boxplot(Birds$YR.ISOL, main = "Year Isolated")
boxplot(Birds$ALT,  main = "Altitude")

# Histograms
par(mfrow = c(2, 3))
hist(Birds$ABUND, main = "Abundance")
hist(Birds$AREA,  main = "Area")       # Note: likely right-skewed
hist(Birds$DIST,  main = "Distance")   # Note: likely right-skewed
hist(Birds$LDIST, main = "L Distance") # Note: likely right-skewed
hist(Birds$YR.ISOL, main = "Year")
hist(Birds$ALT,  main = "Altitude")
```

### Step 2 — Transformations

::: {.callout-note icon=false}
## 💬 Professor's Note
*"With this dataset, with this exercise, you'll have to do some data transformations — the log of area, the log of distance."*
:::

`AREA`, `DIST`, and `LDIST` are right-skewed (many small values, a few very large ones). Log transformation makes them more symmetric and improves model fit.

```{r}
#| eval: false

# Apply log transformations
Birds$LOGAREA  <- log10(Birds$AREA)
Birds$LOGDIST  <- log10(Birds$DIST)
Birds$LOGLDIST <- log(Birds$LDIST)

# Check histograms after transformation
par(mfrow = c(1, 3))
hist(Birds$LOGAREA,  main = "log10(Area)")
hist(Birds$LOGDIST,  main = "log10(Distance)")
hist(Birds$LOGLDIST, main = "log(LDistance)")
```

### Step 3 — Fit the Full ANCOVA Model

The model to fit is:

$$y = \beta_0 + \beta_1\log(\text{AREA}) + \beta_2\log(\text{DIST}) + \beta_3\log(\text{LDIST}) + \beta_4\text{ALT} + \beta_5\text{YR.ISOL} + \beta_6 G_2 + \beta_7 G_3 + \beta_8 G_4 + \beta_9 G_5 + e$$

where $G_2, G_3, G_4, G_5$ are dummy variables for GRAZE levels 2–5 (GRAZE = 1 is the reference group).

```{r}
#| eval: false

model1 <- lm(ABUND ~ LOGAREA + LOGDIST + LOGLDIST + YR.ISOL + ALT + GRAZE,
             data = Birds)
summary(model1)
```

### Step 4 — Worksheet Questions to Answer

::: {.callout-important}
## ⭐ Practice Questions (Worksheet 8)
Work through these questions — they are the same style as exam questions.

**Q1.** Write down the equation of the fitted regression model using the estimated coefficients from R.

**Q2.** Interpret the coefficient for `log(AREA)`. Is it positive or negative? What does that mean for bird abundance?

**Q3.** Test the overall model:
$$H_0: \beta_1 = \beta_2 = \cdots = \beta_9 = 0$$
$$H_A: \text{at least one } \beta_i \neq 0$$
What do the results imply?

**Q4.** Test the hypothesis for `log(LDIST)`:
$$H_0: \beta_3 = 0 \qquad H_A: \beta_3 \neq 0$$
What do the results imply for the model?

**Q5.** Calculate **Variance Inflation Factors (VIFs)** for the model. What do they tell you about **collinearity**?

```r
library(car)
vif(model1)
```

**Q6.** Run diagnostic plots (`plot(model1)`). Check all three assumptions:
- Residuals vs Fitted — any pattern?
- QQ plot — do residuals follow a straight line?
- Scale-Location — is variance constant?

**Q7.** Fit the simpler model with only `LOGAREA` and `GRAZE`. Plot the five regression lines (one per grazing level). Are the lines parallel? What does that tell you?

```r
model2 <- lm(ABUND ~ LOGAREA + GRAZE, data = Birds)
summary(model2)
```
:::

---

# Summary Points

- ✅ **ANCOVA = ANOVA + a continuous covariate.** Use it when you have both categorical and continuous predictors.
- ✅ The **categorical variable is the main interest**; the covariate is included only to **control for its influence** on the response.
- ✅ The full ANCOVA model is: $Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ + e$ — where $X$ is continuous, $Z$ is the dummy variable, and $XZ$ is their interaction.
- ✅ **Always check the interaction term first.** If not significant (p > 0.05), the lines are parallel — consider removing it.
- ✅ The **Parsimony Principle:** prefer the simpler model if it fits equally well. Drop terms starting with the highest p-value.
- ✅ **R² always decreases** when you drop a term, even an unhelpful one. Use **Adjusted R²** to compare models fairly.
- ✅ An **increase in Adjusted R²** after dropping a term confirms the term was not contributing.
- ✅ The three assumptions are: **independence, normality, and equal variance** — checked using residual plots and statistical tests.
- ✅ In the exam: be ready to **write the model, state hypotheses, interpret coefficients, and compare models** from R output.

---

# Pre-Exam Checklist

Before moving on, make sure you can do all of the following:

- [ ] Explain what ANCOVA is and when to use it (vs ANOVA, vs regression)
- [ ] Explain what a covariate is and why we include it
- [ ] Identify all components of an experimental design (units, factors, levels, treatments, covariate, response)
- [ ] Classify a design as single-factor, CRD, factorial, or full factorial
- [ ] Write the ANCOVA model equation from a scenario description
- [ ] Interpret each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) in plain language
- [ ] Derive separate group equations by substituting Z = 0 and Z = 1
- [ ] State $H_0$ and $H_A$ for the overall model, the interaction, and each main effect
- [ ] Read R output and decide: significant or not, positive or negative effect?
- [ ] Explain why you always check the interaction term first
- [ ] Apply the Parsimony Principle to drop non-significant terms
- [ ] Explain the difference between R² and Adjusted R², and which to use when comparing models
- [ ] Interpret QQ plots, residuals vs fitted plots, and histogram of residuals
- [ ] Run the full ANCOVA workflow in R (load data → plot → fit model → diagnose)

---

# Connections to Previous Weeks

| Concept Needed | Covered In |
|----------------|------------|
| ANOVA framework, F-test, treatment effects | Week 5 — One-Way ANOVA |
| Interaction terms, two-factor models | Weeks 6/7 — Two-Way ANOVA |
| Regression coefficients, R², diagnostic plots | Week 8 — Simple Linear Regression |
| Multiple predictors, dummy variables, VIF | Week 9 — Multiple Regression |
| Log transformations for skewed data | Week 9 / Transformations notes |

---

# Additional Resources

- 📘 **Past Exam Paper:** `STAT9004_Exam_19.pdf` — on Canvas
- 🌐 **QQ Plot Interpretation Guide:** [http://seankross.com/2016/02/29/A-Q-Q-Plot-Dissection-Kit.html](http://seankross.com/2016/02/29/A-Q-Q-Plot-Dissection-Kit.html)
- 📂 **Data files:** `fruit.txt` and `Birds.txt` — available on Canvas
- 📄 **R Script:** `WS_8_ANCOVA.R` — full working code for both exercises

---

# Annotation Space

*Use this section to add your own notes as you review.*

---

**My questions / things I need to ask the Professor:**

&nbsp;

&nbsp;

&nbsp;

---

**Things I found confusing and how I resolved them:**

&nbsp;

&nbsp;

&nbsp;

---

**My own worked examples or practice answers:**

&nbsp;

&nbsp;

&nbsp;
