---
title: "Week 2: Statistical Inference"
subtitle: "STAT9004 — Regression Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show R code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

::: {.callout-important}
## ⭐ Assessment Reminders

- This week is a **review week** — the professor confirmed that all topics covered here form the **foundation for everything that comes next** in STAT9004, including all linear model assessments.
- The professor stated directly: *"For your assessment as well, [you need to] know the importance of statistical significance and all that."*
- **Key exam-relevant areas identified from the exercise sheet:** p-values, Type II errors, sample size calculations, power analysis, confidence intervals, and effect size interpretation.
- When writing conclusions in any report or assignment, you **must** quote the p-value and state conclusions in terms of the **population parameter**.
:::

---

# Key Topics

This week covers the following main ideas:

- Descriptive vs. Inferential Statistics
- Population vs. Sample (Parameters vs. Statistics)
- Probability Sampling
- Point Estimates and Sampling Error
- The Standard Error of the Mean (SEM)
- Sampling Distributions and the Central Limit Theorem (CLT)
- Confidence Intervals
- Hypothesis Testing: Null and Alternative Hypotheses
- The p-value and Level of Significance
- Type I and Type II Errors
- Statistical Power
- Effect Size (Cohen's d and R²)
- Statistical vs. Clinical (Practical) Significance

---

# Core Content

## Descriptive vs. Inferential Statistics

**Descriptive statistics** simply describe the data you already have. For example, you might calculate the mean exam score or draw a chart. You are not trying to say anything beyond what is in front of you.

**Inferential statistics** go further. They use methods such as **hypothesis testing**, **confidence intervals**, and **regression analysis** to make conclusions about a large group (a **population**) based on a smaller group (a **sample**). These methods are grounded in **probability theory** — we use probability to say how confident we are in our conclusions.

::: {.callout-note}
## 💡 Analogy

Think of a soup chef who wants to know if the soup is salty enough. They do not drink the entire pot — they take one spoonful (the **sample**) and taste it (the **statistic**). Their judgment about the whole pot is an **inference** about the **population**.
:::

> *Professor's note:* *"Statistical inferences are based on a process of reasoning based on probability theory."*

---

## Population vs. Sample

| Term | Definition | Symbol |
|------|------------|--------|
| **Population** | All individuals under study — the full group you want to learn about | $N$ |
| **Sample** | A smaller subset taken from the population | $n$ |
| **Parameter** | A number that describes the population (usually **unknown**) | $\mu$, $\sigma$ |
| **Statistic** | A number calculated from the sample, used to **estimate** a parameter | $\bar{x}$, $s$ |

In practice, the population is almost always too large to measure fully. We use sample statistics to **estimate** unknown population parameters.

For the estimate to be reliable, the sample must be collected in an **unbiased** way.

> *Professor's note:* *"In general, the parameters of a population are unknown. We use statistics to estimate parameters."*

---

## Probability Sampling

For a sample to be a fair representation of the population, we use **probability sampling**. This means:

- Every member of the population has a **known, non-zero chance** of being selected.
- Units are chosen **at random**.
- All statistical inference methods assume that the sample was collected this way.

If the sample is not random, the estimates may be **biased** — they may systematically over- or under-estimate the true value.

::: {.callout-warning}
## ⚠️ Common Mistake

Students sometimes confuse "random" with "haphazard". In statistics, **random** means that every unit has an equal and known probability of selection — it is a mathematical concept, not just picking things carelessly.
:::

---

## Point Estimates and Sampling Error

From a random sample, we calculate a **point estimate** — a single number that is our best guess for an unknown population parameter.

For example:

- The **sample mean** $\bar{x}$ is a point estimate of the **population mean** $\mu$.

However, because the sample is not the whole population, there will always be a gap between the estimate and the true value. This gap is called **sampling error**.

$$
\text{Sampling Error} = \hat{\theta} - \theta
$$

where $\hat{\theta}$ is the estimate and $\theta$ is the true parameter.

### Ways to Reduce Sampling Error

| Method | How It Helps |
|--------|-------------|
| Increase sample size | Larger samples are more representative |
| Use random/stratified sampling | Reduces systematic bias |
| Calculate the Standard Error of the Mean | Quantifies how much error to expect |
| Use Confidence Intervals | Gives a range that accounts for uncertainty |

---

## Standard Error of the Mean (SEM)

The **Standard Error of the Mean** tells us how much the sample mean $\bar{x}$ is expected to vary from sample to sample. It is calculated as:

$$
\text{SEM} = \frac{\sigma}{\sqrt{n}}
$$

Where:

- $\sigma$ = population standard deviation
- $n$ = sample size

::: {.callout-important}
## ⭐ Key Formula

$$\text{SEM} = \frac{\sigma}{\sqrt{n}}$$

As $n$ increases, SEM gets **smaller** — the estimate becomes more **precise**. This formula appears in confidence intervals, sampling distributions, and power analysis.
:::

::: {.callout-note}
## 💡 Analogy

Imagine measuring the height of people in a country. If you measure 5 people, your average could easily be wrong. If you measure 500 people, you are much more likely to be close to the true average. The SEM captures this idea mathematically.
:::

---

## Sampling Distributions

If you took many different random samples from the same population and calculated the mean for each, those means would form their own distribution — the **sampling distribution of the mean**.

- The sample mean is a **random variable** — it changes each time you take a new sample.
- If the underlying population follows a Normal distribution with mean $\mu$ and standard deviation $\sigma$:

$$
\bar{X} \sim N\left(\mu,\ \frac{\sigma}{\sqrt{n}}\right)
$$

This means the sampling distribution has:

- The **same mean** as the population ($\mu$)
- A **smaller standard deviation** equal to the SEM ($\sigma / \sqrt{n}$)

---

## The Central Limit Theorem (CLT)

::: {.callout-important}
## ⭐ One of the Most Important Results in Statistics

**Central Limit Theorem (CLT):** For large sample sizes ($n \geq 30$), the distribution of the sample mean is **approximately Normal**, *regardless* of the shape of the underlying population.

$$
\bar{X} \approx N\left(\mu,\ \frac{\sigma}{\sqrt{n}}\right) \quad \text{when } n \geq 30
$$

This is what allows us to use Normal-distribution methods (confidence intervals, z-tests) even when the original data is not normally distributed.
:::

### R Code: Demonstrating the CLT

The exercise sheet asks you to simulate rolling a die to see the CLT in action. As the sample size grows, the distribution of the mean becomes more and more bell-shaped.

```{r clt-demo, eval=FALSE}
n_samples <- 10000   # Number of simulated samples
n <- c(1, 2, 5, 10, 100, 1000)  # Sample sizes to try

X <- matrix(0, n_samples, length(n))

par(mfrow = c(3, 2))  # Arrange plots in 3 rows, 2 columns

for (i in n) {
  j <- which(n == i)
  for (k in 1:n_samples) {
    X[k, j] <- mean(as.integer(runif(i, 1, 7)))
  }
  hist(X[, j],
       main = paste("Distribution of the mean of", i, "rolls"),
       freq = FALSE,
       xlim = c(0.5, 6.5))
}
```

> *What to observe:* With n = 1, the histogram looks flat (uniform). As n increases toward 1000, the histogram becomes a narrow, bell-shaped curve centred at 3.5 (the true mean of a fair die). This is the CLT at work.

---

## Confidence Intervals

A **confidence interval (CI)** is a range of values that we are highly confident contains the true population parameter.

### The Logic Behind a 95% CI

If we took many samples and built a 95% CI for each one, approximately 95% of those intervals would contain the true population mean $\mu$.

For a single sample, we say: we are **95% confident** that $\mu$ lies within two standard errors of $\bar{x}$:

$$
\bar{x} \pm 2 \cdot \frac{\sigma}{\sqrt{n}}
$$

::: {.callout-important}
## ⭐ Why Confidence Intervals Matter

- A **larger sample** produces a **narrower** CI (greater certainty).
- CIs show the **range of plausible effect sizes**, not just whether a result is "significant".
- They help you distinguish between **statistical significance** and **practical significance**.

> *Professor's note:* *"If you provide a confidence interval for the parameter you are estimating, this will estimate the size of an effect. The use of confidence intervals facilitates the distinction between statistical significance and clinical significance."*
:::

---

## Hypothesis Testing

**Hypothesis testing** is a formal way of using data (evidence) to answer research questions. It ensures that — in the long run — we reach the correct conclusion a certain percentage of the time.

### Steps in Hypothesis Testing

1. **State the hypotheses** — null ($H_0$) and alternative ($H_1$).
2. **Set the significance level** ($\alpha$, usually 0.05).
3. **Collect data and compute a test statistic**.
4. **Calculate the p-value**.
5. **Make a decision** — reject or fail to reject $H_0$.
6. **State your conclusion** in terms of the population parameter, and always quote the p-value.

> *Professor's note:* *"When you are doing any reports or anything, you should always quote your p-value in your conclusions. And remember, when you're drawing your conclusions, if you're doing hypothesis testing, your conclusions are about the population parameter that you're studying."*

---

### The Null Hypothesis ($H_0$)

The **null hypothesis** is the starting point — a statement of "no effect" or "no difference". We assume it is true until the data gives us enough evidence to doubt it.

- Always stated with an **equality sign**: $H_0: \mu = 0$
- The statement is made about the **population parameter**, not the sample statistic.

> *Professor's note:* *"We assume this statement is true until the data indicates otherwise."*

---

### The Alternative Hypothesis ($H_1$)

The **alternative hypothesis** is your **research theory** — what you believe is actually true.

- It contains all values **not** covered by the null hypothesis.
- Together, $H_0$ and $H_1$ must cover all possible situations.

| Type | Hypotheses | Called |
|------|-----------|--------|
| Two-sided | $H_0: \mu = 0$, $H_1: \mu \neq 0$ | Two-tailed test |
| One-sided (greater) | $H_0: \mu = 0$, $H_1: \mu > 0$ | One-tailed test |
| One-sided (less) | $H_0: \mu = 0$, $H_1: \mu < 0$ | One-tailed test |

> *Professor's note:* *"The alternative hypothesis is your research theory. The researcher actually believes this statement to be true."*

---

## The p-value

The **p-value** is the result of the significance test. It tells you how likely the null hypothesis is to be true, given the data you observed.

More precisely: it is the probability of observing data at least as extreme as yours, *if the null hypothesis were true*.

::: {.callout-important}
## ⭐ The p-value — Most Important Concept in Hypothesis Testing

- A **small p-value** → the null hypothesis is unlikely to be true → we have **strong evidence against** $H_0$.
- Always report the exact p-value in your conclusions.

> 🎵 *Professor's classroom rhyme (do not write in exams!):*
> **"If p is low, null must go."**
> That is: if $p \leq \alpha$ (e.g., $p \leq 0.05$), **reject** $H_0$.

The **only two valid conclusions** you can write are:

1. *"We reject the null hypothesis."* (when $p \leq \alpha$)
2. *"We fail to reject the null hypothesis."* (when $p > \alpha$)

**Anything else is incorrect.**

> *Professor's note:* *"When I teach hypothesis testing, there's only two statements that you're allowed to use. Everything else is incorrect."*
:::

---

## Level of Significance ($\alpha$)

The **significance level** $\alpha$ is set **before** the test is run. It is the amount of error you are willing to accept in your conclusion.

| $\alpha$ | Meaning |
|----------|---------|
| 0.05 (5%) | Most common — you accept a 5% chance of a Type I error |
| 0.01 (1%) | More stringent — used when false positives are very costly |
| 0.001 (0.1%) | Very stringent — used in fields like genetics |

**Decision rule:**

- If $p \leq \alpha$: **Reject $H_0$** → result is **statistically significant**
- If $p > \alpha$: **Fail to reject $H_0$** → not sufficient evidence

> *Professor's note:* *"This is the degree of uncertainty we are prepared to accept in a conclusion. It depends on your study."*

::: {.callout-warning}
## ⚠️ Common Mistakes

- A **significant** result does **not** mean the result is certainly true — only that we have evidence.
- A **non-significant** result does **not** mean the null hypothesis is true — only that we do not have enough evidence to reject it.
- Never say: *"We accept $H_0$"* or *"We accept $H_1$"*. These are **wrong**.
:::

---

## Type I and Type II Errors

Because we base conclusions on sample data, we can sometimes make mistakes.

| | $H_0$ is actually **TRUE** | $H_0$ is actually **FALSE** |
|---|---|---|
| **We Reject $H_0$** | ❌ **Type I Error** (probability = $\alpha$) | ✅ Correct — this is **Power** |
| **We Fail to Reject $H_0$** | ✅ Correct | ❌ **Type II Error** (probability = $\beta$) |

- **Type I Error** = "False positive" — rejecting a true null hypothesis. You can control this by setting $\alpha$.
- **Type II Error** = "False negative" — failing to detect a real effect. Its probability is $\beta$. You cannot directly control it without knowing the true population value.

::: {.callout-important}
## ⭐ Key Point on Errors

We can **control** Type I errors by setting $\alpha$. We **cannot directly control** Type II errors — because to calculate the exact probability of $\beta$, we would need to know the true population parameter, which is unknown.

> *Professor's note:* *"We controlled for Type 1 errors by setting the level of significance, your alpha, or your 5%, or your 1%."*
:::

::: {.callout-note}
## 💡 Real-World Analogy

A Type I Error is like a **false fire alarm** — the alarm goes off (you reject $H_0$) but there is no fire (no real effect). A Type II Error is like a **missed fire** — there is a real fire (a real effect), but the alarm does not go off (you fail to detect it).
:::

---

## Statistical Power

**Power** is the probability of correctly detecting a real effect — of rejecting $H_0$ when it truly is false.

$$
\text{Power} = 1 - \beta
$$

- **High power** is desirable — you want your study to have a good chance of finding an effect if one really exists.
- **The trade-off:** Decreasing $\alpha$ (being more careful about Type I errors) automatically *increases* $\beta$ (makes Type II errors more likely), which *reduces* power.

### Increasing Power

The best solution to this trade-off is to **increase the sample size** before the study begins. With enough data, you can keep both $\alpha$ and $\beta$ at acceptable levels simultaneously.

```{r power-example, eval=FALSE}
# Install packages if needed
# install.packages("pwr")
# install.packages("asbio")

library(pwr)
library(asbio)

# Q6: One-sample test — power of detecting a 0.5 hr difference
# sigma = 1.35, n = 50, alpha = 0.05, effect = 0.5
power.z.test(sigma = 1.35, n = 50, alpha = 0.05, effect = 0.5,
             test = "two.tail", strict = FALSE)

# Q6d: Minimum sample size for 90% power, detecting 15-minute (0.25 hr) difference
power.z.test(sigma = 1.35, power = 0.9, alpha = 0.05, effect = 0.25,
             test = "two.tail", strict = FALSE)

# Q7a: Two-sample t-test — power, n=20 per group, pooled SD=6.2, effect=5 complaints
pwr.t.test(n = 20, d = (5/6.2), sig.level = 0.05,
           type = "two.sample", alternative = "two.sided")

# Q7b: Recalculate with alpha = 0.01
pwr.t.test(n = 20, d = (5/6.2), sig.level = 0.01,
           type = "two.sample", alternative = "two.sided")

# Q7c: Minimum sample size for 90% power
pwr.t.test(power = 0.9, d = (5/6.2), sig.level = 0.05,
           type = "two.sample", alternative = "two.sided")

# Q8c: HbA1c intervention — one-sample, one-sided, 80% power
pwr.t.test(power = 0.8, d = (1/2.2), sig.level = 0.05,
           type = "one.sample", alternative = "less")
```

> *Professor's note:* *"We can select a sample size before the study commences so that alpha and beta are both at satisfactory levels."*

---

## Effect Size

The **effect size** tells you how **big or strong** an effect is — separate from whether it is statistically significant.

A result can be statistically significant but practically tiny. The effect size tells you if the result actually matters in the real world.

- **Larger effect size** = stronger relationship = more likely to be practically important
- Associated with every effect size is a **standard error** (a measure of its uncertainty)

::: {.callout-important}
## ⭐ Why Effect Size is Important

> *Professor's note:* *"It's good in practice to report your effect size. You're doing it anyway — you don't know you're doing it when you're reporting a confidence interval."*

Reporting a confidence interval is a way of reporting effect size, even if you do not realise it.
:::

### Cohen's d (for t-tests)

Used when comparing two means:

$$
d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}
$$

The difference between two means, divided by the pooled standard deviation. This standardises the difference so it can be compared across studies.

### $R^2$ (for Linear Regression)

In linear regression, effect size is measured by $R^2$ — the square of Pearson's correlation coefficient $r$:

$$
R^2 = r^2
$$

$R^2$ is the **coefficient of determination**. It tells you what proportion of the variation in $Y$ is explained by $X$.

::: {.callout-important}
## ⭐ How to Interpret $R^2$

> *Professor's note:* *"If you had a coefficient of determination that was 40%, the interpretation would be: 40% of the variation in Y is explained by the variation in X. Which leaves 60% — that's unexplained variation, maybe due to something else you haven't considered."*

**Always express $R^2$ as a percentage** when interpreting it, and identify what the unexplained variation might be due to.
:::

---

## Statistical vs. Clinical (Practical) Significance

This is one of the most important distinctions in applied statistics.

| | Statistical Significance | Clinical / Practical Significance |
|---|---|---|
| **What it means** | Evidence that an effect exists in the data | The effect is large enough to matter in practice |
| **How it is judged** | p-value vs. $\alpha$ | Effect size |
| **Risk when sample is large** | Small, trivial effects may be "significant" | May still have very small effect size |
| **Risk when sample is small** | A real effect may not be detected | Under-powered study may miss important differences |

**For a result to be truly meaningful, it must be:**

1. First — **statistically significant** ($p \leq \alpha$)
2. Then — have a **meaningful effect size** (large enough to matter practically)

::: {.callout-important}
## ⭐ The Blood Pressure Example

> *Professor's note:* *"If you're studying blood pressure — small changes in blood pressure might be practically significant. It depends on the subject that you're studying."*

In medicine, a statistically significant reduction of even 2 mmHg in blood pressure across a large population could prevent thousands of strokes. So "small" effect sizes can still be practically important — it always depends on **context**.
:::

::: {.callout-warning}
## ⚠️ Common Mistake: Over-relying on the p-value

> *Professor's note:* *"Instead of attaching too much importance on statistical significance, we should pay attention to the actual data as well as the p-value."*

- A **non-significant result** does NOT mean there is no effect. It means you did not have enough evidence to detect one (perhaps your sample size was too small).
- A **significant result** with a tiny effect may have little practical value.
- Always interpret **both** the p-value and the confidence interval together.
:::

---

# Exercise Sheet Summary

This table maps the main exercises to the concepts they test.

| Question | Topic | Key Concept |
|----------|-------|------------|
| Q1 | p-value | What exactly does a p-value tell you? |
| Q2 | Type II error | Definition of Type II error |
| Q3 | Power & sample size | Effect of increasing n on Type II error probability |
| Q4 | Bias | Definition of an unbiased estimator |
| Q5 | CLT simulation | Visualising how sample mean distribution becomes Normal |
| Q6 | Power analysis | One-sample z-test; power calculation by hand and in R |
| Q7 | Power analysis | Two-sample t-test; power and minimum sample size |
| Q8 | Power analysis | Clinical trial; HbA1c; one-sided test; Type I/II interpretation |
| Q9 | Sample size for CI | Margin of error: battery life, $n$ for 95% CI |
| Q10 | Sample size for CI | Margin of error: sugar content, $n$ for 99% CI |
| Q11 | Sample size for CI | Margin of error: flight delays, $n$ for 99% CI |

### Sample Size for a Confidence Interval (Questions 9–11)

When you want to estimate a population mean with a specific **margin of error** $E$ and confidence level, the required sample size formula is:

$$
n = \left( \frac{z_{\alpha/2} \cdot \sigma}{E} \right)^2
$$

Where:

- $z_{\alpha/2}$ = critical value (1.96 for 95%, 2.576 for 99%)
- $\sigma$ = population standard deviation
- $E$ = desired margin of error

Always **round up** to the nearest whole number.

---

### Practice Question with Answer

::: {.callout-important}
## ⭐ Practice: State H₀ and H₁

A company targets a mean order processing time of 6 hours. A sample of 50 orders had a mean of 5.2 hours and standard deviation of 1.35 hours. State the hypotheses.

**Answer:**

$$H_0: \mu = 6 \text{ hours}$$

$$H_1: \mu \neq 6 \text{ hours}$$

Note: $H_0$ uses an **equality sign** and refers to the **population mean** $\mu$, not the sample mean $\bar{x}$.
:::

---

# Key Formulas Reference

| Formula | What It Is |
|---------|-----------|
| $\text{SEM} = \sigma / \sqrt{n}$ | Standard Error of the Mean |
| $\bar{X} \sim N(\mu,\ \sigma/\sqrt{n})$ | Sampling distribution (Normal population) |
| $\bar{x} \pm 2 \cdot (\sigma/\sqrt{n})$ | Approximate 95% Confidence Interval |
| $\text{Power} = 1 - \beta$ | Power of a test |
| $d = (\bar{x}_1 - \bar{x}_2) / s_{\text{pooled}}$ | Cohen's d (effect size for t-tests) |
| $R^2 = r^2$ | Coefficient of determination (effect size for regression) |
| $n = (z_{\alpha/2} \cdot \sigma / E)^2$ | Sample size for estimating a mean |

---

# Summary Checklist

- ✅ I can explain the difference between descriptive and inferential statistics.
- ✅ I can distinguish between a population parameter and a sample statistic.
- ✅ I know why random (probability) sampling is essential.
- ✅ I can define and calculate the Standard Error of the Mean (SEM).
- ✅ I can state the Central Limit Theorem and explain what it means in plain language.
- ✅ I can explain the logic of a confidence interval.
- ✅ I can state a null and alternative hypothesis correctly.
- ✅ I know what a p-value tells me and how to use it to make a decision.
- ✅ I know the two and only two valid conclusions in hypothesis testing.
- ✅ I can define Type I and Type II errors and explain which one we can control.
- ✅ I can define statistical power and explain the trade-off between $\alpha$ and $\beta$.
- ✅ I can explain what effect size is and name two measures: Cohen's d and $R^2$.
- ✅ I can interpret $R^2$ as a percentage (e.g., "40% of the variation in Y is explained by X").
- ✅ I can distinguish between statistical significance and clinical/practical significance.
- ✅ I always report the p-value and a confidence interval together in my conclusions.

---

# Pre-Topic Checklist: Before Moving to Week 3

- [ ] I can calculate the SEM by hand for given values of $\sigma$ and $n$.
- [ ] I can explain why the CLT is important and when it applies ($n \geq 30$).
- [ ] I can write a pair of hypotheses ($H_0$ and $H_1$) for a given research question.
- [ ] Given a p-value and $\alpha$, I can make the correct decision and write the correct conclusion statement.
- [ ] I understand the difference between Type I and Type II errors and can give an example of each.
- [ ] I can explain power ($1 - \beta$) and describe how to increase it.
- [ ] I can use the `pwr` package in R to calculate power and minimum sample sizes.
- [ ] I can calculate the required sample size for a confidence interval using the margin of error formula.
- [ ] I can interpret $R^2$ and explain what the "unexplained" variation represents.
- [ ] I can correctly interpret a confidence interval — I do NOT say "there is a 95% probability that $\mu$ is in this interval" (incorrect), but rather that "95% of such intervals would contain the true mean".

---

# Additional Resources

- R packages used: `pwr`, `asbio` — install with `install.packages("pwr")` and `install.packages("asbio")`
- R script for Exercise 2 is available on Canvas (under Units → Regression Analysis)
- *Fethney (2010)* paper on statistical and clinical significance — available in the course materials

---

# Annotation Space

Use the space below for your own handwritten or digital notes during revision.

> ✏️ *Lecture notes / clarifications:*

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

---

> ✏️ *Things I found confusing and need to revisit:*

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

---

> ✏️ *My own examples:*

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

---
